---
title: "Does multilingualism really protect against accelerated ageing? Some critical comments"
date: "2025-12-15"
categories: [correlational studies, R]
format:
  html:
    code-fold: true
    message: false
    df-print: paged
---

Amoruso et al.'s 
[_Multilingualism protects against accelerated aging in cross-sectional and longitudinal analyses of 27 European countries_](https://10.1038/s43587-025-01000-2)
struck a chord with multilingualism enthusiasts.
But whenever a study claims to have found something that suits us ideologically
or economically, it behooves us, as academics, to take a closer look at it.

So I did.

You can find my critical comments below.
Most are not very technical in nature, and many can be identified
just by reading the paper and looking at the data and code the authors shared---without actually needing to run it.
I believe that these comments should lead to a substantial revision of the published article.

Everything below is based on Amoruso et al.'s article, its supplementary materials,
and the data and code provided on the project's [GitHub](https://github.com/euroladbrainlat/Biobehavioral-age-gaps-Multilingualism) page,
which I've [forked](https://github.com/janhove/Biobehavioral-age-gaps-Multilingualism).
I didn't consult the sources that the authors obtained their data from.

## tl;dr

1. **Missing predictors:** The article suggests that 14 predictors were used,
but the data and code suggest that only 12 predictors were used.
1. **Missing respondents and countries:** The article claims that the evidence is based on data from 86,149 respondents in 27 European countries.
However, the cross-sectional analyses are based on data from only 84,127 respondents in _26_ countries;
the longitudinal analyses are based on data from only 81,329 respondents.
1. **Unjustified causal claims:** 
By their own admission in the discussion section, 
the authors can't make any causal claims, 
but already the title and the abstract make a bold causal claim.
1. **Questionable construction of the outcome variable:** 
The construction of the outcome variable ("BAGs") requires some more scrutiny. 
For the time being, I'll content myself by pointing out 
that it is a dichotomised version of a continuous metric.
1. **Conflation of country-level and individual multilingualism:** 
The multilingual data that the authors used is situated at the country level, 
but the authors regularly frame their findings in terms of individual multilingualism.
1. **Meaningless odds ratios:** 
The odds ratios reported in Table 1 are presented
without mentioning how the data were coded.
A closer look at the Jupyter notebooks reveals that they are essentially meaningless.
1. **Insufficient confounder control:** 
Several confounders are controlled for in the analyses---but only one at a time.
1. **GDP incorrectly controlled for:**
The authors claim to have used per capita GDP as a control variable.
Their data and code show that they used GDP---_not_ per capita GDP---instead.
1. **Insufficient accounting for nesting:**
The nested nature of the data (respondents in cohorts in countries)
isn't accounted for sufficiently.

## A first look
The first few critical comments can be raised by just reading
the article and comparing the claims made in it 
to the data and code made available by the authors.

### Data and code availability
Let's take a look at how Amoruso et al.'s analytical pipeline is supposed to work.

The dataset 
[`scripts/main/data/data.csv`](https://github.com/janhove/Biobehavioral-age-gaps-Multilingualism/blob/main/scripts/main/data/data.csv) contains 86,149 observations of 14 variables:
the respondents' country, age, sex, and scores and indicators of 11
biobehavioural factors:

  - `Education`,
  - [`Barthel`](https://en.wikipedia.org/wiki/Barthel_scale),
  - `Diabetes`,
  - `Hypertension`,
  - `Heart_Disease`,
  -  `Physical_activity`,
  - `Cognition`,
  - `Well_being_domain`,
  - `Sleep_problems`,
  - `Audition_problems`,
  - `Vision_problems`.

The authors derived what they dubbed _biobehavioural age gaps_ ('BAGs')
from this dataset in the Jupyter notebook `scripts/main_BAGs_computation.ipynb`.
Lots of the code in this notebook isn't really important, 
but if you look for the string `vars_list`,
you'll see that the eleven biobehavioural factors as well as the participants'
sex were used as predictors, for a total of twelve predictors.

But their article (particularly Figure 1b) suggests that fourteen predictors
were used to derive the BAGs. 
The factors _unhealthy weight_ and _alcohol consumption_ are mentioned
both in this Figure and in the article's Methods section,
but they simply aren't in the dataset and apparently weren't used
when deriving the BAGs.

Further, the Jupyter notebook doesn't actually output the BAGs.
Instead, two new datasets are fed to the second round of analyses.
The first of these new datasets is 
[`scripts/main/data/BAG_OR_cross.csv`](https://github.com/janhove/Biobehavioral-age-gaps-Multilingualism/blob/main/scripts/main/data/BAG_OR_cross.csv).
It contains 84,127 observations of 21 variables.
I'll discuss these variables later.
This dataset was used for the cross-sectional analyses carried out
in the Jupyter notebook 
[`scripts/main/ORs_cross.ipynb`](https://github.com/janhove/Biobehavioral-age-gaps-Multilingualism/blob/main/scripts/main/ORs_cross.ipynb).
The second is 
[`scripts/main/data/BAG_or_long.csv`](https://github.com/janhove/Biobehavioral-age-gaps-Multilingualism/blob/main/scripts/main/data/BAG_OR_long.csv),
which contains 81,329 observations of 22 variables
and was used for the longitudinal analyses carried out 
in the Jupyter notebook 
[`scripts/main/ORs_long.ipynb`](https://github.com/janhove/Biobehavioral-age-gaps-Multilingualism/blob/main/scripts/main/ORs_long.ipynb).

The difference in sample sizes is not commented on anywhere.
The difference between 86,149 (`data.csv`) and 84,127 (`BAG_OR_cross.csv`) is 2,022,
which happens to be exactly the size of the Slovak sample.
The supplementary Table 6 does not contain any information
about Slovakia.
But it does contain information about Serbia, 
which isn't represented in `data.csv`.
I suspect that the authors blundered 
when merging the BAGs with the country-level information.
Neither the actual BAG computation nor the merging is documented in
any of the notebooks, though.

Another 2,798 observations are missing from `BAG_OR_long.csv`.
I didn't investigate these further;
for the time being, I'll focus on the cross-sectional analyses.

### Unjustifiable causal claims
As the authors correctly point out in their discussion,
their design and analysis do not allow them to establish any causal
links between the degree of multilingualism in a country
and the extent to which its inhabitants show signs of accelerated ageing.
Unfortunately, they don't seem to have kept this insight in mind
when they came up with
the title ("Multilingualism protects...") and the abstract ("These results
underscore the protective role of multilingualism...").

Further, the authors write

> "[R]andomized controlled trials could explicitly
> manipulate multilingualism and directly assess its effects on aging clocks."

Could they really, though?

### Construction of the outcome variable {#sec-bag}
The study's outcome variable was the presence or absence of "accelerated ageing".
Let's see how this outcome variable was constructed.

First, the authors gathered data on individual "biobehavioural" factors (see `data.csv`).
These are based on self-reports and are fairly coarse.
For example, the variable `Hypertension` is simply the yes/no response
to whether a physician has ever told the respondent that their blood pressure was too high.
The list of specific questions can be found in the supplementary materials.

Then, they used the dataset to fit a model with the respondents'
age as the outcome variable and all of these biobehavioural factors as predictors.
The model itself is gradient boosting, but that is not especially important conceptually.
What is important is that the model is able to pick up on nonlinearities and interactions.
The model's hyperparameters were estimated from the data,
and a couple of different cross-validation techniques were used.
For our present purposes, I'm going to focus on the _leave one country out_ cross-validation,
because it makes the most sense to me.
What this means is that 27 models predicting age from biobehavioural factors were fitted, 
with each leaving out the data from another country.

Next, the models fitted were used to predict the age of the inhabitants of the country
that was left out when fitting them. 
That is, the Belgian respondents' ages were predicted on the basis of the model 
fitted to the data from the respondents in the other 26 countries.

Then, the authors computed, for each respondent,
$$\texttt{GAP} := \textrm{predicted age} - \textrm{actual age},$$
that is, the negative model residuals.
As the authors correctly point out, though, the predictive model will
tend to underestimate the age of the oldest respondents and
overestimate the age of the youngest respondents.
To account for this phenomenon, they fitted linear models
with $\texttt{GAP}$ as the outcome and the participants' actual age as the predictor;
let's call the predicted values of \emph{this} model the predicted $\texttt{GAP}$ values.
They then defined
$$\textrm{corrected } \texttt{GAP} := \texttt{GAP} - \textrm{predicted } \texttt{GAP}.$$

Finally, the authors defined the variable "biobehavioural age gap" ($\texttt{BAG}$)
as
$$\texttt{BAG} := \begin{cases}1, &\textrm{if corrected } \texttt{GAP} > 0, \\ 0, & \textrm{otherwise}.\end{cases}$$
I'm always somewhat skeptical when the input to one statistical model is the output of another one.
The $\texttt{BAG}$ values that serve as the outcome variable in the models
testing for a multilingualism effect are themselves the dichotomised output
of another statistical model whose input, in turn, is the
input \emph{and} the output of yet another statistical model.
At the very least, the literal meaning of these $\texttt{BAG}$ values is highly
convoluted.

Further, I do not see why the corrected $\texttt{GAP}$s had to be dichotomised in the first place,
thereby losing the distinction between a corrected $\texttt{GAP}$ of, say, half a year
and one of 12 years.
The authors write that these dichotomous variables were created
because otherwise no odds ratios (and relative risks)
could be computed. 
But this is hardly a satisfactory justification;
it's a bit like scattering dust throughout your living room
before vacuum cleaning it as otherwise there would be no dust to get rid of.

Lastly, note that any given respondent's $\texttt{BAG}$ value doesn't just
depend on this respondent's biobehavioural data: both the model to compute
their raw $\texttt{GAP}$ value as well as the one to compute their corrected $\texttt{GAP}$ value
are based on the data from the respondents in all the other countries.
In fact, you'd expect roughly half the respondents to have a positive $\texttt{GAP}$ value.

### Country-level vs individual multilingualism
The authors use country-level data on multilingualism,
namely the estimated percentage of monolinguals in the country,
the estimated percentage of people in the country who know (exactly) one additional language,
those who know (exactly) two additional languages,
and those who know at least three additional languages.
Throughout their article, though, they move back and forth between
interpreting country-level and individual-level interpretations.
For instance, they write

> "In the cross-sectional analysis, monolinguals
> (that is, those speaking only their mother tongue)
> were 2.11 times more likely to experience accelerated aging (...)."

and

> "Individuals speaking one additional language were 1.11 less likely
> to develop accelerated aging (...), those speaking two additional languages
> were 1.25 times less likely (...), and those speaking three
> or more additional languages were 1.41 less likely (...)."

Again, in the discussion they point out that their data are situated
at the country level, which limits their ability to draw conclusions
about individual multilingualism.
A cleaner and more consistent choice of wording in the title and throughout
the article would have been preferable.


### On the interpretation of the reported odds ratios
Now that we've established that the multilingualism data are situated
at the country level, what are we to make of the odds ratio of 2.11
in the claim that
"monolinguals ... were 2.11 times more likely to experience accelerated aging"?

Their Python code (`ORs_cross.ipynb`) shows that the authors scaled
the multilingualism data to a range from 0.05 to 0.95.
The highest degree of country-level monolingualism was 63.2,
the lowest was 2.86.
So, with $p$ the percentage of monolinguals in a given country,
the scaled monolingualism variable $s(p)$ becomes

$$s(p) = \left(\frac{p - 2.86}{63.2 - 2.86}\right)(0.95 - 0.05) +0.05.$$
The odds ratio output by the model is the estimated ratio
$$\textrm{OR(monolingualism)} = \frac{\textrm{odds of accelerated ageing if } s(p) = 1}{\textrm{odds of accelerated ageing if } s(p) = 0}.$$
As some basic algebra reveals, $s(p) = 0$ corresponds to a
percentage of monolinguals of
$$\frac{-0.05}{0.95 - 0.05}(63.2-2.86) + 2.86 = -0.49,$$
that is negative (!) 49 monolinguals per 10,000 inhabitants.
Similarly, $s(p) = 1$ corresponds to a percentage of monolinguals of
$$\frac{0.95}{0.95 - 0.05}(63.2-2.86) + 2.86 = 66.55,$$
that is 6,655 monolinguals for every 10,000 inhabitants.

What the odds ratio of 2.11 for monolingualism means, then,
is that the model estimates that the inhabitants of a country
with 6,655 monolinguals for every 10,000 inhabitants are 2.11 times more likely
to show signs of accelerated ageing compared to the inhabitants of a country
with negative 49 monolinguals per 10,000 inhabitants.

It would have been more sensible to set the baseline at 0% monolinguals
and have the OR refer to a difference in monolingualism of, say, 1 or 10 percentage points.

The interpretation of the other odds ratios suffers from the same problem, of course.
Further, these other odds ratios aren't too useful for an additional reason
Take, for example, the following statement:

> "speaking one foreign language were [sic.] 1.3 less likely (...)."

The highest percentage of speakers of exactly one additional language was
62.08; the lowest was 9.48.
The same computation as above shows that this odds ratio has the following interpretation:
According to the model, inhabitants of a country with 6,500 speakers of exactly
one additional language per 10,000 inhabitants
are 1.3 times less likely to show signs of accelerated ageing compared
to the inhabitants of a country with only 656 speakers of exactly one additional
language per 10,000 inhabitants.

Crucially, a country with 6,500 speakers of exactly one additional language
per 10,000 inhabitants isn't necessarily more multilingual
than a country with only 656 speakers of exactly one additional
language per 10,000 inhabitants:
In the 2016 data, Luxembourg only had 16.2% speakers of exactly one additional language
compared to, say, Austria's 49.6%.
But Luxembourg had 72% speakers of at least three additional languages
compared to Austria's 13.4%---more speakers of exactly one additional language doesn't mean that the country is more multilingual!

The same remark goes for the analyses concerning the percentage of
speakers of exactly two additional languages or of speakers of at least
three additional languages.

Further, the result that "individuals (sic.) speaking at least one additional foreign
language were 2.17 times less likely (...) to experience accelerated ageing" is superfluous:
if we already know the odds ratio for the degree of monolingualism,
we also know the odds ratio for the degree of multilingualism: it must be the inverse.
The fact that it isn't _exactly_ the inverse in this report is something I will
come back to when reproducing the analyses.

Finally, the operative phrase in the interpretations above is _according to the model_.
While the article contains quite a few graphs, none shows a data plot
for the association between degree of mono/multilingualism versus prevalence
of accelerated ageing.
It is therefore entirely possible that the actual relationship between
these two variables is nonlinear.

Incidentally, for some reason or other, the notebook `ORs_cross.ipynb`
doesn't actually fit the logistic regression models that the
authors claim to have fitted: They fit log-linear regression models.
Moreover, for reasons unspecified,
the data were split up into a training set and a test set,
but the test set was never used for anything.
This isn't mentioned anywhere in the article, either.

### Insufficient confounder control
The authors write in the abstract that the "[e]ffects persisted
after adjusting for linguistic, physical, social and sociopolitical exposomes."
This is true but also easily misunderstood:
They only controlled for one exposome at a time.
Hence, it is possible that the mono/multilingualism variables acted as a proxy
for some of the exposomes that weren't considered in the analysis.

Further, the authors write that one of the exposomes taken into consideration
was the GDP _per capita_.
However, the GDP variable in `BAG_OR_cross.csv` ranges from 
13.5 billion (in nominal USD?) to about 3.9 trillion (in nominal USD?).
I suspect the first figure is for Malta in 2017 and the second for Germany in 2019.
But nowhere in their analysis code do they compute the _per capita_ GDP.

### Nesting not taken into account
The logistic models fitted in `ORs_cross.ipynb` don't take the
nested nature of the data into account: the respondents' data 
were collected in different waves in different countries.
It is quite plausible that the reported confidence intervals
would be considerably wider once this nesting is properly accounted for.
Since the predictors in these models are situated at the country level
(e.g., country-level multilingualism and GDP), simple data aggregation
would be a viable approach.

## Getting stuck in
Now let's get our hands dirty and try 
to (a) reproduce the authors' findings using the code and data they provided
and (b) improve on these analyses.
In doing so, I will assume for the time being that their computation 
of the BAGs (the outcome variable) makes sense---even though I believe
it ought to be scrutinised further.
I will focus entirely on the cross-sectional analyses carried out in `ORs_cross.ipynb`.

I carried out part of the reproduction in Python, using a stripped-down
and refactored version of the notebook that the authors shared 
in their GitHub repository; I'll link to my Python script when we need it.
I additionally also used R.

```{r, warning = FALSE, message = FALSE}
# Load packages and set plotting theme
options(show.signif.stars = FALSE, tidyverse.quiet = TRUE)
library(DT)
library(tidyverse)

theme_set(theme_bw(12))
```

### Reproducing the odds ratios
I tried to reproduce the first couple of odds ratios reported in the
authors' Table 1, namely those reported for the models without additional covariates.
The table caption says that these were estimated using logistic regression models.
In their Python code (`ORs_cross.ipynb`), however, they fit log-link models,
not logistic models:

````{verbatim}
sm.GLM(y_train, X_train_scaled, family=sm.families.Binomial(link=sm.families.links.log())).fit()
````

The output of such models can still be converted to odds ratios, though.
When I run the `ORs_cross.ipynb` notebook and convert the parameters
of the log-linear models fitted in it to odds ratio, 
I obtain slightly different results from the ones reported in the article,
see @tbl-odds-ratios.

```{r}
#| label: tbl-odds-ratios
#| tbl-cap: "The intercept and slope parameter for the log-linear models without any co-variables fitted in `ORs_cross.ipynb`, the odds ratios reported in Amoruso et al.'s Table 1, and the ORs recomputed from the log-linear models."
log_to_OR <- function(intercept, slope) {
  odds0 <- exp(intercept) / (1 - exp(intercept))
  odds1 <- exp(intercept + slope) / (1 - exp(intercept + slope))
  odds1 / odds0
}

tibble(
  Predictor = c("Mono", "One", "Two", "Three", "Total"),
  Intercept = c(-0.916226, -0.699792, -0.628158, -0.704455, -0.527900),
  Slope = c(0.388409, -0.145554, -0.343348, -0.226496, -0.388527),
  `OR reported` = c(2.11, 0.75, 0.54, 0.67, 0.47)
) |>
  mutate(`OR computed` = log_to_OR(Intercept, Slope)) |> 
  mutate_if(is.numeric, round, 3)
```

There could be several reasons for the discrepancies between the reported odds ratios
and the ones that are computed using their code.
I'm not going to speculate on these reasons.

As already mentioned above, 
the authors rescaled the predictor variables (`Mono`, `One`, etc.)
to a range from 0.05 and 0.95
and split off an used test set from the main data.
I don't see any added value in these steps.
Further, I do not see much value
in the results other than the one for either the $\texttt{Mono}$
or the $\texttt{Total}$ variable;
the former is the percentage of monolinguals,
the latter the percentage of non-monolinguals.

Next, let us take a closer look at some of the other results in their Table 1.
These concern logistic regression models for which another predictor
was taken into account.
Because of the issue identified with the GDP variable above,
I've focused on reproducing the findings for this analysis.
As @tbl-OR-gdp shows, the results in the authors' Table 1
are indeed roughly consistent with the models fitted in 
`ORs_cross.pynb` in which GDP (_not_ per capita GDP!)
was used as a control variable.
Again, there could be several reasons for the slight discrepancies;
I'll refrain from speculating on these.

```{r}
#| label: tbl-OR-gdp
#| tbl-cap: "The intercept and slope parameter for the log-linear models with GDP as the co-variable fitted in `ORs_cross.ipynb`, the odds ratios reported in Amoruso et al.'s Table 1, and the ORs recomputed from the log-linear models."
tibble(
  Predictor = c("Mono", "One", "Two", "Three", "Total"),
  Intercept = c(-0.905862, -0.703005, -0.618220, -0.690530, -0.512324),
  Slope = c(0.393635, -0.155711, -0.348275, -0.240334, -0.393752),
  `OR reported` = c(2.16, 0.74, 0.53, 0.65, 0.46)
) |>
  mutate(`OR computed` = log_to_OR(Intercept, Slope)) |> 
  mutate_if(is.numeric, round, 3)
```



### Missing info
Important information is missing from the dataset `BAG_OR_cross.csv`,
on which the cross-sectional analyses are based:
it doesn't specify the country and the wave of data collection that the respondents belong to,
nor does it contain any information about the respondents' age
or the corrected $\texttt{GAP}$ value.
Only country/cohort level covariates and the binary $\texttt{BAG}$ variable are available; 
the latter is called `GAP_bin` in the dataset.
Note that the `ID` variable is just the row number (using zero-based indexing);
it doesn't actually allow you to link a particular respondent's
data in one dataset to their data in another dataset.

```{r}
d_cross <- read.csv("BAG_OR_cross.csv")
```

There are 97 unique combinations of country- and cohort-level predictors.
```{r}
country_cohort <- d_cross |>
  select(-ID, -GAP_bin) |>
  group_by_all() |>
  summarise(n = n(), .groups = "drop")
# write_csv(country_cohort, "country_cohort.csv")
```

Using the information provided in the supplementary materials,
I could piece together which of these combinations corresponded
to which country and wave of data collection.
I also had ChatGPT create a table with the approximate
population of each country in each year of data collection.
I've put this information in the dataset [`country_cohort.csv`](country_cohort.csv).
Let's add it to the dataset for the cross-sectional analyses.
While we're at it, let's compute per capita GDP, too,
expressed in 10,000 units of whatever currency the GDP variable is expressed in.

```{r}
country_cohort <- read.csv("country_cohort.csv")
datatable(country_cohort)
d_cross <- d_cross |>
  left_join(country_cohort) |> 
  mutate(GDP_percapita = (GDP / Pop) / 10000)
```

### Drawing some plots
Amoruso et al. drew lots of plots---just none that show how the predictors
and the BAGs are related.
A first (naive) plot is shown in @fig-naive.
The smooth trend line attempts to estimate the proportion of
respondents showing accelerated ageing (again, I'm taking the 
accelerated ageing variable as given for now; I'll scrutinise it later)
based on the extent of country-level monolingualism.

```{r}
#| label: fig-naive
#| fig-cap: A naive trendline showing the extent of accelerated ageing according to country-level monolingualism.
#| fig-height: 5
#| fig-width: 7
#| out-width: 80%
d_cross |> 
  ggplot(aes(x = Mono, y = GAP_bin)) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cr"),
              se = FALSE) +
  xlab("Percentage of monolinguals in country") +
  ylab("Proportion of accelerated ageing")
```

The bumps in this graph highlight a problem with Amoruso et al.'s analysis:
they didn't take the nested nature of the data into account.
To address this,
I've computed the number and proportion of respondents
showing accelerated ageing for each combination of country
and year of data collection.
@fig-by_country_year shows how the proportion of respondents
with accelerated ageing varied between different waves of data collection
within each country.

```{r}
#| label: fig-by_country_year
#| fig-cap: Each point represents a combination of country and year of data collection. The countries are sorted top to bottom, left to right, in increasing average proportion of respondents showing accelerated ageing.
#| fig-height: 6
#| fig-width: 6
#| out-width: 80%
by_country_year <- d_cross |>
  group_by(Country, Year, Mono, GDP_percapita) |>
  summarise(
    n = n(),
    n_accelerated = sum(GAP_bin > 0),
    prop_accelerated = mean(GAP_bin > 0),
    .groups = "drop"
  )

by_country_year |> 
  ggplot(aes(x = Mono, y = prop_accelerated)) +
  geom_point(shape = 1) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  facet_wrap(vars(reorder(Country, prop_accelerated))) +
  xlab("Proportion of monolinguals") +
  ylab("Proportion of accelerated ageing")
```

The proportion of monolinguals tends to vary fairly little within each country,
and I'm not convinced that we should put too much stock in the yearly changes
in these proportions.
So I'll simply aggregate the data per country instead.
@fig-country_mono shows that there is indeed a tendency for countries
with relatively more monolinguals to have a greater of respondents
with accelerated ageing.

```{r}
#| label: fig-country_mono
#| fig-cap: The percentage of monolinguals shown is the weighted average of the percentages of monolinguals in the country at the different waves of data collection.
#| fig-height: 3.5
#| fig-width: 4.9
#| out-width: 80%
by_country <- d_cross |>
  group_by(Country) |>
  summarise(
    n = n(),
    n_accelerated = sum(GAP_bin > 0),
    prop_accelerated = mean(GAP_bin > 0),
    mean_Mono = mean(Mono),
    mean_GDP = mean(GDP_percapita),
    .groups = "drop"
  )


by_country |> 
  ggplot(aes(x = mean_Mono, y = prop_accelerated)) +
  geom_text(aes(label = Country), size = 3) +
  xlim(0, 65) +
  xlab("Percentage of monolinguals") +
  ylab("Proportion of accelerated ageing")
```

@fig-country_gdp shows the association of the proportion of respondents
per country with accelerated ageing and the country's per capita GDP.

```{r}
#| label: fig-country_gdp
#| fig-cap: The GDP per capita is the weighted average of the country's GDP per capita at the different waves of data collection.
#| fig-height: 3.5
#| fig-width: 4.9
#| out-width: 80%
#| classes: preview-image
by_country |> 
  ggplot(aes(x = mean_GDP, y = prop_accelerated)) +
  geom_text(aes(label = Country), size = 3) +
  xlab("GDP per capita (in 10k)") +
  xlim(0, 12.5) +
  ylab("Proportion of accelerated ageing")
```

@fig-mono_gdp shows the association between the per capita GDP 
and the proportion of monolinguals in the country.

```{r}
#| label: fig-mono_gdp
#| fig-cap: The GDP per capita and the proportion of monolinguals are the weighted average of the country's values at the different waves of data collection.
#| fig-height: 3.5
#| fig-width: 4.9
#| out-width: 80%
by_country |> 
  ggplot(aes(x = mean_GDP, y = mean_Mono)) +
  geom_text(aes(label = Country), size = 3) +
  xlab("GDP per capita (in 10k)") +
  xlim(0, 12.5) +
  ylab("Proportion of monolinguals")
```

### Refitting the models
To take the nested nature of the data into account,
we could fit a binomial model to the data aggregated by country.
I expressed the monolingualism variable in such a way that a one-unit increase
in this variable corresponds to a difference of 10 percentage points in monolingualism.
I also centred this variable at its median, corresponding to about 19% monolinguals.

```{r}
by_country$c.Mono10 <- by_country$mean_Mono / 10
by_country$c.Mono10 <- by_country$c.Mono10 - median(by_country$c.Mono10)
mono.glm <- glm(cbind(n_accelerated, n - n_accelerated) ~ c.Mono10,
                data = by_country,
                family = binomial(link = "logit"))
summary(mono.glm)
```

The problem with this model is the _massive_ overdispersion: $\chi^2 = 2023.3/24 = 84.3 \gg 1$.
The binomial logistic model assumes that the number of respondents showing
accelerated ageing in the $i$-th country is
$$Y_i \sim \textrm{Binomial}(n_i, \pi_i),$$
where $n_i$ is the number of respondents in the $i$-th country and 
$\pi_i$ is the probability parameter modelled as
$$\pi_i = \textrm{Logistic}(\beta_0 + \beta_1x_i) = \frac{1}{1 + \exp({-(\beta_0 + \beta_1x_i)})},$$
with $x_i$ the percentage of monolinguals in the $i$-th country.

Clearly, this model underspecifies the true data-generating mechanism.
In normal (ordinary least squares) regression models, this isn't a huge 
problem in and of itself: 
the error term doesn't depend on the modelled mean and
can absorb the variance associated with unmodelled factors.

But the binomial model does not _have_ a separate error term.
Instead, it assumes that the variance around $Y_i$ is the variance
of the binomial distribution, i.e., $\textrm{Var}(Y_i) = n_i\pi_i(1-\pi_i)$.
To the extent that other factors beyond the ones modelled
contribute to the outcome, $\textrm{Var}(Y_i)$ can be considerably
larger than $n_i\pi_i(1-\pi_i)$,
and the standard errors and confidence intervals need to be widened to
take this into account.

One option is to fit the data in a quasibinomial regression model instead.
The resulting parameter estimates are the same as before, 
but the standard error increased by a factor of about 9.

```{r}
mono.glm <- glm(cbind(n_accelerated, n - n_accelerated) ~ c.Mono10,
                data = by_country,
                family = quasibinomial(link = "logit"))
summary(mono.glm)
mono.conf <- confint(mono.glm)
mono.conf[1, ] |> plogis()
mono.conf[2, ] |> exp()
```

The correct interpretation of these estimates is as follows.
If the model for the $\pi_i$s is correct,
we'd expect that in a country with the median percentage of monolinguals (i.e., about 19%),
about $\textrm{Logistic}(-0.19509) \approx 45\%$
of the respondents show accelerated ageing; 95% CI: $(42\%, 48\%)$.
A 10-percentage point difference in the percentage of monolinguals
is associated with an odds ratio of about $\exp(0.09992) \approx 1.105$;
95% CI: $(1.02, 1.19)$.

Let's now add per capita GDP to the equation.
The per capita GDP variable was centred at its sample median
of about 26,800 units per capita of whatever currency the data are expressed in.

```{r}
by_country$c.mean_GDP <- by_country$mean_GDP - median(by_country$mean_GDP)
gdp.glm <- glm(cbind(n_accelerated, n - n_accelerated) ~ c.Mono10 +
                 c.mean_GDP,
               data = by_country,
               family = quasibinomial(link = "logit"))
summary(gdp.glm)
gdp.conf <- confint(gdp.glm)
gdp.conf[1, ] |> plogis()
gdp.conf[2, ] |> exp()
```

The model predicts a country with a per capita GDP of about 26.8k 
and 19% monolinguals to have about 
$\textrm{Logistic}(-0.1174) \approx 47\%$ of respondents with accelerated ageing;
95% CI: $(43.7\%, 50.4\%)$.
Keeping per capita GDP constant, a 10-percentage point difference in monolingualism
is associated with $\exp(0.06063) \approx 1.06$ times larger odds of 
accelerated ageing in the country, 95% CI: $(0.98, 1.15)$.
Hence, once GDP is taken into account, the association
between accelerated ageing and monolingualism is no longer significant.

Alternative model specifications are, of course, possible.
Some, for instance generalised linear mixed models
using individual-level BAGs but country-level predictors,
lead to parameter estimates that can be fairly difficult to wrap one's head round.
Moreover, here I only looked at a single possible confounder (per capita GDP)
as I had earlier identified a problem with it.
This doesn't mean that _only_ this confounder should be considered.

### Using a continuous outcome
Up till now, I used the $\texttt{BAG}$ values that Amoruso et al. provided
in the `data.csv` dataset.
To recall (see @sec-bag), 
these values are binary indicators of whether the respondent's predicted
age is older than their actual age, after correcting for the
overprediction of younger ages and the underprediction of older ages.
The predicted ages aren't actually shared in any of the datasets:
Amoruso et al. fit the models predicting age from biobehavioural factors
in a couple of scripts, but these don't contain the code for exporting
the predicted ages.
So I adapted their Jupyter notebooks in order to compute and export
the predicted ages myself.

I stuck to Amoruso et al.'s approach, even though I'm pretty skeptical about it.
Based on their Python code, I wrote my own Python script, [`compute_bag.py`](compute_bag.py).
In it, 26 models are trained using gradient boosting to predict
respondents' age from their biobehavioural factors, always leaving out
a different country. 
I excluded the Slovak data seeing as these did not enter into Amoruso et al.'s
analyses.
For each model, a hyperparameter search was conducted.
The ages for the out-of-fold respondents were then predicted,
and their $\texttt{GAP}$ scores computed accordingly.
To correct for the overestimation of the ages of the youngest participants
and the underestimation of those of the oldest participants,
I fitted linear regression models just like in the original
to the training data.
The corrected $\texttt{GAP}$ scores were computed as in the original.
One additional difference to the original is that didn't set all
the random seeds to 42.

The predicted ages and $\texttt{GAP}$ values are stored
in [`data_predictions.csv`](data_predictions.csv).
Let's first reassure ourselves that the model output is comparable to that
reported by Amoruso et al.
As the scatterplot in @fig-predicted-ages shows, the prediction is far from good.
Up till age 66, the respondents' age is overestimated;
from then onwards, their age is underestimated.
```{r}
#| label: fig-predicted-ages
#| fig-cap: The predicted ages are out-of-fold predictions based on gradient boosting. The red lines shows the identity; the blue line is a scatterplot smoother highlighting the mean trend.
#| fig-height: 4.5
#| fig-width: 6.3
#| out-width: 80%
d <- read.csv("data_predictions.csv")

ggplot(d,
       aes(x = Age, y = predicted_age)) +
  geom_boxplot(aes(group = Age), outlier.shape = 1, varwidth = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid", colour = "red") +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "tp")) +
  xlab("True age (years)") +
  ylab("Predicted age (years)") +
  labs(
    title = "Predicted vs. true age"
  )
```

The root mean squared error for the age prediction is 8.71 years,
which is similar to the number reported by Amoruso et al. 
(8.65 in the leave-one-country-out cross-validation).
The mean absolute error is 7.15 years, 
again similar to the number reported in the article (7.09).
The mean directional error is about 0.03 years 
(an average overestimation of about 12 days);
which is similar to the number reported in the article (0.03).
```{r}
sqrt(mean((d$predicted_age - d$Age)^2))
mean(abs(d$predicted_age - d$Age))
mean(d$predicted_age) - mean(d$Age)
```

These differences can be explained by the fact that
I excluded Slovakia from the data when fitting the models; Amoruso et al. didn't.
Further, Amoruso et al. set all their random seeds to 42; I decided to leave them unspecified by default.


Let's now take a look at the corrected $\texttt{GAP}$ values,
@fig-gap.
Note that a slight nonlinear trend remains.
```{r}
#| label: fig-gap
#| fig-cap: The GAP values after correction for linear under- and overestimation.
#| fig-height: 4.5
#| fig-width: 6.3
#| out-width: 80%
ggplot(d,
       aes(x = Age, y = GAP_corrected)) +
  geom_boxplot(aes(group = Age), outlier.shape = 1, varwidth = TRUE) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "tp")) +
  xlab("True age (years)") +
  ylab("Corrected GAP (years)") +
  labs(
    title = "Corrected GAP vs. true age"
  )
```

While the datasets `data.csv` and `data_predictions.csv`
do specify the respondents' countries,
they don't specify which year the data was collected in.
For this reason, I calculated mean monolingualism and mean GDP per capita
values per country and added these to the dataset.

```{r}
country_cohort <- read.csv("country_cohort.csv") |> 
  mutate(GDP = (GDP / Pop) / 10000) |> 
  group_by(Country) |> 
  summarise(
    mean_Mono = weighted.mean(Mono, n),
    mean_GDP = weighted.mean(GDP, n),
    .groups = "drop"
  ) |> 
  mutate(
    c.Mono10 = (mean_Mono / 10) - median(mean_Mono / 10),
    c.GDP = mean_GDP - median(mean_GDP)
  )
d <- d |>
  left_join(country_cohort, join_by("country" == "Country"))
```

I'll go through a similar analysis as before
but this time without discretising the `GAP_corrected` variable.
@fig-mono-gap confirms that there is some association
between the percentage of monolinguals in a country
and the average corrected GAP value in this country.

```{r}
#| label: fig-mono-gap
#| fig-cap: The percentage of monolinguals shown is the weighted average of the percentages of monolinguals in the country at the different waves of data collection. 
#| fig-height: 3.5
#| fig-width: 4.9
#| out-width: 80%
by_country <- d |>
  group_by(country, mean_Mono, mean_GDP, c.Mono10, c.GDP) |>
  summarise(
    n = n(),
    mean_GAP = mean(GAP),
    mean_corrected_GAP = mean(GAP_corrected),
    mean_Age = mean(Age),
    mean_predicted_Age = mean(predicted_age),
    .groups = "drop"
  )


by_country |> 
  ggplot(aes(x = mean_Mono, y = mean_corrected_GAP)) +
  geom_text(aes(label = country), size = 3) +
  xlim(0, 65) +
  xlab("Percentage of monolinguals") +
  ylab("Mean corrected GAP (years)")
```

As @fig-gdp-gap shows, per capita GDP is likewise associated
with the average GAP value per country.

```{r}
#| label: fig-gdp-gap
#| fig-cap: The GDP per capita is the weighted average of the country's GDP per capita at the different waves of data collection.
#| fig-height: 3.5
#| fig-width: 4.9
#| out-width: 80%
by_country |> 
  ggplot(aes(x = mean_GDP, y = mean_corrected_GAP)) +
  geom_text(aes(label = country), size = 3) +
  xlim(0, 12.5) +
  xlab("Per capita GDP") +
  ylab("Mean corrected GAP (years)")
```

We don't need to worry about logistic models and odds ratios this time.
A simple OLS regression model at the country level is all we need.

```{r}
mono.lm <- lm(mean_corrected_GAP ~ c.Mono10, data = by_country)
summary(mono.lm)
confint(mono.lm) * 365.25
```

This country predicts country with 19% monolinguals
to have an average corrected GAP value of about $-0.22$ years (about $-80$ days),
95% CI: $(-217, 54)$ days.
A 10-percentage point difference in the extent of monolingualism is associated
with a change in the average corrected GAP value of about $0.36$ years (about 133 days),
95% CI: $(58, 207)$ days.

We can also control for per capita GDP:

```{r}
gdp.lm <- lm(mean_corrected_GAP ~ c.Mono10 + c.GDP, data = by_country)
summary(gdp.lm)
confint(gdp.lm) * 365.25
```

So, this model predicts that a country with a per capita GDP of about 26.8k
and 19% monolinguals to have an average corrected GAP of about $-0.09$ years
(about $-32$ days), 95% CI: $(-182, 119)$ days).
Keeping per capita GDP constant, it predicts a 10-percentage point change
in the extent of monolingualism to be associated with 
a change in the average corrected GAP of about $0.287$ years (some 105 days),
95% CI: $(22, 188)$ days.

Again, I only took into consideration per capita GDP because Amoruso et al. didn't.
This should in no way be taken to mean that I think that _only_ per capita GDP 
should be taken into account as a possible confounder.

## Now what?
Suffice it to say that I think the authors ought to substantially revise their article,
both in terms of how the analyses are conducted and interpreted and in terms
of how the overall method is described.
The analyses that I carried out using per capita GDP (as opposed to overall GDP)
as a control variable are a start---not a suggestion of what the final model should look
like.
Further, I've only looked at the cross-sectional analyses as I think
my comments on these already ought to prompt readers to take the entire
study with a grain of salt.
This doesn't mean that I think that the longitudinal analyses 
that are also reported in the article are trustworthy---I've literally not looked at them.
Finally, we've taken the authors' method for quantifying accelerated ageing as-is,
but I am somewhat skeptical of it.

**Update 2026/01/09:** A slightly edited
version of this blog post is now available
as a preprint from [PsyArxiv Preprints](https://osf.io/preprints/psyarxiv/rmnv9_v1).
The main change is that I log-transformed the per-capita GDP data for the preprint.

## Software versions
```{r, warning = FALSE}
devtools::session_info("attached")
```
