---
draft: true
title: "Cluster analysis: A skeptic's guide"
date: "2026-09-18"
categories: [correlational studies, research questions]
format:
  html:
    code-fold: show
    message: false
    df-print: paged
    fig-width: 6
    fig-height: 9
    fig-align: center
    echo: false
---

I've been seeing more and more articles in applied linguistics
that use clustering techniques to identify subgroups
("profiles", "classes", "types") of language learners
on the basis of cognitive test or questionnaire data.
Often, the learners are then classified into clusters,
and cluster membership is used as a predictor of an
outcome such as performance on a language test.
I have yet to see a study in applied linguistics that convinces 
me that identifying and interpreting learner clusters adds value 
over treating learner differences as continuous.
I'll explain why that is.

## Refresher: What is cluster analysis?
Cluster analysis refers to a host of unsupervised learning techniques.
That's technobabble for saying that 
we're given lots of multivariate observations and 
we're looking for a sensible way to group these observations 
into a handful of categories.
Crucially, there are no observations 
for which we _know_ which category they belong to.
This makes the problem a different one from supervised learning,
which is useful in situations in which you know the correct label
for at least some of the observations.

There are a _lot_ of clustering algorithms, including
_k_-means clustering, Gaussian mixture models, 
and latent profile and latent class analysis.
It's up to the user to specify which variables the clustering should be based
on, to preprocess the data, and to set the desired number of clusters.
The algorithms go about the task of identifying the desired number of clusters
and of assigning the observations to these clusters in rather different ways
and are bound to end up with different cluster assignments.
I'll come back to this point below.

## What are the clusters for?
Bauer and Curran (2004) distinguish between direct and indirect uses 
of cluster analysis. 
In **direct uses** of cluster analysis, the clusters themselves are of interest:
researchers may want to figure out how many clusters there are,
what proportion of observations (e.g., learners) belong to each cluster,
where these clusters are located in the multivariate data space, etc.
Cluster analysis, then, is used to discover some interesting structure
in the data.

By contrast, in **indirect uses** of cluster analysis, 
the researchers don't really care about the clusters themselves.
It may seem a bit strange at first to do cluster analysis if you don't really
care about the clusters.
But let's say you're dealing with a wonky data distribution $P$
that is difficult to handle mathematically.
Instead of handling $P$ itself, however, you can try to 
first approximate $P$ using a distribution $\widehat P$
that is a combination of distributions $Q_1, \dots, Q_C, C \geq 2$, 
that are easier to handle: 
$$P \approx \widehat P = \sum_{c = 1}^C\pi_cQ_c,$$
with weights $\pi_c > 0, c = 1, \dots C$, that satisfy $\sum_{c=1}^C \pi_c = 1$.
Distributions like $\widehat P$ are called **mixture distributions**,
and if $Q_1, \dots, Q_C$ are Gaussian (i.e., Normal) distributions,
they are called **Gaussian mixtures**.
But neither the weights $\pi_c$ nor the mixture components $Q_c$
themselves are of any particular interest: 
they merely make it easier to do useful stuff related to $P$,
and there isn't necessarily any implication that the mixture components
(i.e., the clusters) carry any physical or cognitive meaning.
For a nice example of such an indirect use of clustering algorithms,
see Faul et al. (2026).

Indirect uses of clustering are fairly forgiving.
Since the goal is merely to obtain a mathematically tractable approximation to $P$, 
it is not necessarily a problem if different clustering solutions 
partition the distribution in rather different ways,
as they do in @fig-different-solutions.
In this case, the choice between competitive approximations may be guided by convenience:
in some applications, solutions with fewer components may be preferred;
in others, solutions with components of similar sizes or with components
with diagonal covariance matrices may be easier to handle.

```{r}
#| label: fig-different-solutions
#| fig-width: 8
#| fig-height: 3.8
#| fig-cap: "Two different but equally well-fitting Gaussian mixtures applied to the same data. In the left panel, the solution has two Gaussian components, within each of which the two variables are somewhat correlated. In the right panel, the solution has three Gaussian components, within each of which the variables are not correlated."

suppressPackageStartupMessages({
  library(MASS)
  library(mclust)
})

squish <- function(x) (x - min(x))/(max(x) - min(x))

set.seed(17+09+2026)

par(las = 1)
op <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), cex = 0.4)

squish <- function(x) (x - min(x))/(max(x) - min(x))
to_unif <- function(x) rank(x)/length(x)

n <- 500
d1 <- mvrnorm(n * 0.7, c(0, 0), Sigma = rbind(c(1, -0.4), c(-0.4, 1)))
d2 <- mvrnorm(n * 0.3, c(1, 1), Sigma = rbind(c(0.5, 0.1), c(0.1, 0.6)))
d <- rbind(d1, d2)
d[, 1] <- squish(d[, 1])
d[, 2] <- squish(d[, 2])
colnames(d) <- c("X", "Y")

op <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
gmm <- Mclust(d, 1:9, "VVV")
plot(gmm, "classification")

lap <- Mclust(d, 1:9, "VEI")
plot(lap, "classification")
par(op)
```

The situation is very different for direct uses of clustering.
If we want to identify qualitatively distinct types of language learners
based on some theoretical considerations,
then a solution with two elliptical components (as in the left plot above) 
would have different consequences from a solution with three 
diagonal components (as in the right plot above).
Further, even if competitive solutions have the same number of components,
the shapes and sizes of the clusters may differ considerably between them,
and learners may be assigned to different clusters in different solutions.
If the cluster analysis serves a direct use, principled reasons for preferring
one solution over another are needed.

Hand (2026, chapter 7) makes a similar distinction between using 
cluster analysis for **discovery**, where the clusters are interpreted
substantively, and **imposition**, where the clusters are used 
as a compression technique for practical purposes.

In the social sciences, including applied linguistics, I suspect
that most intended uses of cluster analysis are direct and discovery-oriented,
in which case a few additional questions need to be addressed.

## What clusters, exactly?
When explaining what cluster analysis does,
I refrained from defining what clusters are.
The reason is simple: There is no generally accepted definition of a cluster.

Hennig (2015) provides an excellent discussion of different notions of clusters 
and how these result in different suitable clustering techniques.
Hand (2026, chapter 7) gives an example in which two different ways of 
conceptualising clusters results in partitionings that could hardly be any more different:

- In some applications, researchers may wish to minimise the maximum distance
  between observations assigned to the same cluster. This tends to result
  in fairly compact clusters that may, however, not have any separation between
  them.
- In other settings, researchers may prefer clusters in which each observation
  belongs to the same cluster as its closest neighbour. This may result
  in elongated clusters with a clean gap between them.

Since there is no universal definition of what a cluster is,
none of these solutions are wrong per se.[^1]
But this makes it all the more important for researchers to define
formally what their definition of a cluster is,
discuss how this definition is meaningful in light of the substantive
theory,
and argue why the clustering algorithm chosen is suitable for detecting
the clusters so defined.

[^1]: In simulation studies, researchers often generate data from a known
      mixture model and study how well different algorithms can recover the
      components (e.g., Toffalini et al. 2022). In doing so, these researchers
      define the clusters to be the components of the generating mixture model.
      But this doesn't mean that cluster solutions that fail to recover the
      components are wrong: such solutions may be recovering
      clusters according to a different, yet equally sensible, definition
      of what a cluster is.
      
Admittedly, this is a tall order.
But it is crucially important for users of clustering algorithms to appreciate
how well their own notion of clusters aligns with that of the algorithm they've
chosen.
By way of an example, consider **Latent Profile Analysis** (LPA).
As Bauer and Curran (2004) explain, LPA is based the assumption
that the $d$-dimensional distribution of the variables
$\boldsymbol Y = (Y_1, \dots, Y_d)$ is a Gaussian mixture
with $C \geq 1$ components, 
_within each of which the variables are independent!_
Formally,

$$
\begin{align*}
  \boldsymbol Y &\sim P, \\
  P &= \sum_{c=1}^C\pi_c \mathcal{N}_d\left(\boldsymbol \mu_c, 
    \begin{pmatrix}
    \sigma^2_{c1} & 0 & \cdots & 0 \\
    0             & \sigma^2_{c2} & \ddots & 0 \\
    \vdots & \ddots & \ddots & \vdots \\
    0      &  0     & \cdots & \sigma^2_{cd}
    \end{pmatrix}\right)
\end{align*},
$$
where $\pi_c > 0, c = 1, \dots, C$, and $\sum_{c = 1}^C \pi_c = 1$.

This is quite the assumption, and I'd like to see researchers justify
explicitly that this is indeed what their own notion of clusters corresponds to.
Just because the goal of a study is to identify "latent profiles"
doesn't mean that a tool called "Latent Profile Analysis" is most appropriate
to the task.

## Does the pipeline work?
I agree with Harrell (2024) that the onus is on researchers to demonstrate
that their data contains meaningful clusters.
To this end, I'd like researchers, once they have clearly defined
what they consider to be clusters and once they have _argued_ 
why the tool they've chosen aligns well with this defintion,
to demonstrate that their entire analytical pipeline
does a reasonable job under realistic conditions.

Ideally, I'd like for researchers to demonstrate that their pipeline
works in a simulation study.
Having defined what clusters are to them, researchers can generate
realistic data sets in terms of their size, 
the number of clusters present,
the structure of these clusters,
and the granularity of the data that will be collected (e.g., using 5-point scales
instead of perfectly continuous variables).
They can then examine if their pipeline can reliably recover the intended clusters.

Equally importantly, researchers ought to address the question: 
_If our theory is wrong, would we know?_

For instance, Chen et al.'s (2026) first research question is
_Can L2 learners be profiled concerning their (...) strategy use and writing self-efficacy?_
But it's not clear to me how they could conceivably ever have answered _no_ to this question---unless
in the rather unrealistic scenario that none of their variables were correlated with one another.
So, I'd like researchers to also generate realistic data sets in which there are,
according to their own definition, no clusters, and see if their pipeline
will accordingly not partition the data into clusters.

Note that the point of such simulations is not merely to gauge 
whether the clustering algorithm chosen is appropriate.
It is the entire workflow that is put to the test,
including how the data are preprocessed and how the desired number
of clusters is determined, when applied to data that isn't infinitely fine-grained.

Incidentally, on the topic of choosing the desired number of clusters,
researchers typically compare different solutions and select a suitable
one on the basis of several criteria, including both numerical and 
substantive ones.
This may be quite sensible. But since this choice is rarely made algorithmically,
it becomes difficult to evaluate the statistical consequences of deciding 
on the desired number of clusters only once the data are in.
The simulation study should take this into account.

## What does the solution look like?
Once a defensible clustering has been obtained, I'd like to _see_ it.

For low-dimensional data, a scatterplot matrix showing all bivariate
relationships between the relevant variables in which the observations
are coloured for their cluster memberships or in which projections of the
mixture components are shown strikes me as a good start.
For high-dimensional data, dimension-reduction techniques such as
principal component analysis or multidimensional scaling may be
useful.

The purpose of the visualisation is two-fold.
First, it allows for visual inspection of whether the model assumptions
are plausible in the actual data.
Second, they allow readers to assess if their own notion of clusters
accords with that of the researchers and of the tool used.
For instance, @fig-lpa would allow readers and reviewers
to realise that the clusters identified by a Latent Profile Analysis
aren't nicely demarcated
and that they themselves might prefer not to partition the data
in this way---or indeed at all.

```{r}
#| label: fig-lpa
#| fig-width: 6
#| fig-height: 6
#| fig-cap: "A visualisation of a LPA fit."
set.seed(18+09+2026)
n <- 300
d <- mvrnorm(n, c(0, 0), Sigma = rbind(c(3, 0.7*sqrt(3)), c(0.8, 1)))
d[, 1] <- to_unif(d[, 1])
d[, 2] <- to_unif(d[, 2])
d[, 2] <- qbeta(d[, 2], 5, 5)
d[, 1] <- qbeta(d[, 1], 1.4, 3.4)
colnames(d) <- c("X", "Y")

par(las = 1)
lap <- Mclust(d, 1:9, "VEI")
plot(lap, "classification")
```

More visualisation may also help researchers and readers alike to realise, first,
that the group labels often attached to clusters ('high-aptitude, highly motivated learners')
may not be apt for all learners in the cluster, and second,
that clustering is fundamentally about carving up continuous data into categories.

## When running follow-up analyses, how is the uncertainty in the cluster assignments taken into account?
Researchers often carry out follow-up analyses with the cluster assignments.
For instance, Chen et al. (2026) compared the learners assigned to the different
clusters in terms of their L2 writing performance,
whereas Roehr-Brackin et al. (2024) wanted to see if cluster membership
interacted with a pedagogical intervention type when learning L2 Polish grammar.

But clustering algorithms don't assign observations to clusters with absolute
certainty: minor changes to the data elsewhere may cause some learners
to be assigned to a different cluster, and some clustering algorithms
reasonably assign cluster membership probabilistically.
If cluster membership is used as a predictor
in a follow-up ANOVA or regression model,
I'd like or researchers to take into account the uncertainty about the 
membership assignments.
Bakk and Kuha (2020) discuss the relevant literature.

## Was it worth it?
Even if all the previous questions have been satisfactorily answered,
I think researchers using clustering algorithms ought to address
what all their effort has actually bought us.

While some clustering algorithms (including Latent Profile Analysis)
assume that the observed data are generated by the (latent) clusters,
the _reconstructed_ clusters are a function of the observed data.
As a result, any cluster assignment _has_ to discard information in the data.
The question is not whether any information has been lost,
but how much useful information has been discarded.

For this reason, I think it is sensible, if cluster membership is used
to model a distal outcome (e.g., L2 writing in Chen et al.'s study),
for researchers to compare the resulting model with alternative models
in which the variables that the clustering is based on are used directly.
The latter models have to fit at least equally well as the former one does;
the question becomes how much worse the cluster-based model fares.
The reason why I think this is sensible is that my default assumption is
that cognitive and attitudinal predictors affect outcomes such as L2 performance
in a continuous rather than in a discrete way.

## Conclusion
I appreciate that the questions asked above set a high bar.
This is a consequence of the lack of a consensus of what clusters are in the first place,
which ought to force researchers to clearly outline 
what their notion of clusters is,
to argue why this notion is theoretically sensible,
and to demonstrate that their analytical pipeline reconstructs clusters
so defined if they are present and does not identify them if they are
absent.
This, I hope, should strike researchers wishing to apply clustering techniques
to their own data as reasonable asks.
Some of the other questions I've put forward are motivated by my own stance
that cognitive and attitudinal predictors tend to have continuous rather
than discrete effects on language learning outcomes.

## References
References prefixed with an asterisk are recommended reading for 
applied linguists and other social scientists thinking about applying
clustering techniques to identify hidden classes in their data.

Bakk, Zsuzsa & Jouni Kuha.
2020.
[Relating latent class membership to external variables: An overview.](https://doi.org/10.1111/bmsp.12227)
_British Journal of Mathematical and Statistical Psychology_ 74.
340--362.

*Bauer, Daniel J. & Patrick J. Curran.
2004.
[The integration of continuous and discrete latent variable models: Potential problems and promising opportunities.](https://doi.org/10.1037/1082-98X.9.1.3)
_Psychological Methods_ 9(1).
3--29.

Chen, Jing, Lawrence Jun Zhang & Xiaotong Chen.
2026.
[L2 learners' self-regulated learning strategies and self-efficacy for writing achievement: A latent profile analysis.](https://doi.org/10.1177/13621688221134967)
_Language Teaching Research_ 30(1).
14--35.

Faul, Antoine, David Ginsbourger & Ben Spycher.
2026.
[Easy conditioning far beyond Gaussian.](https://doi.org/10.48550/arXiv.2409.16003)
_arXiv_ 2409.16003.

*Hand, David J.
2006.
[_What's the question? Deciding what you really want to know._](https://doi.org/10.1201/9781003726821)
Boca Raton, FL: CRC Press.

Harrell, Frank.
2024.
[The burden of demonstrating statistical validity of clusters.](https://www.fharrell.com/post/cluster/)
_Statistical Thinking_ (blog), 6 October 2024.

*Hennig, Christian.
2015.
[What are the true clusters?](https://doi.org/10.1016/j.patrec.2015.04.009)
_Pattern Recognition Letters_ 64.
53--62.

Roehr-Brackin, Karen, Karoline Baranowska, Renato Pavlekovic & Paweł Scheffler.
2024.
[The role of individual learner differences in explicit language instruction.](https://doi.org/10.1111/modl.12963)
_The Modern Language Journal_ 108.
815--845.

Toffalini, Enrico, Paolo Girardi, David Giofrè & Gianmarco Altoè.
2022.
[_Entia non sunt multiplicanda…_ Shall I look for clusters in my cognitive data?](https://doi.org/10.1371/journal.pone.0269584)
_PLoS ONE_ 17(6).
e0269584.
