---
title: "R tip: Ordering factor levels more easily"
date: "2016-08-18"
categories: [R, graphics]
---

By default, `R` sorts the levels of a factor alphabetically. When drawing graphs, this results in ['Alabama First'](http://dx.doi.org/10.1080/09332480.2001.10542269) graphs, and it's usually better to sort the elements of a graph by more meaningful principles than alphabetical order. This post illustrates three convenience functions you can use to sort factor levels in `R` according to another covariate, their frequency of occurrence, or manually.

**Update (2023-08-08):** The `reorder()` function has pretty much the same functionality as the functions introduced in this blog post. In the original blog post, I loaded some packages that are now part of the `tidyverse` suite separately; now I just use the `tidyverse`.

First you'll need the `tidyverse`:

```{r, eval = FALSE}
install.packages(c("tidyverse"))
```

You can download the convenience functions from my [Github page](https://github.com/janhove/janhove.github.io/blob/master/RCode/sortLvls.R) or read them in directly into `R`:

```{r}
source("https://janhove.github.io/RCode/sortLvls.R")
```

## Sorting factor levels by another variable

The code below creates an example dataset with a factor and a covariate:

```{r, message = FALSE}
# Load packages
library(tidyverse)

# Generate same data
set.seed(18-08-2016)

# Create data frame
df <- data.frame(factorBefore = factor(rep(letters[1:5], 3)),
                 covariate = rnorm(15, 50, 30))

# Current order of factor levels (alphabetically)
levels(df$factorBefore)

# Covariate mean per factor level
df |> group_by(factorBefore) |> summarise(mean(covariate))
```

What we want is to sort the levels of the factor by the covariate mean per factor level (i.e., a-b-e-c-d). The function `sortLvlsByVar.fnc` accomplishes this:

```{r}
# Reorder
df$factorAfter1 <- sortLvlsByVar.fnc(df$factorBefore, df$covariate)

# New order of factor levels
levels(df$factorAfter1)
```

By setting the `ascending` parameter to `FALSE`, the factor levels are sorting descendingly according to the covariate mean:

```{r}
# Reorder descendingly
df$factorAfter2 <- sortLvlsByVar.fnc(df$factorBefore, df$covariate, ascending = FALSE)
levels(df$factorAfter2)
```

How this looks like when graphed:

```{r, fig.width = 6, fig.height = 2, message = FALSE}
# Alphabetical order
p1 <- ggplot(df, aes(x = factorBefore, y = covariate)) +
  geom_boxplot()
# Sorted ascendingly
p2 <- ggplot(df, aes(x = factorAfter1, y = covariate)) +
  geom_boxplot()
# Sorted descendingly
p3 <- ggplot(df, aes(x = factorAfter2, y = covariate)) +
  geom_boxplot()
gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

You can change the `R` code from the Github page so that the levels are sorted by another summary statistics, e.g., the covariate median per factor level.

## Sorting factor levels by their frequency of occurrence

Again we'll first create some data:

```{r, fig.width = 6, fig.height = 2, message = FALSE}
df2 <- data.frame(factorBefore = factor(rep(letters[1:5], c(7, 3, 80, 15, 107))),
                  covariate = rnorm(sum(c(7, 3, 80, 15, 107)), 50, 30))
table(df2$factorBefore)
```

We want to order these factor levels by their frequency of occurrence in the dataset (i.e., b-a-d-c-e). `sortLvlsByN.fnc()` accomplishes this:

```{r, fig.width = 6, fig.height = 2, message = FALSE}
df2$factorAfter1 <- sortLvlsByN.fnc(df2$factorBefore)
table(df2$factorAfter1)
```

Or descendingly:

```{r, fig.width = 6, fig.height = 2, message = FALSE}
df2$factorAfter2 <- sortLvlsByN.fnc(df2$factorBefore, ascending = FALSE)
table(df2$factorAfter2)
```

When plotted:

```{r, fig.width = 6, fig.height = 2, message = FALSE}
p4 <- ggplot(df2, aes(x = factorBefore, y = covariate)) +
  geom_boxplot(varwidth = TRUE)
p5 <- ggplot(df2, aes(x = factorAfter1, y = covariate)) +
  geom_boxplot(varwidth = TRUE)
p6 <- ggplot(df2, aes(x = factorAfter2, y = covariate)) +
  geom_boxplot(varwidth = TRUE)
gridExtra::grid.arrange(p4, p5, p6, ncol = 3)
```

## Customising the order of factor levels

If you want to put the factor levels in a custom order, you can use the `sortLvls.fnc()` function.

```{r}
# Create data
df3 <- data.frame(factorBefore = factor(rep(letters[1:5], 3)),
                  covariate = rnorm(15, 50, 30))
levels(df3$factorBefore)
```

Let's say we, for some reason, want to put the current 5th level (e) first, the current 3rd level (c) second, the 4th 3rd, the 4th 2nd and the 1st last:

```{r}
df3$factorAfter1 <- sortLvls.fnc(df3$factorBefore, c(5, 3, 4, 2, 1))
levels(df3$factorAfter1)
```

You can also just specify which factor levels need to go up front; the order of the other ones stays the same:

```{r}
# Put the current 3rd and 2nd in front; leave the rest as they were:
df3$factorAfter2 <- sortLvls.fnc(df3$factorBefore, c(3, 2))
levels(df3$factorAfter2)
```

## Software versions

```{r}
devtools::session_info()
```
