---
title: "Exact significance tests for 2 × 2 tables"
date: "2024-09-10"
categories: [R, significance]
---

Two-by-two contingency tables look so simple 
that you'd be forgiven for thinking they're straightforward to analyse. 
A glance at the statistical literature on the analysis of contingency tables, 
however, reveals a plethora of techniques and controversies surrounding them 
that will quickly disabuse you of this notion 
(see, for instance, Fagerland et al. 2017). 
In this blog post, I discuss a handful of different study designs that give
rise to two-by-two tables and present a few exact significance tests that can be
applied to these tables.
A more exhaustive overview can be found in Fagerland et al. (2017).

## Preliminaries

### Two-by-two contingency tables
Two-by-two contingency tables arise when cross-tabulate obserations that
have two binary properties ($X$: $x_1$ vs. $x_2$; $Y$: $y_1$ vs. $y_2$). 
The number of obervations for which $X = x_1$ and $Y = y_1$ is referred to
as $n_{11}$, and similarly for the other combinations of $X$ and $Y$ values.
We further define the row totals $n_{1+} := n_{11} + n_{12}$
and $n_{2+} := n_{21} + n_{22}$
as well as the column totals $n_{+1} := n_{11} + n_{21}$
and $n_{+2} := n_{12} + n_{22}$.
We call the total number of observations $n_{++}$.

\begin{array}{l|c|c||c}
  & Y = y_1 & Y = y_2 & \textrm{Row total} \\
  \hline
  X = x_1 & n_{11} & n_{12} & n_{1+} \\
  X = x_2 & n_{21} & n_{22} & n_{2+} \\
  \hline
  \textrm{Column total} & n_{+1} & n_{+2} & n_{++} \\
\end{array}

### Exact and approximate tests
This blog post is about **exact significance tests**.
An exact test has the following defining property:
If the null hypothesis is true, the $p$-value that the test yields is
a random variable $P$ with the property that
$$\mathbb{P}(P \leq \alpha) \leq \alpha$$
for each $\alpha \in (0, 1)$.
This means that, if the null hypothesis is true, 
there is at most a 5% probability that the test will yield a $p$-value of 0.05 or below.
Similarly, if the null hypothesis is true, 
there is at most a 20% probability that the test will yield a $p$-value of 0.20 or below,
and so for all values between 0 and 1.
Ideally, $\mathbb{P}(P \leq \alpha)$ should be close to, but no larger than, $\alpha$:
If $\mathbb{P}(P \leq \alpha)$ tends to be considerably lower than $\alpha$,
i.e., if the test is **conservative**, this will negatively affect the test's power.

Tests that aren't exact can still be **approximate**.
A possible problem with approximate tests is that their justification depends
on results derived for large samples; 
for smaller samples, $\mathbb{P}(P \leq \alpha)$
may be substantially larger than $\alpha$.
The best-known approximate test used for the analysis of two-by-two contingency
tables is Pearson's $\chi^2$-test.

## Contingency tables with both marginals fixed
Two-by-two contingency tables can be the result of different research designs --
some fairly common, others exceedingly rare.

**Example 1** (Fisher's exact test, one-sided)**.** 
Say we want to establish if a learner of English is able to tell the /æ/ phoneme
in _bat_ and the /ɛ/ phoneme in _bet_ apart.
To this end, we make 35 recordings, 
21 of which contain the word _bet_ and 14 contain the word _bat_.
The learner is then asked to identify those 21 audio files that he thinks are
recordings of _bet_; the remaining 14 audio files are suspected recordings of _bat_.
The results are summarised in the following contingency table:

\begin{array}{l|cc}
  & \textrm{Categorised as \textit{bet}} & \textrm{Categorised as \textit{bat}}  \\
  \hline
  \textrm{Is \textit{bet}} & 15 & 6 \\
  \textrm{Is \textit{bat}} & 6  & 8 \\
\end{array}


Note that we insisted that the learner select exactly 21 suspected recordings 
of _bet_, no more and no fewer.
As a result, the column total $n_{+1}$ was known in advance.
Moreover, we knew beforehand that 21 of the recordings actually featured _bet_
and 14 of the recordings actually featured _bat_. Hence, the row
totals $n_{1+}, n_{2+}$ were also known in advance.
As a result, $n_{++}$ and $n_{+2}$ were also known in advance.
Consequently, this study design fixes both the row and column marginals.
What wasn't known in advance was the number of suspected _bet_ recordings
that actually were _bet_ recordings ($n_{11}$).

The null hypothesis in this setting is that the learner is incapable of 
distinguishing _bet_ from _bat_ recordings and just selected 21 random
audio files as suspected _bet_ recordings in order to comply with the instructions.
Under this null hypothesis, 
the top left entry in the contingency table ($n_{11}$) follows a hypergeometric
distribution with parameters $n_{1+} = 21, n_{2+} = 14$ and $n_{+1} = 21$.
(Different authors parametrise this distribution differently;
I use the parametrisation that's used in R.)
That is,

$$H_0 : n_{11} \sim \textrm{Hypergeometric}(\underbrace{21}_{\textrm{is bet}}, \underbrace{14}_{\textrm{is bat}}, \underbrace{21}_{\textrm{to be categorised as bet}}).$$

@fig-hypergeometric shows the probability mass and cumulative probability
functions of the $\textrm{Hypergeometric}(21, 14, 21)$ distribution.
The observed top-left entry, i.e., 15, is highlighted in blue.

```{r, fig-hypergeometric, echo = FALSE, fig.cap = "The probability mass function (left) and the cumulative probability function (right) of the $\\textrm{Hypergeometric}(21, 14, 21)$ distribution.", fig.height = 2.8, fig.width = 7, out.width = "80%"}
par(las = 1,
    bty = "l",
    mar = c(3,3.5,2,1),
    mgp = c(2,0.3, 0),
    tck = -.01,
    cex = 0.8,
    mfrow = c(1, 2))

plot(0:21, dhyper(0:21, 21, 14, 21), pch = 16, 
     xlab = bquote(k), ylab = expression(Prob(n[11] == k)))
abline(v = 15, lty = "dashed", col = "blue", lwd = 2)

plot(0:21, phyper(0:21, 21, 14, 21), pch = 16, 
     xlab = bquote(k), ylab = expression(Prob(n[11] <= k)))
segments(-1, phyper(15, 21, 14, 21),
         15, lty = "dashed", col = "blue", lwd = 2)
segments(15, phyper(15, 21, 14, 21),
         y1 = -1, lty = "dashed", col = "blue", lwd = 2)
```

If the learner did not just pick 21 audio files at random but was in fact
able to tell _bet_ and _bat_ recordings apart to some degree,
this top-left entry can be expected to be large as opposed to small.
This means that we want to compute a right-sided $p$-value,
which we do by calculating 

$$\mathbb{P}(n_{11} \geq 15) = 1 - \mathbb{P}(n_{11} \leq 14):$$

```{r}
1 - phyper(15 - 1, 21, 14, 21)
```

This computation amounts to running **Fisher's exact test**:

```{r}
tab <- rbind(c(15, 6), c(6, 8))
fisher.test(tab, alternative = "greater")$p.value
```

**Example 2** (Fisher's exact test, two-sided)**.** 
Let's slightly change the design of the study in Example 1.
Instead of recording _bet_ 21 times and _bat_ 14 times and asking the learner
to select 21 suspected _bet_ recordings, we record both _bet_ and _bat_ 18
times and ask the learner to select 18 suspected _bet_ recordings.
The results are summarised in the following contingency table:

\begin{array}{l|cc}
  & \textrm{Categorised as \textit{bet}}  & \textrm{Categorised as \textit{bat}}  \\
  \hline
  \textrm{Is \textit{bet}} & 5            & 13 \\
  \textrm{Is \textit{bat}} & 13           & 5 \\
\end{array}

Under the null hypothesis that the learner possesses no relevant discriminatory
ability, the top-left entry ($n_{11}$) follows a $\textrm{Hypergeometric}(18, 18, 18)$
distribution; see @fig-hypergeometric-twosided.
The observed top-left entry (5) is highlighted by the dashed blue line.

```{r, fig-hypergeometric-twosided, echo = FALSE, fig.cap = "The probability mass function of the $\\textrm{Hypergeometric}(18, 18, 18)$ distribution.", fig.height = 2.8, fig.width = 3.5, out.width = "38%"}
par(las = 1,
    bty = "l",
    mar = c(3,3.5,2,1),
    mgp = c(2,0.3, 0),
    tck = -.01,
    cex = 0.8,
    mfrow = c(1, 1))

p_k <- dhyper(0:18, 18, 18, 18)
colour <- ifelse(p_k <= p_k[6], "blue", "black")
plot(0:18, p_k, pch = 16, col = colour,
     xlab = bquote(k), ylab = expression(Prob(n[11] == k)))
abline(v = 5, lty = "dashed", col = "blue", lwd = 2)
```

Of note, the learner seems to be able to tell _bet_ and _bat_ apart
to some extent -- it's just that he seems to identify _bet_ recordings
as _bat_ and vice versa. 
Since we're interested in the learner's discriminatory ability,
regardless of whether he is then also able to correctly _label_
the two categories, we want to compute a two-sided $p$-value.
The most common way to do so is to sum the probability masses
$\mathbb{P}(n_{11} = k), k = 1, \dots, n_{+1},$ that are no greater
than the probability mass of the actually observed top-left entry.
These are the probability masses coloured blue in @fig-hypergeometric-twosided.

```{r}
p_k <- dhyper(0:18, 18, 18, 18)
sum(p_k[p_k <= dhyper(5, 18, 18, 18)])
```

Fisher's exact test carries out the same computation.
```{r}
tab <- rbind(c(5, 13), c(13, 5))
fisher.test(tab)$p.value
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# chisq test only approximate
par(las = 1,
    bty = "l",
    mar = c(3,3.5,2,1),
    mgp = c(2,0.3, 0),
    tck = -.01,
    cex = 0.8,
    mfrow = c(1, 1))
library(tidyverse)
.p_chisq <- function(k, n1plus = 18, n2plus = 18, nplus1 = 18, correct = FALSE) {
  tab <- rbind(c(k, n1plus - k),
               c(nplus1 - k, n2plus - nplus1 + k))
  chisq.test(tab, correct = correct)$p.value
}
p_chisq <- Vectorize(.p_chisq, vectorize.args = "k")
n1plus <- 5
n2plus <- 8
nplus1 <- 5
k <- 0:n1plus
pvals <- p_chisq(k, n1plus, n2plus, nplus1, correct = FALSE)
probs <- dhyper(k, n1plus, n2plus, nplus1)
df <- data.frame(pvals, probs)
df <- df |> 
  group_by(pvals) |> 
  summarise(probs = sum(probs)) |> 
  arrange(pvals) |> 
  mutate(cumprob = cumsum(probs))
plot(df$pvals, df$cumprob, pch = 16,
     xlab = expression(alpha),
     ylab = expression(Prob(P[chi^2] <= alpha)))
points(c(df$pvals[-1], 1.5), df$cumprob, pch = 1)
segments(x0 = c(df$pvals),
         x1 = c(df$pvals[-1], 1.5),
         y0 = c(df$cumprob),
         y1 = c(df$cumprob))
abline(a = 0, b = 1, lty = 2)
```

## Contingency tables with one marginal fixed
Contingency tables in which both the row and column marginals are fixed in advance
are a rare sight.
More common in some areas of research are contingency tables where only the row 
marginals are fixed by design.
Such tables can be found in, for instance, experimental research in which a
fixed number of participants are assigned to one experimental condition
and another fixed number of participants are assigned to the other experimental
condition and where for each participant, we have a single binary outcome
(e.g., died vs. survived, or passed vs. failed).

**Example 3** (Boschloo's test)**.**
Let's imagine we're in charge of an agency that designs self-study 
courses to help students prepare for an entrance exam.
We've developed a course whose efficacy we want to compare against
that of its predecessor.
More specifically, we're interested in finding out whether the new course
is better than the old one in terms of helping the students pass the entrance
exam.
We recruit 24 students willing to participate in an evaluation study
and, using complete randomisation,
we assign 12 of them to work with the new course and 12 to work with the old one.
The results look as follows:


\begin{array}{l|cc}
  & \textrm{Pass}  & \textrm{Fail}  \\
  \hline
  \textrm{New course} & 10          & 2 \\
  \textrm{Old course} & 6           & 6 \\
\end{array}

In contrast to the previous two examples, 
it's only the row marginals that were known beforehand in this example.
Nevertheless,
applying Fisher's exact test to this contingency table is 
reasonable.
In doing so, 
we would be **conditioning** the analysis on the observed column marginals
($n_{+1} = 16$, $n_{+2} = 8$), 
even though the study design did not fix these marginals.
The corresponding null hypothesis is

$$H_0 | n_{+1} = 16, n_{+2} = 8~:~ n_{11} \sim \textrm{Hypergeometric}(12, 12, 16).$$

An exact test remains exact when it is used conditionally (Lydersen 2009:1165), 
so the resulting $p$-value is perfectly valid:

```{r}
tab <- rbind(c(10, 2), c(6, 6))
fisher.test(tab, alternative = "greater")$p.value
```

That said, conditional exact tests tend to be pretty conservative.
Intuitively, the reason is that the conditional exact test only considers
$\min(n_{+1}, n_{1+}) + 1$ of the possible tables that could have been observed under the null
hypothesis; in this example, these would be the thirteen different tables 
where $n_{1+} = n_{2+} = 12$ and $n_{+1} = 16$, i.e., with $n_{11} = 0, 1, \dots, 12$.
The actual sample space under the null hypothesis in this design, however,
comprises $(n_{1+} + 1) \cdot (n_{2+} + 1)$ tables.
In this example, these would be the 169 different tables where $n_{1+} = n_{2+} = 12$.
By only considering a small part of the sample space, the conditional exact test
in essence misses out on opportunities to return small $p$-values.

Unconditional exact tests consider the whole sample space and are consequently
less conservative than conditional exact tests.
The contingency table in the fixed row marginals design can be considered the result
of two draws from binomial distributions: one with $n_{1+}$ attempts and
success probability $\pi_1$, and one with $n_{2+}$ attempts and success
probability $\pi_2$. The null hypothesis is that $\pi_1 = \pi_2$.
That is,
\begin{align*}
H_0~:~ & n_{11} \sim \textrm{Binomial}(n_{1+}, \pi), \\ 
       & n_{21} \sim \textrm{Binomial}(n_{2+}, \pi),
\end{align*} 
with some $\pi \in [0,1]$ that is common to both binomial distributions.
This $\pi$ parameter is known as a **nuisance parameter**: its value is
unknown and not of primary interest, 
but if we want to carry out an unconditional test, 
we need to somehow take it into account.

An unconditional exact test for the fixed row marginals design
that is often recommended is **Boschloo's test** (Boschloo 1970).
The idea behind this test is as follows.
First, we define a test statistic that captures the extent to which
observed contingency tables differ from the contingency table you'd expect
to find under the null hypothesis, given $n_{1+}$, $n_{2+}$ and some candidate
value for $\pi$.
Second, for each table that we could have observed, we compute both how likely 
it was to occur (this depends on $n_{1+}$, $n_{2+}$ and $\pi$) and the 
test statistic it would have resulted in.
Third, using the results of these calculations, we compute the probability
that we would observe a test statistic at least as extreme as the test statistic
that we actually did observe under the null hypothesis and for 
the given $n_{1+}, n_{2+}$ and $\pi$ values. This probability is a $p$-value
conditional on the $\pi$ value considered.
We repeat this procedure for different $\pi$ values -- say, for 101 equally
spaced $\pi$ values in the interval $[0, 1]$.
Our unconditional exact $p$-value is now the maximum of the 101
values so computed.
(Theoretically, it should be the supremum of the $p$-values when
varying $\pi$ over the entire $[0,1]$ interval. But the maximum of the 101
$p$-values compute will be close enough to this supremum.)
In Boschloo's test, the test statistic used is in fact the one-sided
$p$-value obtained from Fisher's exact test.
But this $p$-value is used as a test statistic, not as a $p$-value in its own
right.

Let's walk through the computation step by step.
First, we run a one-sided Fisher's exact test in order to obtain the observed
test statistic:

```{r}
(obs_test_stat <- fisher.test(rbind(c(10, 2), c(6, 6)), alternative = "greater")$p.value)
```

Next, we create a grid with all possible combinations of $(n_{11}, n_{21})$ entries
that we could have observed:
```{r}
n_row1 <- 12
n_row2 <- 12
tables <- expand.grid(n11 = 0:n_row1, n21 = 0:n_row2)
```

We now fix some $\pi \in [0, 1]$, for instance, $\pi = 0.43$.
For each possible table, we compute how likely it would have been to observe
this table if $\pi = 0.43$. For instance, the 61st row in the grid
corresponds to the table

\begin{array}{l|cc}
  & \textrm{Pass}  & \textrm{Fail}  \\
  \hline
  \textrm{New course} & 8         & 4 \\
  \textrm{Old course} & 6         & 6 \\
\end{array}

The probability of observing the first row is given by the probability mass
of $k = 8$ under a Binomial(12, 0.43) distribution;
the probability of observing the second row is given by the probability mass
of $k = 6$, also under a Binomial(12, 0.43) distribution.
So the probability of observing this table is the product of these two probabilities:
```{r}
dbinom(8, 12, 0.43) * dbinom(6, 12, 0.43)
```

We compute this probability for all 169 tables:
```{r}
tables$probability <- dbinom(tables$n11, 12, 0.43) * dbinom(tables$n21, 12, 0.43)
```

For each table, we also compute the test statistic:

```{r}
tables$test_statistic <- NA
for (i in 1:nrow(tables)) {
  current_table <- rbind(c(tables$n11[i], 12 - tables$n11[i]),
                         c(tables$n21[i], 12 - tables$n21[i]))
  tables$test_statistic[i] <- fisher.test(current_table, alternative = "greater")$p.value
}
```

We can now compute the probability that we'd observe a test statistic at least
as extreme as the test statistic associated with the table we actually observed,
assuming the null hypothesis is true and $\pi = 0.43$.
Since the $p$-value produced by Fisher's exact test is smaller for tables
that are more surprising under the null hypothesis, 'at least as extreme'
corresponds to 'at most as large'. 
We can compute the figure we're interested in by computing the proportion
of tables resulting in test statistics no larger than the one we observed
but weighted by their probability of occurring under the assumption that $\pi = 0.43$:

```{r}
weighted.mean(tables$test_statistic <= obs_test_stat, w = tables$probability)
```

Assuming $\pi = 0.43$, the resulting $p$-value is hence $0.043$.
We repeat this procedure for different candidate values of $\pi$.
The $p$-value of Boschloo's test is then the maximum of the resulting $p$-values.
In this specific example, the maximum $p$-value is $p = 0.0495$, which is obtained
for $\pi = 0.68$; see @fig-boschloo.

```{r, fig-boschloo, echo = FALSE, fig.cap = "In Boschloo's test, a $p$-value is computed for each candidate value of $\\pi$. The test then outputs the maximum of these $p$-values.", fig.height = 1.5*2.8, fig.width = 1.8*3.5, out.width = "66%"}
#| classes: preview-image
par(las = 1,
    bty = "l",
    mar = c(3,3.5,2,1),
    mgp = c(2,0.3, 0),
    tck = -.01,
    cex = 0.8,
    mfrow = c(1, 1))
tables <- expand.grid(n11 = 0:12, n21 = 0:12)
probability <- vector(length = nrow(tables))
test_statistic <- vector(length = nrow(tables))
pvals <- vector(length = 101)

for (i in 1:nrow(tables)) {
  current_table <- rbind(c(tables$n11[i], 12 - tables$n11[i]),
                         c(tables$n21[i], 12 - tables$n21[i]))
  test_statistic[i] <- fisher.test(current_table, alternative = "greater")$p.value
}
is_extreme <- test_statistic <= obs_test_stat

for (j in 1:length(pvals)) {
  current_pi <- (j - 1) / 100
  for (i in 1:nrow(tables)) {
  current_table <- rbind(c(tables$n11[i], 12 - tables$n11[i]),
                         c(tables$n21[i], 12 - tables$n21[i]))
  probability[i] <- dbinom(tables$n11[i], 12, current_pi) * dbinom(tables$n21[i], 12, current_pi)
  }
  pvals[j] <- weighted.mean(is_extreme, w = probability)
}
plot(((1:101) - 1) / 100, pvals,
     col = ifelse(pvals == max(pvals), "red", "grey"),
     pch = ifelse(pvals == max(pvals), 16, 1),
     xlab = expression(pi),
     ylab = "p-value")
abline(h = max(pvals), lty = 2, col = "red", lwd = 2)
abline(v = (which.max(pvals) - 1) / 100, lty = 2, col = "red", lwd = 2)
```

If our alternative hypothesis were that the new programme produced worse
results than the old one, we'd have used the left-sided $p$-value of Fisher's
exact test as the test statistic in Boschloo's procedure.
A two-sided $p$-value can be obtained by carrying out Boschloo's test once
using the right-sided $p$-value of Fisher's exact test as the test statistic
and once using the left-sided $p$-value, and then doubling the smaller
of the two resulting $p$-values.

To run Boschloo's test, you can use the following `boschloo_test()` function:

```{r}
boschloo_test <- function(tab, alternative = "two.sided", pi_range = c(0, 1), stepsize = 0.01) {
  # This test assumes fixed row sums.
  # Nuisance parameter values in the interval pi_range are tried out.
  # stepsize governs granularity of search through nuisance parameter value candidates.
  
  if (!all(dim(tab) == c(2, 2))) stop("tab needs to be a 2*2 contingency table.")
  
  if (alternative == "two.sided") {
    # Truncate two-sided p-value at 1
    return(
      min(2 * min(boschloo_test(tab, alternative = "less", pi_range = pi_range, stepsize = stepsize),
                  boschloo_test(tab, alternative = "greater", pi_range = pi_range, stepsize = stepsize)), 
          1)
      )
  }
  
  # Use Fisher's exact test p-value as test statistic
  statistic <- function(x) fisher.test(x, alternative = alternative)$p.value
  
  # Construct grid with possible results
  row_sums <- rowSums(tab)
  my_grid <- expand.grid(n1 = 0:row_sums[1], n2 = 0:row_sums[2])
  my_grid$statistic <- NA
  for (i in 1:nrow(my_grid)) {
    my_tab <- rbind(c(my_grid$n1[i], row_sums[1] - my_grid$n1[i]),
                    c(my_grid$n2[i], row_sums[2] - my_grid$n2[i]))
    my_grid$statistic[i] <- statistic(my_tab)
  }
  
  # Compute observed test statistic
  obs_p <- statistic(tab)
  is_extreme <- my_grid$statistic <= obs_p
  
  # Maximise p-value over range
  pis <- seq(pi_range[1], pi_range[2], by = stepsize)
  max_p <- 0
  for (current_pi in pis) {
    current_p <- weighted.mean(x = is_extreme,
      w = dbinom(my_grid$n1, row_sums[1], current_pi) * 
            dbinom(my_grid$n2, row_sums[2], current_pi))
    if (current_p > max_p) max_p <- current_p
  }
  max_p
}
```

It works like so:

```{r}
boschloo_test(tab = tab, alternative = "greater")
```

Alternatively, you can use the `boschloo()` function in the [`exact2x2` package](https://cran.r-project.org/web/packages/exact2x2/index.html).
See `?boschloo` for details on the parameters.
Here I specify the number of grid points (`nPgrid`) in order to make the results
agree exactly with those produced by `boschloo_test()`:
```{r}
exact2x2::boschloo(6, 12, 10, 12, alternative = "greater",
                   control = exact2x2::ucControl(nPgrid = 101))$p.value
```

## Contingency tables with only the total sum fixed
A third possibility is that neither the row sums ($n_{1+}, n_{2+}$)
nor the column sums ($n_{+1}, n_{+2}$) are known beforehand but that the
total number of observations ($n_{++}$) is.
One general case where such contingency tables arise is in observational 
research in which two binary features are measured in a fixed number of 
randomly sampled units.
A second general case is comprised of experiments that have
a binary outcome variable and in which the participants are assigned to the
conditions using simple randomisation so that the precise number of participants
isn't fixed in advance.

### Observational studies
**Example 4** (Boschloo's test)**.**
During a hike through the Fribourg Prealpes near [Schwarzsee](https://www.myswitzerland.com/en-ch/destinations/schwarzsee/), 
we conduct a linguistic field experiment.
Any time we encounter a hiking party chatting in French, we greet them in German;
any time we encounter a hiking party chatting in German, we greet them in French.
Afterwards, we jot down for each party whether the first person greeting us back
did so in the same language in which they were addressed or in a different language.
(Parties not chatting in French or German are ignored in this field experiment.)
We planned to continue the field experiment until we've encountered the 20th French-
or German-speaking hiking party.
Here's the resulting contingency table:

\begin{array}{l|cc}
                                 & \textrm{Same language}  & \textrm{Different language}  \\
  \hline
  \textrm{French-speaking party} & 2                       & 6 \\
  \textrm{German-speaking party} & 8                       & 4 \\
\end{array}

Note that only the total number of observations
($n_{++} = 20$) was known beforehand.

The null hypothesis is that whether the first greeter in the party responded
in the same language or in a different language is independent of the 
language in which the party was chatting.
More formally, let $\pi_{\textrm{row}}$ be the probability that a randomly encountered
French- or German-speaking party is French-speaking
and let $\pi_{\textrm{col}}$ be the probability that the first greeter in a 
French- or German-speaking party responds in the same language.
Then our null hypothesis is

\begin{align*}
H_0~:~&\mathbb{P}(\textrm{is French-speaking and responds in same language}) = \pi_{\textrm{row}}\pi_{\textrm{col}}, \\
      &\mathbb{P}(\textrm{is French-speaking and responds in different language}) = \pi_{\textrm{row}}(1-\pi_{\textrm{col}}), \\
      &\mathbb{P}(\textrm{is German-speaking and responds in same language}) = (1-\pi_{\textrm{row}})\pi_{\textrm{col}}, \\
      &\mathbb{P}(\textrm{is German-speaking and responds in different language}) = (1-\pi_{\textrm{row}})(1-\pi_{\textrm{col}}).
\end{align*}

More compactly, our null hypothesis is that the tuple $(n_{11}, n_{12}, n_{21}, n_{22})$
is multinomially distributed with the probabilities above, i.e.,

$$H_0~:~(n_{11}, n_{12}, n_{21}, n_{22}) \sim \textrm{Multinomial}(20, \pi_{\textrm{row}}\pi_{\textrm{col}},\pi_{\textrm{row}}(1-\pi_{\textrm{col}}),(1-\pi_{\textrm{row}})\pi_{\textrm{col}}, (1-\pi_{\textrm{row}})(1-\pi_{\textrm{col}})),$$
for some unknown $\pi_{\textrm{row}}, \pi_{\textrm{col}} \in [0,1]$.

In terms of the analysis, we could condition on both marginals
or on one of them and run Fisher's or Boschloo's test, respectively:

```{r}
tab <- rbind(c(2, 6), c(8, 4))

# Condition on both marginals
fisher.test(tab)$p.value

# Condition on row marginals only
boschloo_test(tab)

# Condition on column marginals only
boschloo_test(t(tab))
```

The resulting $p$-values are all valid.
But like before, these tests only consider a part of the sample space:
Fisher's exact test only takes into account the nine tables
where $n_{1+} = 8$ and $n_{+1} = 10$;
Boschloo's test considers the $9 \cdot 13 = 117$ tables
where $n_{1+} = 8$ and $n_{2+} = 12$ when conditioning on the row marginals
and the $11 \cdot 11 = 121$ tables where $n_{+1} = n_{+2} = 10$
when conditioning on the column marginals.
But the table observed is in actual fact one of 
$$\sum_{i = 0}^{20}\sum_{j = 0}^{20-i}\sum_{k= 0}^{20-i-j}1 = 1771$$
tables where $n_{++} = 20$.

One possible solution is to generalise Boschloo's test to two nuisance parameters.
That is, rather computing a $p$-value for each candidate value of $\pi \in [0,1]$
and then taking the maximum, we compute a $p$-value for each pair
of candidate values of $(\pi_{\textrm{row}}, \pi_{\textrm{col}}) \in [0,1] \times [0,1]$.
The `unconditional_test()` function defined below carries out this procedure:
```{r}
unconditional_test <- function(
    tab, 
    alternative = "two.sided", 
    pi_range_row = c(0, 1), 
    pi_range_col = c(0, 1),
    stepsize = 0.01)
  {
  
  # This test assumes a fixed total sum.
  # Nuisance parameter values in the rectangle pi_range_row * pi_range_col are tried out.
  # stepsize governs granularity of search through nuisance parameter value candidates.
  
  if (!all(dim(tab) == c(2, 2))) stop("tab needs to be a 2*2 contingency table.")
  
  if (alternative == "two.sided") {
    # Truncate two-sided p-value at 1
    return(min(
      2 * min(unconditional_test(tab, alternative = "less", stepsize = stepsize),
              unconditional_test(tab, alternative = "greater", stepsize = stepsize)),
      1))
  }
  
  # Use Fisher's exact test p-value as test statistic
  statistic <- function(x) fisher.test(x, alternative = alternative)$p.value
  
  # Helper function for multinomial weights
  weights <- function(pi_row, pi_col, n11, n12, n21, n22) {
    total_sum <- n11 + n12 + n21 + n22
    (pi_row*pi_col)^n11 * (pi_row * (1 - pi_col))^n12 *
      ((1 - pi_row)*pi_col)^n21 * ((1 - pi_row)*(1 - pi_col))^n22 * 
      factorial(total_sum)/(factorial(n11) * factorial(n12) * factorial(n21) * factorial(n22))
  }
  
  # Construct grid with possible results
  total_sum <- sum(tab)
  my_grid <- expand.grid(n11 = 0:total_sum, 
                         n12 = 0:total_sum,
                         n21 = 0:total_sum)
  my_grid$n22 <- total_sum - my_grid$n11 - my_grid$n12 - my_grid$n21
  my_grid <- subset(my_grid, n22 >= 0)
  my_grid$statistic <- NA
  for (i in 1:nrow(my_grid)) {
    my_tab <- rbind(c(my_grid$n11[i], my_grid$n12[i]),
                    c(my_grid$n21[i], my_grid$n22[i]))
    my_grid$statistic[i] <- statistic(my_tab)
  }
  my_grid$statistic[is.na(my_grid$statistic)] <- 1

  # Compute observed test statistic
  obs_p <- statistic(tab)
  is_lower <- my_grid$statistic <= obs_p
  
  # Maximise p value over grid
  pis <- expand.grid(
    pi_row = seq(pi_range_row[1], pi_range_row[2], by = stepsize),
    pi_col = seq(pi_range_col[1], pi_range_col[2], by = stepsize)
  )
  max_p <- 0
  for (i in 1:nrow(pis)) {
    w <- weights(pis$pi_row[i], pis$pi_col[i], 
                 my_grid$n11, my_grid$n12, my_grid$n21, my_grid$n22)
    current_p <- weighted.mean(is_lower, w = w)
    if (current_p > max_p) max_p <- current_p
  }
  
  max_p
}
```

While it takes noticeably longer to run this test, it should be a bit more powerful
than the tests that condition on one or both marginals:
```{r}
unconditional_test(tab)
```

The `exact.test()` function in the [`Exact`](https://cran.r-project.org/package=Exact)
also implements this procedure.
For one-sided tests (i.e., `alternative = "less"` or `"greater"`),
it produces the same $p$-values as `unconditional_test()`,
but it computes the two-sided $p$-value differently:


```{r}
Exact::exact.test(tab, model = "multinomial",
                  npNumber = 101, ref.pvalue = FALSE, method = "boschloo")$p.value
```


### Experiments with simple randomisation
**Example 5** (Boschloo's test)**.**
We conduct the same experiment as in Example 3, but with one change:
We assign the participants to the conditions using simple randomisation
rather than using complete randomisation.
Hence, we're not guaranteed to have exactly 12 participants in each condition,
and the row marginals aren't fixed in advance.
For the sake of comparison, let's assume we obtain the same results as in Example 3:

\begin{array}{l|cc}
  & \textrm{Pass}  & \textrm{Fail}  \\
  \hline
  \textrm{New course} & 10        & 2 \\
  \textrm{Old course} & 6         & 6 \\
\end{array}

We could again condition on the row marginals:
```{r}
tab <- rbind(c(10, 2),
             c(6, 6))
boschloo_test(tab, alternative = "greater")      # condition on row marginals
```

Alternatively, we could run an unconditional test.
Note, however that while $n_{1+}$ and $n_{2+}$ weren't fixed by design,
$\pi_{\textrm{row}}$ is known to be $0.5$.
Hence, we don't need to iterate through different candidate values for $\pi_{\textrm{row}}$:

```{r}
unconditional_test(tab, alternative = "greater", pi_range_row = c(0.5, 0.5))
```

## Contingency tables with nothing fixed
Contingency tables where not even $n_{++}$ is fixed are mentioned
by Fagerland et al. (2017) but not discussed in any detail.
An analysis that conditions on the observed $n_{++}$ seems reasonable.

## References
Boschloo, R. D. 1970.
[Raised conditional level of significance for the 2 × 2 table when testing the equality of two probabilities.](https://doi.org/10.1111%2Fj.1467-9574.1970.tb00104.x)
*Statistica Neerlandica* 24.
1-35.

Fagerland, Morten W., Stian Lydersen & Petter Laake. 2017. 
[*Statistical analysis of contingency tables*.](https://doi.org/10.1201/9781315374116) 
Boca Raton, FL: Chapman and Hall/CRC.

Lydersen, Stian, Morten W. Fagerland & Petter Laake. 2009.
[Recommended tests for association in 2 × 2 tables.](https://doi.org.10.1002/sim.3531)
_Statistics in Medicine_ 28. 1159-1175.