---
title: "A quick introduction to Gaussian process regression"
date: "2026-07-09"
categories: [Gaussian processes, predictive modelling]
format:
  html:
    code-fold: true
    message: false
    df-print: paged
    fig-width: 8
    fig-height: 6
    fig-align: center
knitr:
  opts_chunk:
    out.width: "100%"
filters:
  - shinylive
embed-resources: false
---

For my Master's thesis in statistics,
I took a closer look at a method that allows you to do predictive regression 
modelling when the inputs are represented as probability distributions.



(1) Predictive model, here: for solving regression problems.

(2) Goal: explain basic logic.

The goal of this blog post is to introduce the nuts and bolts of 
Gaussian process regression.

Assume that for any finite collection of outputs $Y_1, \dots, Y_n$,
the vector $\boldsymbol{Y} = (Y_1, \dots, Y_n)^{\top}$
has an $n$-variate Gaussian (i.e., Normal) distribution:
$$\boldsymbol Y \sim \mathcal{N}_n(\boldsymbol \mu, \boldsymbol \Sigma),$$
where $\boldsymbol \mu$ is the distribution's $n$-valued mean vector
and $\boldsymbol \Sigma$ is its $n \times n$ covariance matrix.

For simplicity, we assume that the mean vector of this distribution is $\boldsymbol{0}$.
The $(i,j)$-th element of $\boldsymbol \Sigma$ expresses the covariance of 
the $i$-th and $j$-th element of $\boldsymbol Y$.



```{r}
#| label: fig-kernels
#| fig-cap: "A scatterplot to which we want to add a trend line."
#| fig-height: 10
# Plotting functions -----------------------------------------------------------
plot_gpr <- function(x, K, n = 4, main = "") {
  cols <- RColorBrewer::brewer.pal(n, "Dark2")
  Y <- MASS::mvrnorm(n = n, mu = rep(0, length(x)), Sigma = K)
  plot(x = x, y = Y[1, ], ylim = range(Y), type = "n", las = 1,
     xlab = expression(italic(x)), 
     ylab = expression(italic(y)), 
     main = main)
  for (i in seq_len(n)) {
    points(x = x, y = Y[i, ], type = "o", col = cols[i])
  }
}
plot_corr <- function(K, main = "") {
  corrplot::corrplot(cov2cor(K),
    method = "color", title = main, tl.pos = "n")
}

# Plot kernels -----------------------------------------------------------------
op <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))

# Random seed
set.seed(2026-07-07)

# Predictor data and distances between them
x <- seq(-5, 5, by = 0.2)
D <- outer(x, x, "-") |> abs()

# Linear kernel with an offset
K_linear <- outer(x, x, "*") + matrix(1, nrow = length(x), ncol = length(x))
plot_corr(K_linear, main = "Linear kernel")
plot_gpr(x, K_linear, main = "Linear kernel")

# Periodic kernel
K_periodic <- exp(-sin(D)^2)
plot_gpr(x, K_periodic, main = "Periodic kernel")

# RBF kernel
K_rbf <- exp(-D^2)
plot_gpr(x, K_rbf, main = "Gaussian RBF kernel")

# Matérn 1/2 kernel
K_matern12 <- exp(-D)
plot_gpr(x, K_matern12, main = "Matérn(1/2) kernel")

# Matérn 3/2 kernel
K_matern32 <- (1 + sqrt(3) * D) * exp(-sqrt(3)*D)
plot_gpr(x, K_matern32, main = "Matérn(3/2) kernel")

# Matérn 5/2 kernel
K_matern52 <- (1 + sqrt(5) * D + 5 * D^2 / 3) * exp(-sqrt(5)*D)
plot_gpr(x, K_matern52, main = "Matérn(5/2) kernel")

par(op)
```

## References and further reading


## Session info
```{r}
devtools::session_info("attached")
```

