---
title: ""
date: "2026-06-03"
categories: [Gaussian processes, predictive modelling, Wasserstein distances]
format:
  html:
    code-fold: true
    message: false
    df-print: paged
    fig-width: 8
    fig-height: 6
    fig-align: center
knitr:
  opts_chunk:
    out.width: "100%"
---

Lexical frequency as predictor of text quality ratings.
Typically summarised as scalar (e.g., average log-frequency)
or binned as lexical frequency profile.

But frequency distributions may have same scalar summaries yet differ
considerably in their shape.
Further, binning is based on arbitrary choices (number of bins, cut-offs).

Alternative: use distributions as predictors.

Example: German data.

(1) Frequency distributions (Zipf).
(2) Mean Zipf.
(3) Frequency bands à la Nation & Laufer.

Construct kernels. Use Gaussian process regression. Compare performance.

```{r}
library(slicer)
library(tidyverse)

# Scalar inputs and output
d <- read_csv("german_summary.csv") |> 
  select(TextType, Time, Batch, 
         contains("LemmaFreq"), LemmameanZipf, LemmameanZipfUnique,
         meanRating)

# Each text as a distribution
load("word_distributions_matrix.Rda")
length(matrix_list)
colnames(matrix_list[[314]])
matrix_list[[314]]
```

```{r}
test_idx <- which(d$Batch %in% c("F", "G", "I", "N"))
training_idx <- (1:nrow(d))[-test_idx]
```

### Mean Zipf values (unique and not unique)
```{r}
mean_zipf_dist <- outer(d$LemmameanZipf, d$LemmameanZipf, "-") |> abs()
mean_unique_zipf_dist <- outer(d$LemmameanZipfUnique , d$LemmameanZipfUnique , "-") |> abs()
```

```{r}
# training data
plot(meanRating ~ LemmameanZipf, data = d[training_idx, ],
     xlab = "mean Zipf (lemma-based)", ylab = "Mean rating")
plot(meanRating ~ LemmameanZipfUnique, data = d[training_idx, ],
     xlab = "mean Zipf (unique lemma-based)", ylab = "Mean rating")
```


```{r}
mean_zipf_gpr <- fit_gpr(D2 = mean_zipf_dist^2, training_idx, test_idx,
  y_train = d$meanRating[training_idx], y_test = d$meanRating[test_idx],
  runs = 10L)
plot(d$LemmameanZipf[test_idx], mean_zipf_gpr$test_predictions)
mean_unique_zipf_gpr <- fit_gpr(D2 = mean_unique_zipf_dist^2, training_idx, test_idx,
  y_train = d$meanRating[training_idx], y_test = d$meanRating[test_idx],
  runs = 10L)
plot(d$LemmameanZipfUnique[test_idx], mean_unique_zipf_gpr$test_predictions)
# RMSE: 0.99
```

### Comparison of frequency bands
```{r}
d$LemmaFreqTop1000 <- d[, 4:13] |> rowSums()
top1000_dist <- outer(d$LemmaFreqTop1000, d$LemmaFreqTop1000, "-") |> abs()
top2000_dist <- outer(d$LemmaFreq2000, d$LemmaFreq2000, "-") |> abs()
lfp_dist <- list(top1000_dist^2, top2000_dist^2)
plot(meanRating ~ LemmaFreqTop1000, data = d[training_idx, ],
     xlab = "Proportion in Top 1000", ylab = "Mean rating")
plot(meanRating ~ LemmaFreq2000, data = d[training_idx, ],
     xlab = "Proportion in Band 1001-2000", ylab = "Mean rating")
```

```{r}
lfp_gpr <- fit_gpr(D2 = lfp_dist, training_idx, test_idx,
  y_train = d$meanRating[training_idx], y_test = d$meanRating[test_idx],
  runs = 10L)
# $RMSE
# [1] 1.049372
# 
# $length_scale
# [1] 8.965691e-02 9.390859e-16
# 
# $scaling_factor
# [1] 0.2483085 0.3552805
# 
# $noise_variance
# [1] 1.002978
# 
# $nll
# [1] 1226.377
```


### Entire distributions
```{r}
e_vect <- function(i, n) {
  if (i == 0) {
    return(c(1, rep(0, n - 1)))
  }
  if (i == n) {
    return(c(rep(0, n - 1), 1))
  }
  c(rep(0, i - 1), 1, rep(0, n - i))
}

thetas <- matrix(
  c(
    e_vect(13, 21) # frequency information
  ), 
  nrow = 1, byrow = TRUE)
zipf_distances <- compute_all_distances(
  matrix_list, thetas, verbose = TRUE, keep_projections = TRUE
)

distribution_gpr <- fit_gpr(D2 = zipf_distances, training_idx, test_idx,
  y_train = d$meanRating[training_idx], y_test = d$meanRating[test_idx],
  runs = 10L)
# $RMSE
# [1] 0.9280229
# 
# $length_scale
# [1] 0.3431806
# 
# $scaling_factor
# [1] 1.413498
# 
# $noise_variance
# [1] 0.7332158
# 
# $nll
# [1] 1154.967
```


## References

## Session info
```{r}
devtools::session_info("attached")
```

