---
title: "Testing strong and weak null hypotheses in randomised experiments"
date: "2026-07-13"
categories: [assumptions, experiments, design features, nonparametric tests, R, significance]
format:
  html:
    code-fold: true
    message: false
    fig-width: 9
    fig-height: 3.3
---

In a [previous post](../2025-11-09-population-randomisation/),
I explained how randomisation tests work and 
why they are a sensible choice when analysing the data from a randomised
experiments with a null hypothesis test.
In this post, I want to take a closer look at the distinction between 
so-called **strong** or **sharp** (Fisherian) and **weak** (Neymanian) null hypotheses
and how they can be tested.
This post is in a way a synopsis of chapters 3, 4, and 8
of Ding's (2024) _First course in causal inference_.

## Recap: Randomisation tests of the strong null hypothesis
Under the randomisation model, we posit that there are two 
potential outcomes $y_i(0)$ and $y_i(1)$ for each unit $i = 1, \dots, m$
in a randomised experiment with two conditions, only one of which is observed:

* either $y_i(0)$ if the $i$-th unit is assigned to the first condition;
* or $y_i(1)$ if the unit is assigned to the second condition.

The quantities $y_i(0), y_i(1), i = 1, \dots, m$ are unknown but fixed quantities
in the present set-up -- they are _not_ random variables.
What are random variables are the data that are actually observed.
The randomness is introduced by the random assignment of units to conditions.

According to Fisher's **strong null hypothesis**, 
there is no differential treatment effect for _any_ of the units:

$$H_{oS}: y_i(0) = y_i(1) \text{ for } i = 1, \dots, m.$$

As I've explained [before](../2025-11-09-population-randomisation/),
this strong null hypothesis can be tested exactly 
by means of a randomisation test using _any_ definable test statistic.
This test statistic could be the observed mean difference between the two
conditions, the observed difference between the condition medians,
the probability of superiority -- anything the analyst deems sensible.

A test statistic that I did not use in the other blog post
but that is important now
is the **studentised mean difference**:

$$
  t := \frac{\widehat{y}(1) - \widehat{y}(0)}{\sqrt{\widehat{V}}}.
$$

In this equation, $\widehat{y}(0), \widehat{y}(1)$ are the condition means in the first
and second condition, respectively, 
and $\widehat{V}$ is defined as

$$
  \widehat{V} := \frac{\widehat{S}^2(0)}{n_0} + \frac{\widehat{S}^2(1)}{n_1},
$$

where $\widehat{S}^2(0)$ and $n_0$ are the sample variance and the group size
in the first condition
and $\widehat{S}^2(1)$ and $n_1$ are the sample variance and the group size
in the second condition.
Incidentally, 
this definition of $t$ agrees with the one used in Welch's _t_-test,
i.e., the $t$-statistic you get when running R's `t.test()`
with the default parameter setting `var.equal = FALSE`.

The following code snippet defines a function for computing 
the studentised mean difference. 

```{r}
studentised_mean_diff <- function(x, group) {
  v_hat <- var(x[group == 1]) / length(x[group == 1]) + 
    var(x[group == 0]) / length(x[group == 0])
  (mean(x[group == 1]) - mean(x[group == 0])) / sqrt(v_hat)
}
```

I've also copied the function for running a randomisation test 
from the other blog post.

```{r}
randomisation_test <- function(
    # Randomisation test of strong null in two-group comparison.
    x,           # outcome
    group,       # grouping variable (0 or 1)
    fun,         # function for computing test statistic from x and group
    delta = 0,   # delta for strong null hypothesis
    alternative = "two_sided", # two-sided, left-sided or right-sided p-value?
    M = 19999,   # number of new randomisations generated
    plot = TRUE, # plot histogram of test statistic under H0?
    ...          # parameters passed to histogram
) {
  test_stat <- fun(x, group)
  x[group == 1] <- x[group == 1] - delta
  test_stat_H0 <- replicate(M, {
    fun(x, sample(group))
  })
  test_stat_H0 <- c(test_stat, test_stat_H0)
  
  if (plot) {
    hist(test_stat_H0, ...)
    abline(v = test_stat, lwd = 2)
  }
  
  p_left <- mean(test_stat_H0 <= test_stat)
  if (substr(alternative, 1, 1) == "l") return(p_left)
  
  p_right <-mean(test_stat_H0 >= test_stat)
  if (substr(alternative, 1, 1) == "r") return(p_right)
  
  min(2 * min(p_left, p_right), 1)
}
```

These functions can be used like so:
```{r}
#| code-fold: show
# Define some data
d <- structure(list(
  Group = c(0, 0, 0, 0, 1, 0, 1, 0, 1, 1, 1, 0,
            1, 1, 1, 1, 0, 1, 0, 1, 0, 0, 1, 1, 1, 1, 0), 
  Outcome = c(7, 8, 9.5, 8.5, 7.5, 9, 8, 7.5, 10, 8, 8.5, 
              8, 9, 10.5, 8, 9, 7.5, 9, 9.5, 11, 8, 8, 7.5, 9, 9.5, 8.5, 8.5)), 
  class = "data.frame", row.names = seq_len(27))

# Randomisation test with studentised mean difference
randomisation_test(d$Outcome, d$Group, studentised_mean_diff, plot = FALSE)
```

Randomisation testing pretty easily achieves exact finite-sample inference
for the strong null hypothesis and can straightforwardly be justified in
randomised experiments.
However, two drawbacks in particular bear mentioning.

The first concerns the construction of sensible **confidence sets**.
It _is_ possible to construct confidence sets for, for instance,
a constant treatment shift $\delta$.
This involves testing the strong null hypothesis that the differential
treatment effect is $\delta$ for all units:

$$
H_{oS}(\delta): y_i(1) = y_i(0) + \delta \text{ for } i = 1, \dots, m.
$$
A $(1-\alpha)$ confidence set for the constant treatment shift is then
given by the set of all $\delta$ for which $H_{oS}$ cannot be rejected at the
$\alpha$ level.
Confidence sets for constant multiplicative effects (i.e., $y_i(1) = \gamma y_i(0)$
for $1 \leq i \leq m$) can similarly be constructed.

However, if the data are not compatible with such a constant additive
(or multiplicative) effect, the confidence set may be empty.
Moreover, since it often seems unrealistic 
that any non-zero effect should be the same in all units the $\delta$ 
(or $\gamma$) parameter for which we construct a confidence set 
may have little bearing to reality.

The second drawback of randomisation testing under Fisher's strong null hypothesis
is that it assumes no differential treatment effect for _any_ unit.
But perhaps we're more interested in testing the less restrictive 
null hypothesis that there is no differential treatment effect _on average_.

## The weak null hypothesis
Whereas the strong null hypothesis posits that $y_i(1) = y_i(0)$ for all
units, the weak null hypothesis only posits that this is true _on average_
in the sample.
Let 
$$\overline y(0) := \frac{1}{m}\sum_{i=1}^m y_i(0)$$
and 
$$\overline y(1) := \frac{1}{m}\sum_{i=1}^m y_i(1)$$
denote the average outcomes if every unit in the experiment had been assigned
to the first and to the second condition, respectively.
Neyman's weak null hypothesis posits that the difference 
$\tau := \overline y(1) - \overline y(0)$ between these average is zero:
$$H_{oW}: \tau = 0.$$
Fisher's strong null hypothesis implies Neyman's weak null hypothesis,
but the latter can hold even if the former doesn't,
namely when the individual treatment effects aren't all zero
but happen to cancel each other out.

For each unit, we observe either $y_i(0)$ or $y_i(1)$,
so we don't observe $\overline y(0), \overline y(1)$ or $\tau$, either.
Instead, we estimate $\tau$ as sample mean difference 
between the outcomes observed in the first and second conditions:
$$\widehat \tau := \widehat y(1) - \widehat y(0).$$
It can be shown that in a randomised experiment in which
$y_i(0), y_i(1), i = 1, \dots, m$ are treated as fixed quantities,
the variance of the random quantity $\widehat \tau$ can be estimated conservatively
as
$$
  \widehat{V} = \frac{\widehat{S}^2(0)}{n_0} + \frac{\widehat{S}^2(1)}{n_1},
$$
which is the quantity we used in the denominator of the studentised mean difference
in the previous section (Ding 2024, Theorem 4.1).
What we mean when we say that $\widehat V$ conservatively estimates the variance
of $\widehat \tau$ is that the expected value of $\widehat V$ is at least as large
as the variance of $\widehat \tau$, i.e., $\mathbb{E}(\widehat V) \geq \mathrm{Var}(\widehat \tau)$.

Further, it can be shown that the quantity
$$
  \frac{\widehat \tau - \tau}{\sqrt{\text{Var}(\widehat \tau)}}
$$
has an asymptotic standard normal distribution under some regularity conditions
(Ding 2024, Theorem 4.2).
What this means is that, under some conditions, we can approximate the distribution
of $(\widehat \tau - \tau)/\sqrt{\text{Var}(\widehat \tau)}$ as a standard normal
distribution and that this approximation will be more precise for large $m$.
If we knew what the value of $\textrm{Var}(\widehat \tau)$ was, this would enable
us to construct an approximate $(1-\alpha)$ confidence interval for $\tau$ as
$$\widehat \tau \pm z_{1 - \alpha/2}\sqrt{\textrm{Var}(\widehat \tau)},$$
where $z_{1 - \alpha/2}$ is the $(1-\alpha/2)$ 
quantile of the standard normal distribution.

Usually, we don't know the value of $\textrm{Var}(\widehat \tau)$,
but since it is conservatively estimated by $\widehat V$, we may use the latter
in its stead to obtain the asymptotically conservative $(1 - \alpha)$
**confidence interval**
$$\widehat \tau \pm z_{1 - \alpha/2}\sqrt{\widehat V}$$
for $\tau$.

```{r}
#| code-fold: show
neyman_ci <- function(x, group, conf_level = 0.95) {
  y0_hat <- mean(x[group == 0])
  y1_hat <- mean(x[group == 1])
  v_hat <- var(x[group == 1]) / length(x[group == 1]) + 
    var(x[group == 0]) / length(x[group == 0])
  z <- qnorm(1 - (1 - conf_level)/2)
  y1_hat - y0_hat + c(-1, 1) * z * sqrt(v_hat)
}
```

In our fictitious examples, a 95% confidence interval for $\tau$
can be computed as follows:

```{r}
#| code-fold: show
neyman_ci(d$Outcome, d$Group)
```

A similar confidence interval can be obtained using the Welch approach,
which uses the same test statistic $(\widehat \tau - \tau)/\sqrt{\widehat V}$.
In the Welch approach, this test statistic is compared not against the 
standard normal distribution but against some _t_-distribution.
_t_ distributions are wider than the standard normal distribution,
but for large samples, they are similar.
This explains why the following confidence interval obtained from the Welch
approach is similar but wider than the one obtained from `neyman_test()`:
```{r}
#| code-fold: show
t.test(d$Outcome[d$Group == 1], d$Outcome[d$Group == 0])$conf.int
```
The Welch confidence interval is, however, derived from a different
set of assumptions from 
the confidence interval from `neyman_ci()`, 
namely that the data are random samples from
normally distributed populations.
The Neyman confidence interval does not assume random sampling
but random assignment.

The asymptotic normality of $(\widehat \tau - \tau)/\sqrt{\text{Var}(\widehat \tau)}$
and the conservativeness of $\widehat V$ as an estimator of $\textrm{Var}(\widehat \tau)$
can also be used to compute an asymptotically conservative **_p_-value**
by comparing $(\widehat \tau - \tau)/\sqrt{\widehat V}$ to the quantiles of 
standard normal distribution.
This is implemented in the `neyman_test()` function below.
```{r}
#| code-fold: show
neyman_test <- function(
    # Asymptotic test of weak null in two-group comparison.
    x,     # outcome
    group, # grouping variable (0 or 1)
    alternative = "two_sided" # two-sided, left-sided or right-sided p-value?
) {
  test_statistic <- studentised_mean_diff(x, group)

  p_left <- pnorm(test_statistic, lower.tail = TRUE)
  if (substr(alternative, 1, 1) == "l") return(p_left)

  p_right <- pnorm(test_statistic, lower.tail = FALSE)
  if (substr(alternative, 1, 1) == "r") return(p_right)

  min(2 * min(p_left, p_right), 1)
}
```

It can be used like so:
```{r}
#| code-fold: show
neyman_test(d$Outcome, d$Group)
```

By the same logic as above, the _p_-value obtained from the Welch _t_-test
will be similar but a bit wider than the one obtained from `neyman_test()`:
```{r}
#| code-fold: show
t.test(d$Outcome[d$Group == 1], d$Outcome[d$Group == 0])$p.value
```

Interestingly, the randomisation test 
(as implemented in `randomisation_test()`)
provides an asymptotically conservative test for 
the weak null hypothesis,
provided that the studentised mean difference is used
as the test statistic (Ding and Dasgupta 2017).

## Simulations
Let's run a few simulations to illustrate some of the points made above.
For the simulation, I generate
values for $y_i(0)$ and $y_i(1)$, $i = 1, \dots, m$,
in which the weak (but not the strong) null hypothesis is literally true.
These data are treated as fixed.
For $n_0$ randomly chosen units, we observe the $y_i(0)$ values;
for the remaining $n_1 = m - n_0$ units, we observe the $y_i(1)$ values.
For each full set of $y_i(0)$ and $y_i(1)$, the random assignment is carried out
500 times, 
and each time the data are analysed using four methods:

* A randomisation test using the mean difference $\widehat y(1) - \widehat y(0)$
as the test statistic.
* A randomisation test using the studentised mean difference $\frac{\widehat y(1) - \widehat y(0)}{\sqrt{\widehat V}}$ as the test statistic.
* A $z$-test based on this studentised mean difference; that is, using the `neyman_test()` function.
* Welch's _t_-test.

For each $(n_0, n_1)$ pair, ten full sets of $y_i(0)$ and $y_i(1)$
values are generated.

The code snippet below contains the simulation function.

```{r}
# Packages
suppressPackageStartupMessages(library(tidyverse))
theme_set(theme_bw(12))

# Random seed
set.seed(2026-07-13)

# Simulation function
simulate_weak <- function(
    n0 = 40,
    n1 = 10,
    scenario = "scenario",
    runs = 500,
    M = 999
) {
  # Construct science table under weak null hypothesis
  y0 <- rpois(n0 + n1, 4)
  y1 <- rpois(n0 + n1, 12)
  y0 <- y0 - mean(y0)
  y1 <- y1 - mean(y1)
  groups <- c(rep(0, n0), rep(1, n1))
  
  mean_diff <- function(x, group) {
    mean(x[group == 1]) - mean(x[group == 0])
  }
  
  # Randomise 'runs' times and compute p-values
  p_vals_meandiff <- numeric(runs)
  p_vals_stud_meandiff <- numeric(runs)
  p_vals_neyman <- numeric(runs)
  p_vals_welch <- numeric(runs)
  
  for (run in seq_len(runs)) {
    group0 <- sample(seq_len(n0 + n1), n0)
    group1 <- setdiff(seq_len(n0 + n1), group0)
    outcome <- c(y0[group0], y1[group1])
    p_vals_meandiff[[run]] <- randomisation_test(
      outcome, groups, mean_diff, M = M, plot = FALSE
    )
    p_vals_stud_meandiff[[run]] <- randomisation_test(
      outcome, groups, studentised_mean_diff, M = M, plot = FALSE
    )
    p_vals_neyman[[run]] <- neyman_test(outcome, groups)
    p_vals_welch[[run]] <- t.test(outcome ~ groups)$p.value
  }
  d <- data.frame(
    scenario = rep(scenario, 4 * runs),
    run = rep(seq_len(runs), 4),
    method = rep(c("mean diff.",
                   "studentised m.d.",
                   "Neyman",
                   "Welch"), each = runs),
    pvalue = c(p_vals_meandiff,
               p_vals_stud_meandiff,
               p_vals_neyman,
               p_vals_welch)
  )
  d$method <- factor(d$method)
  levels(d$method) <- c(
    "mean diff.", "studentised m.d.", "Neyman", "Welch"
  )
  d
}
```

First, we consider fairly small groups, with
$n_0 = 20$ and $n_1 = 8$.
@fig-20-8 shows cumulative probability plots of the 500 _p_-values
generated for each method and each set of $y_i(0)$ and $y_i(1)$ values.
The different curves within each facet show the curves for the different
sets of $y_i(0)$ and $y_i(1)$ values.
Importantly, for a significance test to be a valid test of the weak
null hypothesis, these curves should all lie under the line $y = x$,
which is highlighted by a dashed black line.
This is because for a random variable $P$ to be a valid _p_-value,
it should satisfy $\textrm{Prob}(P \leq \alpha) \leq \alpha$
for all $\alpha \in (0, 1)$.

```{r}
#| label: fig-20-8
#| fig-cap: "Distribution of p-values from four methods under the weak null hypothesis for ten fixed sets of potential outcomes ($n_0 = 20, n_1 = 8$)."
map_dfr(1:10, 
  ~ simulate_weak(n0 = 20, n1 = 8, scenario = .x)) |> 
  ggplot(aes(x = pvalue, 
             colour = factor(scenario), 
             group = factor(scenario))) +
  stat_ecdf() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("p-value") +
  ylab("Cumulative probability") +
  scale_colour_brewer(palette = "Paired") +
  facet_grid(
    cols = vars(method)
  ) +
  labs(
    title = "Distribution of p-values under weak null hypothesis",
    subtitle = "Group sizes 20 and 8"
  ) +
  theme(legend.position = "none")
```
As is evident from these plots, the three tests relying on the 
studentised mean difference (i.e., the second randomisation test,
the Neyman test, and the Welch test) all yield valid tests for all
sets of $y$-values.
This is not the case for the randomisation test in which the 
raw mean difference is used as the test statistic: for some sets
of $y$-values, this test is anti-conservative.
(But this randomisation test still is a valid test of the 
_strong_ null hypothesis; just not of the weak one.)

Essentially the same results are obtained for larger group sizes,
as shown @fig-80-32.

```{r}
#| label: fig-80-32
#| fig-cap: "Distribution of p-values from four methods under the weak null hypothesis for ten fixed sets of potential outcomes ($n_0 = 80, n_1 = 32$)."
map_dfr(1:10, 
  ~ simulate_weak(n0 = 80, n1 = 32, scenario = .x)) |> 
  ggplot(aes(x = pvalue, 
             colour = factor(scenario), 
             group = factor(scenario))) +
  stat_ecdf() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("p-value") +
  ylab("Cumulative probability") +
  scale_colour_brewer(palette = "Paired") +
  facet_grid(
    cols = vars(method)
  ) +
  labs(
    title = "Distribution of p-values under weak null hypothesis",
    subtitle = "Group sizes 80 and 32"
  ) +
  theme(legend.position = "none")
```

@fig-10-4 illustrates that none of the approaches provides small-sample
guarantees -- the approaches that worked for the group sizes above are 
approximations based on asymptotic behaviour.
Still, at least for the $y$-values I generated, the Neyman and Welch approaches
seem to perform reasonably well.

```{r}
#| label: fig-10-4
#| fig-cap: "Distribution of p-values from four methods under the weak null hypothesis for ten fixed sets of potential outcomes ($n_0 = 10, n_1 = 4$)."
map_dfr(1:10, 
  ~ simulate_weak(n0 = 10, n1 = 4, scenario = .x)) |> 
  ggplot(aes(x = pvalue, 
             colour = factor(scenario), 
             group = factor(scenario))) +
  stat_ecdf() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("p-value") +
  ylab("Cumulative probability") +
  scale_colour_brewer(palette = "Paired") +
  facet_grid(
    cols = vars(method)
  ) +
  labs(
    title = "Distribution of p-values under weak null hypothesis",
    subtitle = "Group sizes 10 and 4"
  ) +
  theme(legend.position = "none")
```

## tl;dr
For a two-group comparison stemming from a completely randomised experiment
in which no covariates are taken into account,
we may draw the following conclusions.

1. The randomisation test (as implemented in `randomisation_test()`)
is finite-sample exact for the strong (or sharp) Fisherian null hypothesis
of no differential treatment effect for any unit.
'Finite-sample exact' means that it is is a valid test that does not rely
on asymptotic results to provide approximations.
Moreover, this holds for _any_ test statistic used.
2. The randomisation test is also asymptotically conservative for the 
weak Neymanian null hypothesis of no differential treatment effect on average,
provided that the studentised mean difference is used as the test statistic.
For other test statistics, such as the mean difference, 
this property may not hold.
3. Asymptotically conservative tests and confidence intervals for the average
differential treatment effect $\tau = \overline y(1) - \overline y(0)$
may be obtained by comparing the studentised mean difference against a standard normal distribution (`neyman_test()`) or by running a Welch _t_-test.
4. If you're interested in testing both the strong and the weak null hypothesis,
a randomisation test using the studentised mean difference gives you both
an exact test of the strong null hypothesis and an approximate test
of the weak null hypothesis.

## References
Ding, Peng.
2024.
_A first course in causal inference_.
Boca Raton, FL: CRC Press.
Also available from [arXiv.org](https://doi.org/10.48550/arXiv.2305.18793).

Ding, Peng and Tirthankar Dasgupta.
2017.
[A randomization-based perspective on analysis of variance: a test statistic robust to treatment effect heterogeneity](https://doi.org/10.1093/biomet/asx059).
_Biometrika_ 105(1).
45--56.

## Software versions
```{r}
devtools::session_info("attached")
```

