---
draft: true
title: "Different distributions, same boxplot"
date: "2026-08-19"
categories: [graphs, R]
format:
  html:
    code-fold: show
    message: false
    df-print: paged
    fig-width: 6
    fig-height: 9
---

When I want to compare the distribution of some more or less continuous
variable between two or more groups, 
the [boxplot](../2016-06-21-drawing-a-boxplot/) is my default graph of
choice.
They compactly summarise a distribution in five numbers:
the minimum,
first quartile, median, third quartile, and maximum.
They avoid arbitrary choices such as histogram binning and, unlike
kernel density plots, don't assume that the underlying distribution
is smooth.

But boxplots can also be deceptive.
One important reason for this is that two distributions can be represented by the same
boxplot yet be qualitatively different.
This is why I encourage students and colleagues
to [overlay the individual data points](../2016-06-21-drawing-a-boxplot/#adding-the-individual-data-points) 
when using boxplots, as this helps to ward off some common misinterpretations
and may reveal some additional structure in the data.

To further drive this point home, I thought it'd be useful to have an R function
that can generate quite radically different distributions that would
produce identical boxplots---in much the same way
that I wrote a function that outputs different bivariate distributions
yielding the same correlation coefficient ([_What data patterns can lie behind a correlation coefficient?_](../2016-11-21-what-correlations-look-like/)).

The next code snippet defines such a function, `boxplot_data()`.
It takes, among other inputs,
a desired minimum, first quartile, median, third quartile, and maximum,
and generates data conforming to this five-number summary.
The parameter `k` governs the number of data points in the distribution,
and the other parameters regulate what the distribution looks like
in the intervals determined by the five-number summary.

```{r}
#| code-fold: true
boxplot_data <- function(
    k,                          
    minv, q1, q2, q3, maxv,    
    s = NULL,                 
    shapes = rep("uniform", 4), 
    margins_lo = c(0.05, 0.05, 0.05, 0.05),
    margins_hi = c(0.05, 0.05, 0.05, 0.05)) {
  # Generate univariate data whose five-number summary matches 
  # specified values. Data between these values follows user-specified
  # distributions.
  
  # k: Integer, k >= 2. Generates n = 4*k observations.
  # minv, q1, q2, q3, maxv: Desired five-number summary in ascending order.
  # s: Numeric vector (length 3) giving the gaps around q1, q2, and q3. 
  #    If NULL, a default gap equal to 10% of the smallest adjacent interval is used.
  # shapes: Character vector (length 4) specifying the distribution shape in each interval.
  # margins_lo, margins_hi: Numeric vectors (length 4) giving lower and upper insets 
  #                         (proportions) that keep generated values away from the interval endpoints.

  if (k < 2) stop("k must be >= 2")
  if (any(c(minv, q1, q2, q3, maxv) != sort(c(minv, q1, q2, q3, maxv)))) {
    stop("five-number summary must be in ascending order.")
  }
  n <- 4 * k
  
  # Symmetric gap around quartiles
  if (is.null(s)) {
    s <- min(q1 - minv, q2 - q1, q3 - q2, maxv - q3) * 0.1
    s <- rep(s, 3)
  }
  
  data <- numeric(n)
  data[1]       <- minv
  data[k]       <- q1 - s[1] # gap around first hinge
  data[k + 1]   <- q1 + s[1]
  data[2 * k]   <- q2 - s[2] # gap around median
  data[2*k + 1] <- q2 + s[2]
  data[3 * k]   <- q3 - s[3] # gap around second hinge
  data[3*k + 1] <- q3 + s[3]
  data[n]       <- maxv
  
  # Squish data into
  squish <- function(x) {
    x <- x - min(x)
    x <- x / max(x)
  }
  right_skewed <- function(m) {
    x <- abs(rnorm(m))
    squish(x)
  }
  left_skewed <- function(m) {
    x <- 1 - right_skewed(m)
    squish(x)
  }
  one_bump <- function(m) {
    x <- rnorm(m)
    squish(x)
  }
  two_bumps <- function(m) {
    x <- c(rnorm(floor(m/2), -5), rnorm(ceiling(m/2), 5))
    squish(x)
  }
  
  
  fill_segment <- function(v0, v1, m, shape, a0, b0) {
    if (m <= 0) return(numeric(0))
    u <- switch(shape,
                uniform    = runif(m),
                right_skew = right_skewed(m),
                left_skew  = left_skewed(m),
                one_bump   = one_bump(m),
                two_bumps  = two_bumps(m),
                stop("unknown shape: ", shape)
    )
    u <- a0 + (1 - a0 - b0) * u   # inset u away from 0/1
    sort(v0 + (v1 - v0) * u)
  }
  
  # Obtain bounds (ranks) for segments
  seg_bounds <- list(c(1, k), c(k+1, 2*k), c(2*k+1, 3*k), c(3*k+1, n))
  for (j in seq_along(seg_bounds)) {
    i0 <- seg_bounds[[j]][1]
    i1 <- seg_bounds[[j]][2]
    m <- i1 - i0 - 1
    if (m <= 0) next
    pts <- fill_segment(data[i0], data[i1], m, shapes[j], a0 = margins_lo[j], b0 = margins_hi[j])
    data[(i0 + 1):(i1 - 1)] <- pts
  }
  
  if (isFALSE(all.equal(fivenum(data), c(minv, q1, q2, q3, maxv)))) {
    warning("Not exact match with desired values.")
  }
  
  data
}
```

In the next snippet, I define the function `draw_boxplots()`,
which uses `boxplot_data()` to generate six datasets 
with identical five-number summaries and then plots them.

```{r}
#| code-fold: true
draw_boxplots <- function(k = 15, five_numbers = c(0, 7, 13, 27, 55), 
  show_mean = TRUE, show_data = FALSE, mfrow = c(2, 3), mar = c(1, 3, 4.1, 1), las = 1, ...) {
  # Plot 6 identical boxplots with different underlying distributions.
  
  draw_boxplot <- function(x, show_mean = TRUE, main = NULL) {
    boxplot(x, main = main, at = 1)
    points(x = jitter(rep(1, length(x)), 4), y = x)
    if (show_mean) points(x = 1, mean(x), col = "darkred", pch = 8)
  }
  
  five_numbers <- sort(five_numbers)
  minv <- five_numbers[1]
  maxv <- five_numbers[5]
  q1 <- five_numbers[2]
  q2 <- five_numbers[3]
  q3 <- five_numbers[4]
  
  if (minv < q1 - 1.5*(q3 - q1)) {
    minv <- q1 - 1.5*(q3 - q1)
    warning(paste0("Minimum value too far from box. Was reset to ", minv, "."))
  }
  if (maxv > q3 + 1.5*(q3 - q1)) {
    maxv <- q3 + 1.5*(q3 - q1)
    warning(paste0("Maximum value too far from box. Was reset to ", maxv, "."))
  }

  op <- par(no.readonly = TRUE)
  par(mfrow = mfrow, mar = mar, las = las, ...)
  
  # Spread out
  d1 <- boxplot_data(k, minv, q1, q2, q3, maxv, 
    shapes = c("uniform", "uniform", "uniform", "uniform"))
  draw_boxplot(d1, main = "Spread out data")
  
  # Gap around median
  d2 <- boxplot_data(k, minv, q1, q2, q3, maxv,  
    shapes = c("left_skew", "right_skew", "left_skew", "right_skew"),
    s = c(min(q1 - minv, 0.01), 0.7*min(q2 - q1, q3 - q2), min(maxv - q3, 0.01)))
  draw_boxplot(d2, main = "Gap around median")
  
  # Five numbers typical
  d3 <- boxplot_data(k, minv, q1, q2, q3, maxv,  
    shapes = c("two_bumps", "two_bumps", "two_bumps", "two_bumps"),
    s = rep(0, 3),
    margins_lo = rep(0, 4),
    margins_hi = rep(0, 4))
  draw_boxplot(d3, main = "Five modes")
  
  # All quartiles atypical
  d4 <- boxplot_data(k, minv, q1, q2, q3, maxv,  
    shapes = c("uniform", "uniform", "uniform", "uniform"),
    s = 0.5*c(min(q1 - minv, q2 - q1), min(q2 - q1, q3 - q2), min(q3 - q2, maxv - q3)),
    margins_lo = c(0.5, 0.5, 0.5, 0.5),
    margins_hi = c(0.5, 0.5, 0.5, 0.5))
  draw_boxplot(d4, main = "Quartiles atypical")
  
  # 5) Nearly empty box
  d5 <- boxplot_data(k, minv, q1, q2, q3, maxv,  
    shapes = c("uniform", "uniform", "uniform", "uniform"),
    s = c(0, min(q2 - q1, q3 - q2), 0),
    margins_lo = c(0, 0, 1, 0),
    margins_hi = rep(0, 4))
  draw_boxplot(d5, main = "Nearly empty box")
  
  # 6) Nearly empty whiskers
  d6 <- boxplot_data(k, minv, q1, q2, q3, maxv,  
    shapes = c("left_skew", "uniform", "uniform", "right_skew"),
    s = c(0, 0, 0),
    margins_lo = c(1, 0, 0, 0),
    margins_hi = c(0, 0, 0, 1))
  draw_boxplot(d6, main = "Nearly empty whiskers")
  
  par(op)
  
  if (show_data) {
    d <- data.frame(
      x = c(d1, d2, d3, d4, d5, d6),
      plot = rep(1:6, each = 4*k)
    )
    return(d)
  }
}
```

@fig-1 shows six boxplots, with the individual data points overlaid, 
that are obtained when using the default settings.

* The top row shows a distribution where the median is a pretty typical value (left)
  and a bimodal distribution in which the observations lie far from the median (right).
* The middle row shows a distribution with five modes around each of the value in the five-number summary (left)
  and a multimodal distribution in which none of the modes are centred around these five numbers (right).
* In the bottom row, either the box is pretty much empty (left), or the whiskers are pretty much empty (right).

```{r}
#| label: fig-1
#| fig-cap: "Six fairly standard boxplots but with different underlying distributions. The red stars show the distributions' means."
draw_boxplots(mfrow = c(3, 2))
```

@fig-2 shows a similar line-up of boxplots, but this time, they are all degenerate:
the first quartile and the minimum coincide.

```{r}
#| label: fig-2
#| fig-cap: "The first quartile is also the minimum."
draw_boxplots(five_numbers = c(0, 0, 13, 27, 55), mfrow = c(3, 2))
```

The situation is even worse in @fig-3: Now the _median_ and the minimum coincide.
(In this case, there can't be a gap around the median.)

```{r}
#| label: fig-3
#| fig-cap: "The median is also the minimum."
draw_boxplots(five_numbers = c(0, 0, 0, 3, 6.5), mfrow = c(3, 2))
```

@fig-4 and @fig-5 highlight a counterintuitive aspect of boxplots:
A _larger_ box corresponds to a _lower_ density in the middle 50% of the distribution,
whereas a more compact box means that the middle 50% of the distribution is more tightly
clustered around the median.

```{r}
#| label: fig-4
#| fig-cap: "A large box and tiny whiskers means that most data is concentrated near the extremes."
draw_boxplots(five_numbers = c(0.2, 0.3, 3, 4.9, 5), mfrow = c(3, 2))
```

```{r}
#| label: fig-5
#| fig-cap: "And a compact box means that the data are more densely distributed in the middle 50% of the distribution."
draw_boxplots(five_numbers = c(0.2, 2.1, 3, 3.5, 5), mfrow = c(3, 2))
```

Hopefully this convinces someone somewhere to add individual data points
to their boxplots!

## Software versions
```{r}
devtools::session_info("attached")
```

