---
title: "When interaction effects are hard to explain, try nested effects"
date: "2026-06-03"
categories: [contrast coding, multiple regression, R, tutorial]
format:
  html:
    code-fold: true
    message: false
    df-print: paged
    fig-width: 8
    fig-height: 6
    fig-align: center
knitr:
  opts_chunk:
    out.width: "100%"
---

When analysing a factorial experiment,
R's default regression output gives you the main effects and interaction effects.
The meaning of the resulting parameter estimates can be tricky to explain,
especially that of the interaction effect:
in essence, it tells you what the difference between two differences is.
Nested effects, which immediately tell you what the differences themselves are,
are often easier to interpret.
This post shows how you can obtain such nested effects in R.

## Example data
@fig-example-data shows [data](berthele2012.csv) from an experiment with a two-by-two design
carried out by Berthele (2012).
French-speaking students at a university for teacher education were 
played a recording of what they thought was a first-language German boy
speaking French.
About half of the students were told that the boy had a typical Swiss-German
name (e.g., Luca); the others were told that he had a name associated
with the Balkan region (e.g., Dragan).
Furthermore, about half of the recordings contained code-switches (i.e.,
the boy would occasionally use German expressions when speaking French);
the other half of the recordings did not contain any such code-switches.
The future teachers were asked to rate the boy's academic potential on a
scale from 1 to 6, with 6 being the highest.

```{r}
#| warning: false
#| message: false
#| fig-height: 4.5
#| label: fig-example-data
#| fig-cap: "Data from Berthele's (2012) experiment. Each circle shows a rating given by a teacher trainee. The red stars highlight the cell means."
library(tidyverse)
theme_set(theme_bw())

d <- read_csv("berthele2012.csv")
ggplot(d, 
       aes(x = CS, y = Potential)) +
  geom_point(shape = 1, position = position_jitter(width = 0.2, height = 0)) +
  geom_point(data = d |> 
               group_by(CS, Name) |> 
               summarise(Potential = mean(Potential), .groups = "drop"),
             shape = 8, size = 5, colour = "darkred") +
  facet_grid(cols = vars(Name)) +
  xlab("Code-switching?") +
  ylab("Rated academic potential")
```

## Interaction effects
@fig-example-data suggests that there is some **interaction** going on in the data:
'Dragan's' ratings tend to be higher than 'Luca's' when there is no code-switching,
but lower where there is code-switching.
It's pretty easy to model this interaction in R:^[Here I'm ignoring the rather coarse nature of the data. If you want to go the whole hog, you can use bootstrapping to verify the standard errors, but this would detract from the blog post's goal.]

```{r}
#| code-fold: show
mod.lm <- lm(Potential ~ Name*CS, data = d)
summary(mod.lm)$coefficients
```

You need a bit of practice to wrap your head around what these 
parameter estimates actually refer to, however.
The parameter estimate for `NameLuca` ($0.40 \pm 0.20$), for instance, 
tells you that the ratings for 'Luca' tend to be $0.40$ points higher
than those for 'Dragan' _if the recording contains code-switches_.
Next, the parameter estimate of `CSwithout` ($0.66 \pm 0.20$),
tells you that the ratings for recordings without code-switches
tend to be $0.66$ points higher than those with code-switches
_if the students were told that the boy is called Dragan_.
The estimate of the interaction parameter `NameLuca:CSwithout` ($-0.93 \pm 0.28$)
tells you that the absence of code-switches was $0.93$ points more detrimental
for 'Luca' than it was for 'Dragan'.
_Equivalently_, it tells you that the _presence_ of code-switches was 
$0.93$ points more detrimental for 'Dragan' than it was for 'Luca'.
_Also equivalently_, it tells you that labelling the recording 'Dragan'
instead of 'Luca' is $0.93$ points more detrimental on average for recordings
with code-switches than for recordings without code-switches.

If you skipped the previous paragraph: I understand.
Even for a fairly simple two-by-two design, 
figuring out the precise meaning of the estimated parameters requires sitting
down with a pencil and a piece of paper; see Chapter 10 
in my [statistics booklet](https://github.com/janhove/AnalysingQuantitativeData).
Thankfully, however,
you can usually make the output of statistical models such as the
one above more readily interpretable both to yourself and to your audience
by changing the coding scheme.
I've explained how this can be done in a [previous blog post](../2020-05-05-contrast-coding).

## Nested effects using R's shorthand notation
Whereas an interaction effect tells you _how much stronger_ an effect of one factor (e.g., code-switching) is
for one level of another factor (e.g., if the name is Luca) than for the other level (e.g., if the name is Dragan),
**nested effects** tell you quite simply _what_ the effect of one factor
is for each level of another factor.
To fit nested effects in R, you can use the slash notation:

```{r}
#| code-fold: show
mod.lm <- lm(Potential ~ Name/CS, data = d)
summary(mod.lm)$coefficients
```

As before, the estimated intercept ($2.96 \pm 0.16$) shows the average rating
given to Dragan recordings with code-switches,
and the `NameLuca` estimate ($0.40 \pm 0.20$) shows that the Luca recordings
with code-switches were rated a bit better on average than these.
Again like before, the `NameDragan:CSwithout` ($0.66 \pm 0.20$) shows
the effect of the absence of code-switches on the ratings given to Dragan.
What's different from before is the `NameLuca:CSwithout` estimate ($-0.27 \pm 0.19$):
it shows that, for 'Luca', recordings without code-switches tended to be rated
$0.27$ worse than recordings with code-switches.

The next snippet shows an alternative coding scheme that suppresses the intercept
using the `0` notation:
```{r}
#| code-fold: show
mod.lm <- lm(Potential ~ 0 + Name/CS, data = d)
summary(mod.lm)$coefficients
```

This makes it clearer that the previous intercept estimate refers to the
Dragan recordings exclusively.
The parameter estimate for `NameLuca` ($3.36 \pm 0.13$) now shows
the average rating for a Luca recording with code-switches
rather than a difference between two averages.
The remaining estimates refers to the same nested effects as before.

Of course, we can switch the order in which the factors appear.
In doing so, we obtain the average ratings for the Dragan recordings
with and without code-switches as well as the effects of labelling
the recordings 'Luca' rather than 'Dragan' for both recordings with
and without code-switches:

```{r}
#| code-fold: show
mod.lm <- lm(Potential ~ 0 + CS/Name, data = d)
summary(mod.lm)$coefficients
```

## Nested effects using a custom coding scheme
One of the representations above is sufficient for our purposes.
If you want even more control over what the estimates refer to,
I recommend you use a custom coding scheme.
In this example, it's a bit unfortunate that the `with` code-switching
condition gets absorbed in the intercept but the `without` condition doesn't.
By using a custom coding scheme, we can have the estimated parameters
mean exactly what we want them to mean.

First, we explicitly label each cell of the design and convert these labels to
a factor:
```{r}
#| code-fold: show
d$Cell <- paste0(d$Name, "-", d$CS) |> 
  as.factor()
table(d$Cell)
```

Next, we write down what we want the intercept to represent.
I think that a sensible choice is to have the intercept represent the mean
of the two cells in which no code-switching occurred, i.e., `Dragan-without`
and `Luca-without`.^[Other choices are defensible, too. In fact, if this technique is new to you, I suggest you try to adapt
the code snippets so that the intercept represents the grand mean (the average of all cell averages)
or just Luca's code-switch-free recordings.]
We write this down in the form of a null hypothesis, even if we don't really
want to test this null hypothesis:

$$\frac{1}{2}\left(\mu_{\textrm{Luca, no CS}} + \mu_{\textrm{Dragan, no CS}}\right) = 0.$$

We then rewrite this null hypothesis so that all four cells are represented on the left-hand side
of the equation. 
In doing so, make sure that the cells occur in the same order in which they appear above,
i.e., first Dragan with code-switches, then Dragan without, then Luca with, and finally 
Luca-without.
(If you find this confusing, either bear with me or read the longer blog post
on [contrast coding](../2020-05-05-contrast-coding).)

$$\frac{1}{2}\left(\mu_{\textrm{Luca, no CS}} + \mu_{\textrm{Dragan, no CS}}\right) = 0
\Leftrightarrow
0\mu_{\textrm{Dragan, with CS}}
+ \frac{1}{2}\mu_{\textrm{Dragan, no CS}}
+ 0\mu_{\textrm{Luca, with CS}}
+ \frac{1}{2}\mu_{\textrm{Luca, no CS}} = 0.$$

We put the coefficients so obtained (i.e., 0, 1/2, 0 and 1/2) into a vector.
The vector's name makes it clear that its purpose is to estimate the 
mean of the averages of the cells in which no code-switching occurred;
the vector entries are named for clarity's sake (e.g., `Ln` means Luca, no code-switches).

```{r}
#| code-fold: show
Mean_NoCS <- c(Dw = 0, Dn = 1/2, Lw = 0, Ln = 1/2)
```

We now write down what we want the next parameter estimate to refer to.
I want it to represent the difference between Dragan's code-switch-free ratings
and Luca's code-switch-free ratings.
We can write this down as a null hypothesis like so:

$$\left(\mu_{\textrm{Dragan, no CS}} = \mu_{\textrm{Luca, no CS}}\right)
\Leftrightarrow
0\mu_{\textrm{Dragan, with CS}}
+ 1\mu_{\textrm{Dragan, no CS}}
+ 0\mu_{\textrm{Luca, with CS}}
- 1\mu_{\textrm{Luca, no CS}} = 0.$$

Again, we put the resulting coefficients in a vector.
```{r}
#| code-fold: show
NoCS_DraganvLuca <- c(Dw = 0, Dn = 1, Lw = 0, Ln = -1)
```

I want the third parameter estimate to show the (nested) code-switching
effect for Luca:

$$\left(\mu_{\textrm{Luca, with CS}} = \mu_{\textrm{Luca, no CS}}\right)
\Leftrightarrow
0\mu_{\textrm{Dragan, with CS}}
+ 0\mu_{\textrm{Dragan, no CS}}
+ 1\mu_{\textrm{Luca, with CS}}
- 1\mu_{\textrm{Luca, no CS}} = 0.$$

Let's put this into a vector, too:
```{r}
#| code-fold: show
Luca_CS <- c(Dw = 0, Dn = 0, Lw = 1, Ln = -1)
```

The fourth and final parameter estimate should show the (nested) code-switching
effect for Dragan:
$$\left(\mu_{\textrm{Dragan, with CS}} = \mu_{\textrm{Dragan, no CS}}\right)
\Leftrightarrow
1\mu_{\textrm{Dragan, with CS}}
- 1\mu_{\textrm{Dragan, no CS}}
+ 0\mu_{\textrm{Luca, with CS}}
+ 0\mu_{\textrm{Luca, no CS}} = 0.$$

```{r}
#| code-fold: show
Dragan_CS <- c(Dw = 1, Dn = -1, Lw = 0, Ln = 0)
```

Now we collect the four vectors into a hypothesis matrix `Hm`:
```{r}
#| code-fold: show
Hm <- rbind(
  Mean_NoCS,
  NoCS_DraganvLuca,
  Luca_CS,
  Dragan_CS 
)
```

Using the `apply_contrasts()` function defined in the next snippet,
the hypothesis matrix can be converted into a set of contrasts that 
can be applied to the `Cell` factor:

```{r}
#| code-fold: show
apply_contrasts <- function(Hm) {
    MASS::fractions(provideDimnames(MASS::ginv(Hm), base = dimnames(Hm)[2:1]))[, -1]
}
contrasts(d$Cell) <- apply_contrasts(Hm)
```

Now fit the model using the `Cell` factor with the custom coding scheme as a predictor:
```{r}
#| code-fold: show
mod.lm <- lm(Potential ~ Cell, d)
summary(mod.lm)$coefficients
```

As you can verify, each parameter estimate refers exactly to what we wanted
it to refer to.

## Interaction effect or nested effects?
In the example above, the nested effects tell you what the effect of code-switching
is for Luca recordings and for Dragan recordings,
whereas the interaction effect tells you by how much these differ.
I find the nested effects easier to interpret.
But if you're interested in the _difference_ between these nested effects
(and this is often the case!), you also need the interaction effect.
What you could do in this case is have your cake and eat it, too:
report the easy-to-interpret nested effects in a table,
and report the interaction effect in the text if it's also relevant.
Perhaps something along the lines of 

> "Luca recordings with code-switches received ratings that
> were $0.27 \pm 0.19$ points higher on average than
> did Luca recordings without code-switches.
> The opposite was true for 'Dragan': if he code-switched, the 
> ratings were $0.66 \pm 0.20$ lower than when he did not.
> The difference between these code-switching effects is statistically significant
> ($0.93 \pm 0.28$, $t(151) = 3.4, p = 0.0009$).

The interaction estimate and standard error as well as the significance test
can be gleaned from the output of the model at the top of this page.


## Conclusion
In sum, if you're analysing a study with a factorial design and 
you find that you're tripping yourself up when explaining the literal meaning
of the parameter estimates, consider using a nested-effects parametrisation
instead of an interaction-effect parametrisation.
If you want full control, use a custom-coding scheme.


## Reference
Berthele, Raphael.
2012.
[The influence of code-mixing and speaker information on perception and assessment of foreign language proficiency: An experimental study](https://doi.org/10.1177/1367006911429514).
_International Journal of Bilingualism_ 16(4).
453--466.

## Session info
```{r}
devtools::session_info("attached")
```

