---
title: "The population model and the randomisation model of statistical inference"
date: "2025-11-09"
categories: [assumptions, experiments, design features, nonparametric tests, R, significance]
format:
  html:
    code-fold: true
    message: false
---

Much of statistical practice---and teaching---relies on an assumption
that is demonstrably false:
the fiction that the data were sampled randomly from some population.
Now, in practice, this fiction turns out to be quite useful.
But I think that it's important to appreciate that it is a fiction
and that it's possible to do statistics without subscribing to it.

Let's first take a look at the ubiquitous Student's _t_-test for independent samples.
This test is used to check if the difference between the means of two lists of numbers,
$\boldsymbol x = (x_1, \dots, x_k)$ and $\boldsymbol y = (y_1, \dots, y_m)$,
could plausibly be explained by chance.
It is based on the following assumptions:

1. The observations $\boldsymbol x$ and $\boldsymbol y$ constitute 
   simple random samples from some populations $P_1$ and $P_2$, respectively.
2. Population $P_1$ follows a normal distribution with mean $\mu_1$ and variance $\sigma^2$.
3. Population $P_2$ follows a normal distribution with mean $\mu_2 = \mu_1 + \delta$ and variance $\sigma^2$.

If these assumptions are well-founded, 
the Student's _t_-test provides an exact test for the null hypothesis that $\delta = \delta_0$ 
for some specified number $\delta_0$, where typically $\delta_0 = 0$.
(In this context, `exact' means that the probability that the test yields a _p_-value
no larger than $\alpha$ if the null hypothesis is true is itself no larger than $\alpha$
for _any_ $\alpha \in (0, 1)$.)

That's a big _if_.

Thanks to the central limit theorem, though, 
the normality assumption isn't _too_ crucial:
for "large enough" samples from non-normal distributions,
the _t_-test behaves "more or less" as it does for normal distributions,
at least under fairly general conditions such as that the distributions
that are sampled have finite variance.
And except for the smallest of samples, 
the equality of variances assumption
can be dealt with by using Welch' _t_-test instead of Student's.
I don't want to go into what "large enough" and "more or less" mean in this context;
interested readers may want to look up what "convergence in distribution" means.

But what strikes me as more fundamental than the shape of the populations $P_1$ and
$P_2$ are the questions
_Who's drawing random samples from a population?_
and _What's the population, anyway?_

Sure, it's easy to set up computer simulations in which the data are sampled from some population.
But at the coalface, 
most research is done with volunteers (or students needing credits)
who've undergone a round of screening.
And even if there is some population that these participants
can be considered to be a sample from,
their recruitment didn't involve any mathematically 
well-defined sampling procedure.
This **population model** of statistical inference,
which treats data as random samples from some population (normal or other), 
is, then, mostly based on a fiction.

The population model of statistical inference contrasts with the **randomisation model**.
In the population model,
the randomness in the results is assumed to be caused by random sampling;
in the randomisation model,
it is assumed to be caused by randomly assigning the units 
(e.g., the participants) in the study to different conditions.
Whereas random sampling is exceedingly rare in the social sciences,
random assignment is quite common.
And if random assignment was used, the randomisation model results in
valid hypothesis tests that do not make any unverifiable distributional assumptions.

In what follows, I'll illustrate the randomisation model in the simplest of settings.

## A two-group comparison
Consider a two-group study in which $k + \ell = m$ participants 
are randomly assigned to one of two groups such that $k$ are assigned 
to the first group and $\ell$ to the second group.
The participants in the first group engage in some activity or are subjected
to some treatment, and those in the second group engage in another activity
or are subjected to another treatment.
For each participant $i = 1, \dots, m$, we observe a numerical outcome $Y_i$,
and the question of interest is whether the two treatments differentially
affected the outcome.

To make the following a bit more concrete, I'll start out by
the fictitious data summarised in @tbl-summary and shown in @fig-boxplot.
Fictitious data are a bit boring, so I'll apply the techniques discussed
to a real but somewhat more complicated example later on.

```{r}
#| label: tbl-summary
#| tbl-cap: "Results summary of a fictitious experiment."
suppressPackageStartupMessages(library(tidyverse))
theme_set(theme_bw())

d <- structure(list(Group = c(0, 0, 0, 0, 1, 0, 1, 0, 1, 1, 1, 0,
1, 1, 1, 1, 0, 1, 0, 1, 0, 0, 1, 1, 1, 1, 0), Outcome = c(7,
8, 9.5, 8.5, 7.5, 9, 8, 7.5, 10, 8, 8.5, 8, 9, 10.5, 8, 9, 7.5,
9, 9.5, 11, 8, 8, 7.5, 9, 9.5, 8.5, 8.5)), class = "data.frame", row.names = c(NA,
-27L))

d |> 
  group_by(Group) |> 
  summarise(
    n = n(),
    Mean = mean(Outcome) |> round(2),
    Median = median(Outcome),
    `Standard deviation` = sd(Outcome) |> round(2)
  ) |> 
  knitr::kable()
```

```{r, fig.height = 4, fig.width = 5, out.width = "80%"}
#| label: fig-boxplot
#| fig.cap: "Data for a fictitious randomised experiment with two groups."
d |> 
  ggplot(aes(x = factor(Group), y = Outcome)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(shape = 1, position = position_jitter(width = 0.1, height = 0)) +
  scale_x_discrete(name = "Condition",
    limits = c("0", "1"), labels = c("0", "1"))
```

### The randomisation model: Theory
For each participant $i = 1, \dots, m$, 
there exist two potential outcomes $y_i(0)$ and $y_i(1)$, 
only one of which is observed:

* either $y_i(0)$ if the participant is assigned to the first condition;
* or $y_i(1)$ if the participant is assigned to the second condition.

In the randomisation model, these potential outcomes are treated as
fixed, though possibly unknown, quantities, and not as random variables
having some distribution.
What _is_ random is which of these two quantities we end up observing.
Specifically, let $B_i$ be the random variable that indicates
which condition the $i$-th participant is assigned to.
Then $B_i = 0$ if the participant is assigned to the first condition
and $B_i = 1$ if the participant is assigned to the second condition.
Then the observed outcome for the $i$-th participant is
$$Y_i = y_i(B_i) = \begin{cases}y_i(0) &\textrm{if } B_i = 0,\\ y_i(1) &\textrm{if } B_i = 1.\end{cases}$$

The researchers control the randomisation,
so they _know_ what the distribution of the assignment vector 
$\boldsymbol B := (B_1, \dots, B_m)$ is.
In the present example, the assignment vector was picked randomly
from the set of vectors with length $12 + 15 = 27$ that contain
exactly $12$ zeroes and $15$ ones, with each such vector having the
same probability of being picked.
Formally, 
$$\boldsymbol B \sim \textrm{Uniform}(\mathcal{B}_{12, 15}),$$
where
$$\mathcal{B}_{k, \ell} := \{\boldsymbol b \in \{0, 1\}^{k + \ell} : \sum_{i=1}^{k + \ell} b_i = \ell\}.$$
According to the so-called **sharp null hypothesis**, 
there is no differential treatment effect whatsoever.
That is,
$$H_0: y_i(1) = y_i(0) \textrm{ for }  i = 1, \dots, m.$$
Under such a sharp null hypothesis,
the values for $y_i(0)$ and $y_i(1)$ can immediately be reconstructed 
from the observed outcomes: $y_i(0) = y_i(1) = Y_i$ for $i = 1, \dots, m$.

More generally, for any number $\delta$, we could set up the sharp
null hypothesis that the second treatment boosts (or would have boosted)
the outcome by $\delta$ for all participants:
$$H_0(\delta): y_i(1) = y_i(0) + \delta \textrm{ for }  i = 1, \dots, m.$$
Under $H_0(\delta)$, too, we can straightforwardly reconstruct the 
$y_i$ values for all participants:
$y_i(0) = Y_i - B_i\delta$ and $y_i(1) = Y_i + (1-B_i)\delta$
for $i = 1, \dots, m$.

An exact _p_-value testing such sharp null hypotheses
can now be computed as follows without needing to assume
that the $y_i(0)$ and $y_i(1)$ values are randomly sampled from some population.

1. Decide on a relevant **test statistic** $T_{\delta}(\boldsymbol y, \boldsymbol b)$.
   The _p_-value that eventually results 
   will be valid regardless of the test statistic chosen.
   But ideally, a test statistic is chosen that tends to be small (or close to zero)
   under the null hypothesis but large (or farther from zero)
   if the null hypothesis isn't true.
   For the current setting, you could compute, for instance, 
   the difference between the group means, 
   or the difference between the group medians,
   or the probability that a randomly chosen participant 
   who was assigned to the second condition has a higher score
   than a randomly chosen participant from the first condition.

2. Compute the test statistic for the data at hand: $T_{\delta, \textrm{obs}} := T_{\delta}(\boldsymbol Y, \boldsymbol B)$.

3. For a left-sided _p_-value, 
   compute the probability that $T_{\delta}(\boldsymbol Y, \boldsymbol {\tilde B}) \leq T_{\delta, \textrm{obs}}$,
   where $\boldsymbol {\tilde B}$ has the same distribution as $\boldsymbol B$ 
   (here: $\boldsymbol {\tilde B} \sim \textrm{Uniform}(\mathcal{B}_{k, \ell})$).
   This can be computed by going through all the possible assignment vectors
   $\boldsymbol b$ and counting how often 
   $T_{\delta}(\boldsymbol Y, \boldsymbol b) \leq T_{\delta, \textrm{obs}}$.
   For a right-sided _p_-value, instead compute the probability that
   $T_{\delta}(\boldsymbol Y, \boldsymbol {\tilde B}) \geq T_{\delta, \textrm{obs}}$.
   A two-sided _p_-value can be computed as twice the smaller of the left-side
   and the right-sided _p_-value.

Step 3 poses a practical challenge: 
the set of possible assignment vectors
is usually too large to handle.
In our case, the set $\mathcal{B}_{k, \ell}$ contains
$${{k + \ell}\choose {k}} = \frac{(k + \ell)!}{k!\ell!}$$
elements.
Even for fairly small $k, \ell$, 
this are too many vectors to iterate over.
For instance, for $k = 12, \ell = 15$, we obtain 17,383,860
possible assignment vectors;
for $k = \ell = 20$, there are 137,846,528,820 such vectors.

What we can do instead is sample $M$ assignment vectors at random
from the distribution of $\boldsymbol B$
and count how often the resulting test statistic is no larger
than $T_{\delta, \textrm{obs}}$ (for a left-sided _p_-value)
or no smaller than $T_{\delta, \textrm{obs}}$ (for a right-sided _p_-value).
This number plus one and then divided by $M+1$ is
an exact left- or right-sided _p_-value.
A two-sided _p_-value can then again be calculated as twice the
smaller of these two one-sided _p_-values.

### The randomisation model: Practice
Let's put this theory into practice.

First, we need to decide on a test statistic.
In the code snippet below, I define five such statistics, mainly
to illustrate the flexibility of the approach.

1. The difference between the means in the 0 condition and in the 1 condition.
2. The difference between the medians in the 0 condition and in the 1 condition.
3. The difference between the [midmeans](https://en.wikipedia.org/wiki/Interquartile_mean) in the 0 condition and in the 1 condition.
4. The rank sum of the observations in the 1 condition. To obtain this statistic, compute the ranks of all observations, and then sum the ranks of those in the 1 condition.
5. The probability of superiority: If you pick a random observation in the 1 condition and compare it to a randomly chosen observation the 0 condition, what's the probability that the 1 observation is higher than the 0 observation? Both the rank sum and the probability of superiority are closely connected to the Wilcoxon--Mann--Whitney tests.

```{r}
# All functions assume that one of the groups is coded as 0 and the other as 1.
mean_diff <- function(x, group) {
  mean(x[group == 1]) - mean(x[group == 0])
}
median_diff <- function(x, group) {
  median(x[group == 1]) - median(x[group == 0])
}
midmean_diff <- function(x, group) {
  mean(x[group == 1], trim = 0.25) - mean(x[group == 0], trim = 0.25)
}
ranksum <- function(x, group) {
  ranks <- rank(x)
  sum(ranks[group == 1])
}
probsup <- function(x, group) {
  n0 <- sum(group == 0)
  n1 <- sum(group == 1)
  (ranksum(x, group) - n1*(n1 + 1)/2) / (n0 * n1)
}
```

For the present data, the five observed test statistics are the following:
```{r}
#| label: tbl-teststats
#| tbl-cap: "Observed test statistics."
d |> 
  summarise(
    `Mean difference` = mean_diff(Outcome, Group),
    `Median difference` = median_diff(Outcome, Group),
    `Midmean difference` = midmean_diff(Outcome, Group),
    `Rank sum` = ranksum(Outcome, Group),
    `Probability of superiority` = probsup(Outcome, Group)
  ) |> 
  pivot_longer(everything(), names_to = "Test statistic",
               values_to = "Value") |>
  mutate(Value = round(Value, 2)) |> 
  knitr::kable()
```

The next code snippet defines the function `randomisation_test()`,
which computes a left-sided, right-sided or two-sided (= default) _p_-value
by applying a randomisation test for the test statistic of choice,
as explained above.
By default, 19,999 new assignment vectors are randomly generated
under the assumption that $\boldsymbol B \sim \textrm{Uniform}({\mathcal{B}_{k, \ell}})$.

```{r}
randomisation_test <- function(
    # Randomisation test of sharp null in two-group comparison.
    x,           # outcome
    group,       # grouping variable
    fun,         # function for computing test statistic from x and group
    delta = 0,   # delta for sharp null hypothesis
    alternative = "two_sided", # two-sided, left-sided or right-sided p-value?
    M = 19999,   # number of new randomisations generated
    plot = TRUE, # plot histogram of test statistic under H0?
    ...          # parameters passed to histogram
) {
  test_stat <- fun(x, group)
  x[group == 1] <- x[group == 1] - delta
  test_stat_H0 <- replicate(M, {
    fun(x, sample(group))
  })
  test_stat_H0 <- c(test_stat, test_stat_H0)
  
  if (plot) {
    hist(test_stat_H0, ...)
    abline(v = test_stat, lwd = 2)
  }
  
  p_left <- mean(test_stat_H0 <= test_stat)
  if (substr(alternative, 1, 1) == "l") return(p_left)
  
  p_right <-mean(test_stat_H0 >= test_stat)
  if (substr(alternative, 1, 1) == "r") return(p_right)
  
  min(2 * min(p_left, p_right), 1)
}
```

@fig-meandiff shows the distribution of the mean difference
under the sharp null hypothesis for $\delta = 0$.
The vertical line highlights the observed mean difference.
The resulting two-sided _p_-value is about 0.12.

```{r, fig.height = 4, fig.width = 5, out.width = "80%"}
#| label: fig-meandiff
#| fig-cap: "The distribution of the mean difference under the sharp null hypothesis for $\\delta = 0$."
set.seed(2025-11-07) # the day I wrote this
randomisation_test(d$Outcome, d$Group, mean_diff, 
  breaks = 50, xlab = "Mean difference", main = "Distribution under H0")
```

While the distribution of the mean difference under $H_0$
looks roughly bell-shaped but with some gaps,
the distribution of the median difference shown in @fig-mediandiff
looks much weirder.
But no distributional assumptions are required for computing the
randomisation-based _p_-value, and so the resulting _p_-value
of 0.18 is also a valid, exact _p_-value testing $H_0$.

```{r}
#| label: fig-mediandiff
#| fig-cap: "The distribution of the median difference under the sharp null hypothesis for $\\delta = 0$."
randomisation_test(d$Outcome, d$Group, median_diff, 
  breaks = 50, xlab = "Median difference", main = "Distribution under H0")
```

Similarly, we can compute exact _p_-values for the other test statistics,
or indeed for any statistic you can come up with.
For the midmean, we obtain $p = 0.20$; for the rank sum and the probability
of superiority $p = 0.14$.

```{r}
randomisation_test(d$Outcome, d$Group, midmean_diff, plot = FALSE)
randomisation_test(d$Outcome, d$Group, ranksum, plot = FALSE)
randomisation_test(d$Outcome, d$Group, probsup, plot = FALSE)
```

All of these _p_-values are exact for tests of the sharp null hypothesis. 
In fact, any test statistic can be used, 
and the resulting inference will still be exact as long as 
the randomisation procedure is correctly mirrored in the computation. 
Computing tests for several test statistics is a bad idea, however,
as you're just creating a multiple-comparisons problem.
At present, I'm not aware of general rules of thumb for choosing among different
test statistics, beyond the simple advice that the test statistic 
should be sensitive to the kinds of departures from the 
null hypothesis that are scientifically relevant.
But I'll look into it.

What may not sit so well with some readers is that repeated runs 
of the same randomisation test won't produce identical results. 
This is due to the randomness involved in choosing the assignment vectors.
For instance, another run of the randomisation test for the mean difference
yields a slightly different result from the first one.
```{r}
randomisation_test(d$Outcome, d$Group, mean_diff, plot = FALSE)
```
Increasing $M$ decreases this variability, but crucially,
the validity of the _p_-value does not depend on the specific choice of $M$.
In practical terms, preregistering a test statistic, a random seed, and $M$
strikes me as a good idea.

### The population model: A fiction but a useful one
Student's two-sample _t_-test serves as the classical population-model counterpart
to the randomisation test for the mean difference.
In the present example, the _p_-value it outputs (0.10) is similar but 
not quite identical to the one we obtained above.

```{r}
t.test(d$Outcome ~ d$Group, var.equal = TRUE)
```

This is typical:
it's surprisingly difficult to construct realistic examples,
even with simulated data, where the _t_-test yields a markedly different
_p_-value from the randomisation test for the mean difference.
This similarity is not coincidental.
Indeed,
the _t_-test and its cousins gained their popularity precisely because, in the pre-computer era,
they provided a computationally tractable approximation of randomisation tests.
But just because the _t_-test generally produces similar _p_-values
to those obtained by an exact procedure doesn't mean that the _t_-test is
also exact.

(Incidentally, population-based approaches tend to produce results
close to those obtained under the randomisation model
because, for moderately large samples and under mild regularity conditions, 
the distribution of the mean difference under
all possible random permutations of the group
labels is approximately normal.)

## The scope of inference
Under the randomisation model, the null hypothesis 
concerns only the specific experimental units included in the study.
As a result, any _statistical_ inference drawn is about
what happened in the experiment, 
not about what might happen in a larger population (whatever it may be).
Since population-model tests used in randomised experiments are useful
by virtue of how well they approximate randomisation-model methods,
any statistical inferences they enable still concern only the experimental
units included in the study.

This is not to say that generalisation beyond the experimental units is impossible.
But it must be based on subject-matter arguments, not on statistical ones.
In practice, this is precisely what reasonable researchers do:
they interpret each others' and their own findings by taking external validity into account.

## What about random group sizes?
Before we apply the randomisation model to a real dataset,
it is worth considering experiments in which
the group sizes aren't fixed in advance.
Suppose we have an experiment with $m$ experimental units (e.g., participants)
that are randomly assigned to two conditions.
Unlike in our previous discussion,
the number of units assigned to each condition isn't predetermined.
For instance, each participant may be assigned independently
to the first or second condition with equal probability.

In this case, the assignment vector $\boldsymbol B$ isn't distributed
according to $\textrm{Uniform}(\mathcal{B}_{k, \ell})$
with fixed group sizes $k, \ell$.
Instead, $\boldsymbol B$ follows some distribution over $\{0, 1\}^m$.
Fortunately, we can still apply the same randomisation test as before:
we simply use the observed number $K$ of units assigned to the first condition 
and the observed number $L$ of units assigned to the second condition
in lieu of some fixed $k$ and $\ell$.
In statistical parlance, we condition on the observed group sizes.

The reason why we can condition on the observed group sizes is as follows.
Let $P$ denote the _p_-value obtained from
the randomisation test conditional on the groups sizes $K$ and $L$
under the sharp null hypothesis.
For any fixed group sizes $k, \ell$, the randomisation test is exact.
Hence, $\textrm{Prob}(P \leq \alpha | K = k, L = \ell) \leq \alpha$
for any $\alpha \in (0, 1)$ and for any $k, \ell$ with $k + \ell = m$.
So, under the null,
\begin{align*}
  \textrm{Prob}(P \leq \alpha)
  &= \sum_{\boldsymbol b \in \mathcal{B}_{m}} \underbrace{\textrm{Prob}(P \leq \alpha | K = m - L, L = \sum_{i = 1}^m b_i)}_{\leq \alpha}\textrm{Prob}(K = m - L, L = \sum_{i = 1}^m b_i) \\
  &\leq \alpha \sum_{\boldsymbol b \in \mathcal{B}_{m}} \textrm{Prob}(K = m - L, L = \sum_{i = 1}^m b_i) \\
  &= \alpha,
\end{align*}
for any $\alpha \in (0,1)$.
So $P$ is an exact _p_-value even when the group sizes are random rather than
fixed.

The same reasoning works if we condition on, say, the total number
of experimental units. So randomisation tests also work when
the participants trickle into the lab and the total number of participants
isn't known in advance.

(Alternatively, we could write a slight variant of the 
`randomisation_test()` function in which the new assignment vectors
$\boldsymbol b$ are sampled from the unconditional distribution.)

## A real example
Klein et al. (2014) replicated a study originally devised by Oppenheimer
and Monin (2009) at 36 different sites.
This is their description:

> ``Oppenheimer & Monin (2009) investigated whether the rarity of an independent, 
> chance observation influenced beliefs about what occurred before that
> event. Participants imagined that they saw a man rolling dice in a casino. In
> one condition, participants imagined witnessing three dice being rolled and
> all came up 6's. In a second condition two came up 6's and one came up 3.
> In a third condition, two dice were rolled and both came up 6's. All 
> participants then estimated, in an open-ended format, how many times the man
> had rolled the dice before they entered the room to watch him. Participants
> estimated that the man rolled dice more times when they had seen him roll
> three 6's than when they had seen him roll two 6's or two 6's and a 3. For the
> replication, the condition in which the man rolls two 6's was removed leaving
> two conditions.''

Let's use the data from the Brasília site as a more realistic example;
see @tbl-summaryklein and @fig-boxplotklein.
The participants were randomly assigned to the conditions, but without
enforcing specific group sizes.
For the randomisation tests, we'll condition on the observed group sizes.

```{r}
d <- read.csv("Klein2014_gambler.csv") |> 
  filter(Sample == "brasilia") |> 
  filter(!is.na(RollsImagined))
d$n.Condition <- ifelse(d$Condition == "three6", 1, 0)
```

```{r}
#| label: tbl-summaryklein
#| tbl-cap: "Results summary of Klein et al.'s replication of the gambler's fallacy study in Brasília."
d |> 
  group_by(Condition) |> 
  summarise(
    n = n(),
    Mean = mean(RollsImagined) |> round(1),
    Median = median(RollsImagined),
    `Standard deviation` = sd(RollsImagined) |> round(1)
  ) |> 
  knitr::kable()
```

```{r, fig-boxplotklein, fig.cap = "Data from the replication of the gambler's fallacy study at the Brasília site.", fig.height = 4, fig.width = 5, out.width = "80%"}
d |> 
  ggplot(aes(x = Condition, y = RollsImagined)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(shape = 1, position = position_jitter(width = 0.1, height = 0)) +
  scale_x_discrete(name = "Condition",
    limits = c("two6", "three6"), labels = c("two sixes", "three sixes")) +
  ylab("Number of previous rolls imagined")
```

@fig-meandiffklein shows the distribution of the mean differences
under the null hypothesis.
Note the weird multimodal distribution of the test statistic.
Nevertheless, the _p_-value of 0.17 is valid for the null hypothesis
that none of the participants' estimations would have been different
in the other condition.

```{r, fig.height = 4, fig.width = 5, out.width = "80%"}
#| classes: preview-image
#| label: fig-meandiffklein
#| fig-cap: "The distribution of the mean difference under the sharp null hypothesis for $\\delta = 0$ for the Brasília replication of the gambler's fallacy."
randomisation_test(d$RollsImagined, d$n.Condition, mean_diff,
  breaks = 50, xlab = "Mean difference", main = "Distribution under H0")
```

Further, despite the wonky distribution of the mean difference under the null
hypothesis, Student's _t_-test produces a highly similar _p_-value.
```{r}
t.test(d$RollsImagined ~ d$n.Condition, var.equal = TRUE)
```

Klein et al. actually square-root-transformed the data.
If we feed the square-root-transformed data to a randomisation test,
we obtain $p = 0.048$.
```{r}
randomisation_test(sqrt(d$RollsImagined), d$n.Condition, mean_diff, plot = FALSE)
```
The corresponding _t_-test again yields quite a similar result,
though one that many researchers would interpret differently ($p = 0.06$).
```{r}
t.test(sqrt(d$RollsImagined) ~ d$n.Condition, var.equal = TRUE)$p.value
```

I suspect that Klein et al. used a square-root transformation
rather than a rank-based test because they wanted to compare results
across replicated studies, and this required them to express the findings
of all studies in terms of mean differences.
But since we don't have this problem, we can turn to other test statistics,
such as the probability of superiority; see @fig-propsupklein.

```{r, fig.height = 4, fig.width = 5, out.width = "80%"}
#| label: fig-propsupklein
#| fig-cap: "The distribution of the probability of superiority under the sharp null hypothesis for $\\delta = 0$ for the Brasília replication of the gambler's fallacy."
randomisation_test(d$RollsImagined, d$n.Condition, probsup,
  breaks = 50, xlab = "Probability of superiority", main = "Distribution under H0")
```
The resulting _p_-value (0.049) is very similar to the one produced
by the corresponding Wilcoxon--Mann--Whitney test.

```{r}
wilcox.test(d$RollsImagined ~ d$n.Condition, correct = FALSE)
```

Incidentally, you can use the $W = 1265$ statistic in the model output
to compute the probability of superiority like so:
```{r}
1 - 1265 / (sum(d$n.Condition == 1) * sum(d$n.Condition == 0))
```

## tl;dr
The randomisation model offers a way to compute valid 
_p_-values that are justified by the research design 
rather than by alluding to an imagined population and an idealised sampling procedure.
Nevertheless, the population model is useful, but mostly as an approximation. 
It allows for convenient, general-purpose methods (e.g., the _t_-test)
that tend to work well for largish samples.
But the asymptotic reasoning that underpins the use of these methods 
in randomised experiments involves complex mathematics
and only guarantees large-sample validity.
Randomisation tests, by contrast, are conceptually straightforward
and the resulting inferences are exact for any sample size.
In the words of Berger et al. (2002):

> ``If credibility for the parametric [= population-model test, JV]
> derives from assurances that its p-value will likely be close
> to the corresponding exact one,
> then this is tantamount to an admission that the exact test
> is the gold standard (or, perhaps, the platinum standard).'' (p. 79)

(Platinum was worth more than gold in 2002.)

Randomisation testing is also flexible: we can make up any test statistic we deem suitable
and obtain a valid _p_-value for the test of the sharp null hypothesis.
If the test statistic we choose has a classical population-model
counterpart (e.g., Student's two-sample _t_-test when testing the mean difference),
the randomisation test and its classical counterpart usually result
in similar _p_-values.
But running the randomisation test instead of the classical counterpart
is a small change that makes the resulting inferences more defensible.

## What's next?
The randomisation approach extends beyond simple two-group comparisons. 
In future posts, I'll explore how it applies to more complex experimental designs.

What's not obvious to me is how sensible confidence intervals could
be constructed in the randomisation model.
This, too, will have to be the subject for another post.

## References
Berger, Vance W., Clifford E. Lunneborg, Michael D. Ernst & Jonathan G. Levine.
2002.
[Parametric analysis in randomized clinical trials](https://doi.org/10.22237/jmasm/1020255120).
_Journal of Modern Applied Statistical Methods_ 1(1).
74--82.

Klein, Richard A., Kate A. Ratliff, Michelangelo Vianello, Reginald B. Adams Jr., Štěpán
Bahník, Michael J. Bernstein, Konrad Bocian et al. 2014. 
[Investigating variation in replicability: A “many labs” replication project](https://doi.org/10.1027/1864-9335/a000178). 
_Social Psychology_ 45(3). 
142--152.

Oppenheimer, Daniel M. & Benoît Monin. 2009. 
[The retrospective gambler's fallacy: Unlikely events, constructing the past, and multiple universes](https://doi.org/10.1017/S1930297500001170). 
_Judgment and Decision Making_ 4(5). 
326--334.

## Software versions
```{r}
devtools::session_info("attached")
```

