---
title: "Testing sharp and weak null hypotheses in randomised experiments"
date: "2026-11-09"
categories: [assumptions, experiments, design features, nonparametric tests, R, significance]
format:
  html:
    code-fold: true
    message: false
---

In a [previous post](../2025-11-09-population-randomisation/),
I explained how randomisation tests work and 
why they are a sensible choice when analysing the data from a randomised
experiments with a null hypothesis test.
In this post, I want to take a closer look at the distinction between 
so-called **sharp** (Fisherian) and **weak** (Neymanian) null hypotheses
and how they can be tested.
This post is in a way a synopsis of chapters 3, 4, and 8
of Ding's (2024) _First course in causal inference_.

## Recap: Randomisation tests of the sharp null hypothesis
Under the randomisation model, we posit that there are two 
potential outcomes $y_i(0)$ and $y_i(1)$ for each participant $i = 1, \dots, m$
in a randomised experiment with two conditions, only one of which is observed:

* either $y_i(0)$ if the $i$-th participant is assigned to the first condition;
* or $y_i(1)$ if the participant is assigned to the second condition.

According to Fisher's **sharp null hypothesis**, 
there is no differential treatment effect for _any_ of the participants:

$$H_{oS}: y_i(0) = y_i(1) \text{ for } 1 \leq i \leq m.$$

As I've explained [before](../2025-11-09-population-randomisation/),
this sharp null hypothesis can be tested exactly by means of a randomisation test
using _any_ definable test statistic.
This test statistic could be the observed mean difference between the two
conditions, the observed difference between the condition medians,
the probability of superiority -- anything the analyst deems sensible.

A test statistic that I did not use in the other blog post
but that is important now
is the **studentised mean difference**:

$$
  t := \frac{\widehat{y}(1) - \widehat{y}(0)}{\sqrt{\widehat{V}}}.
$$

In this equation, $\widehat{y}(0), \widehat{y}(1)$ are the condition means in the first
and second condition, respectively, 
and $\widehat{V}$ is defined as

$$
  \widehat{V} := \frac{\widehat{S}^2(0)}{n_0} + \frac{\widehat{S}^2(1)}{n_1},
$$

where $\widehat{S}^2(0)$ and $n_0$ are the sample variance and the group size
in the first condition
and $\widehat{S}^2(1)$ and $n_1$ are the sample variance and the group size
in the second condition.
Incidentally, 
this definition of $t$ agrees with the one used in Welch's $t$-test.

The following code snippet defines a function for computing 
the studentised mean difference. 
I've also copied the function for running a randomisation test 
from the other blog post.

```{r}
studentised_mean_diff <- function(x, group) {
  v_hat <- var(x[group == 1]) / length(x[group == 1]) + 
    var(x[group == 0]) / length(x[group == 0])
  (mean(x[group == 1]) - mean(x[group == 0])) / sqrt(v_hat)
}

randomisation_test <- function(
    # Randomisation test of sharp null in two-group comparison.
    x,           # outcome
    group,       # grouping variable
    fun,         # function for computing test statistic from x and group
    delta = 0,   # delta for sharp null hypothesis
    alternative = "two_sided", # two-sided, left-sided or right-sided p-value?
    M = 19999,   # number of new randomisations generated
    plot = TRUE, # plot histogram of test statistic under H0?
    ...          # parameters passed to histogram
) {
  test_stat <- fun(x, group)
  x[group == 1] <- x[group == 1] - delta
  test_stat_H0 <- replicate(M, {
    fun(x, sample(group))
  })
  test_stat_H0 <- c(test_stat, test_stat_H0)
  
  if (plot) {
    hist(test_stat_H0, ...)
    abline(v = test_stat, lwd = 2)
  }
  
  p_left <- mean(test_stat_H0 <= test_stat)
  if (substr(alternative, 1, 1) == "l") return(p_left)
  
  p_right <-mean(test_stat_H0 >= test_stat)
  if (substr(alternative, 1, 1) == "r") return(p_right)
  
  min(2 * min(p_left, p_right), 1)
}
```

These functions can be used like so:
```{r}
#| code-fold: show
# Define some data
d <- structure(list(
  Group = c(0, 0, 0, 0, 1, 0, 1, 0, 1, 1, 1, 0,
            1, 1, 1, 1, 0, 1, 0, 1, 0, 0, 1, 1, 1, 1, 0), 
  Outcome = c(7, 8, 9.5, 8.5, 7.5, 9, 8, 7.5, 10, 8, 8.5, 
              8, 9, 10.5, 8, 9, 7.5, 9, 9.5, 11, 8, 8, 7.5, 9, 9.5, 8.5, 8.5)), 
  class = "data.frame", row.names = seq_len(27))

# Randomisation test with studentised mean difference
randomisation_test(d$Outcome, d$Group, studentised_mean_diff, plot = FALSE)
```

Randomisation testing pretty easily achieves exact finite-sample inference
for the sharp null hypothesis and can straightforwardly be justified in
randomised experiments.
However, two drawbacks in particular bear mentioning.

The first concerns the construction of sensible **confidence sets**.
It _is_ possible to construct confidence sets for, for instance,
a constant treatment shift $\delta$.
This involves testing the sharp null hypothesis that the differential
treatment effect is $\delta$ for all participants:

$$
H_{oS}(\delta): y_i(1) = y_i(0) + \delta \text{ for } 1 \leq i \leq m.
$$
A $(1-\alpha)$ confidence set for the constant treatment shift is then
given by the set of all $\delta$ for which $H_{oS}$ cannot be rejected at the
$\alpha$ level.

<!-- * Exact finite-sample inference easily achieved by means of randomisation tests. -->
<!-- * Can construct CI for additive effect by testing -->

<!-- and taking a as the $1 - \alpha$ confidence set those $\delta$ values -->
<!-- for which $H_o(\delta)$ is not rejected at level $\alpha$. -->
<!-- (Similarly for multiplicative effects.) -->

<!-- Problem with CI: May be empty. Unrealistic that any non-zero effect should -->
<!-- be the same in all units. -->

<!-- ## The weak null hypothesis -->
<!-- The sharp null hypothesis posits that none of the units shows any -->
<!-- differential treatment effect. -->
<!-- Weak form: the _average_ treatment effect is zero. That is, -->
<!-- we define -->
<!-- $$\frac{1}{m}\sum_{i=1}^m y_i(0) =: \overline y(0)$$ -->
<!-- and -->
<!-- $$\frac{1}{m}\sum_{i=1}^m y_i(1) =: \overline y(1)$$ -->
<!-- and posit the null hypothesis that -->
<!-- $$H_{oW}: \overline y(1) - \overline y(1) =: \tau = 0.$$ -->
<!-- For each participant, we observe either $y_i(0)$ or $y_i(1)$, -->
<!-- so we don't observe $\overline y(0)$ or $\overline y(1)$ either. -->
<!-- Instead, we estimate $\widehat \tau$. -->



<!-- Define $\overline Y(1) := n^{-1}\sum_i Y_i(1)$ -->
<!-- and $\overline Y(0) := n^{-1}\sum_i Y_i(0)$. -->

<!-- Can show that  -->
<!-- $\widehat V$ is a conservative estimator of $\text{Var}(\widehat \tau)$. -->
<!-- Under some conditions, it can be shown that  -->
<!-- the distribution of -->

<!-- \[ -->
<!--   \frac{\widehat \tau - \tau}{\sqrt{\text{Var}(\widehat \tau)}} -->
<!-- \] -->

<!-- is approximately standard normal for large enough samples (Ding 2024, Theorem 4.2). -->

<!-- Since $\widehat V$ conservatively estimates $\text{Var}(\widehat \tau)$, -->
<!-- an asymptotically conservative $(1 - \alpha)$ confidence interval for $\tau$ -->
<!-- can be constructed as -->

<!-- \[ -->
<!--   \widehat{\tau} \pm z_{1 - \alpha} \sqrt{\widehat{V}}}. -->
<!-- \] -->

<!-- An asymptotic $p$-value is obtained by comparing  -->
<!-- $(\widehat{\tau} - \tau)/\sqrt{\widehat{V}}$ to a standard normal distribution. -->


<!-- ```{r} -->
<!-- d <- read.csv("Klein2014_gambler.csv") |>  -->
<!--   filter(Sample == "brasilia") |>  -->
<!--   filter(!is.na(RollsImagined)) -->
<!-- d$n.Condition <- ifelse(d$Condition == "three6", 1, 0) -->
<!-- d |>  -->
<!--   ggplot(aes(x = Condition, y = RollsImagined)) + -->
<!--   geom_boxplot(outlier.shape = NA) + -->
<!--   geom_point(shape = 1, position = position_jitter(width = 0.1, height = 0)) + -->
<!--   scale_x_discrete(name = "Condition", -->
<!--     limits = c("two6", "three6"), labels = c("two sixes", "three sixes")) + -->
<!--   ylab("Number of previous rolls imagined") -->
<!-- ``` -->


<!-- ```{r} -->

<!-- randomisation_test <- function( -->
<!--     # Randomisation test of sharp null in two-group comparison. -->
<!--     x,           # outcome -->
<!--     group,       # grouping variable -->
<!--     fun,         # function for computing test statistic from x and group -->
<!--     delta = 0,   # delta for sharp null hypothesis -->
<!--     alternative = "two_sided", # two-sided, left-sided or right-sided p-value? -->
<!--     M = 19999,   # number of new randomisations generated -->
<!--     plot = TRUE, # plot histogram of test statistic under H0? -->
<!--     ...          # parameters passed to histogram -->
<!-- ) { -->
<!--   test_stat <- fun(x, group) -->
<!--   x[group == 1] <- x[group == 1] - delta -->
<!--   test_stat_H0 <- replicate(M, { -->
<!--     fun(x, sample(group)) -->
<!--   }) -->
<!--   test_stat_H0 <- c(test_stat, test_stat_H0) -->

<!--   if (plot) { -->
<!--     hist(test_stat_H0, ...) -->
<!--     abline(v = test_stat, lwd = 2) -->
<!--   } -->

<!--   p_left <- mean(test_stat_H0 <= test_stat) -->
<!--   if (substr(alternative, 1, 1) == "l") return(p_left) -->

<!--   p_right <-mean(test_stat_H0 >= test_stat) -->
<!--   if (substr(alternative, 1, 1) == "r") return(p_right) -->

<!--   min(2 * min(p_left, p_right), 1) -->
<!-- } -->

<!-- ``` -->

<!-- ```{r} -->
<!-- set.seed(2026-07-09) # the day I wrote this -->
<!-- randomisation_test(d$RollsImagined, d$n.Condition, studentised_mean_diff,  -->
<!--   breaks = 50, xlab = "Studentised mean difference",  -->
<!--   main = "Distribution under H0") -->
<!-- ``` -->

<!-- Asymptotic Neyman test for weak null hypothesis: -->

```{r}
neyman_test <- function(
    # Asymptotic test of weak null in two-group comparison.
    x, 
    group, 
    alternative = "two_sided"
) {
  test_statistic <- studentised_mean_diff(x, group)

  p_left <- pnorm(test_statistic, lower.tail = TRUE)
  if (substr(alternative, 1, 1) == "l") return(p_left)

  p_right <- pnorm(test_statistic, lower.tail = FALSE)
  if (substr(alternative, 1, 1) == "r") return(p_right)

  min(2 * min(p_left, p_right), 1)
}

# Alternatively, use Welch' t-test.
```

<!-- The  -->

<!-- ```{r} -->
<!-- #| results: asis -->
<!-- # Let's do this for all samples: -->

<!-- d <- read.csv("Klein2014_gambler.csv") |>  -->
<!--   filter(!is.na(RollsImagined)) -->
<!-- d$n.Condition <- ifelse(d$Condition == "three6", 1, 0) -->

<!-- pvals_HoF <- rep(0, length(unique(d$Sample))) -->
<!-- pvals_HoN <- rep(0, length(unique(d$Sample))) -->

<!-- for (i in seq_along(unique(d$Sample))) { -->
<!--   my_sample <- unique(d$Sample)[[i]] -->
<!--   d_sample <- d |> filter(Sample == my_sample) -->
<!--   pvals_HoF[[i]] <- randomisation_test(d_sample$RollsImagined, d_sample$n.Condition,  -->
<!--     studentised_mean_diff, breaks = 50,  -->
<!--     xlab = "Studentised mean difference",  -->
<!--     main = paste("Distribution under HoF in the", my_sample, "sample")) -->
<!--   pvals_HoN[[i]] <- neyman_test(d_sample$RollsImagined, d_sample$n.Condition) -->
<!-- } -->

<!-- ``` -->

## Simulation

```{r}
simulate_sharp <- function() {
  
}
```

```{r}
simulate_weak <- function() {
  
}
```




## Reference
Ding, Peng.
2024.
_A first course in causal inference_.
Boca Raton, FL: CRC Press.
Also available from [arXiv.org](https://doi.org/10.48550/arXiv.2305.18793).

## Software versions
```{r}
devtools::session_info("attached")
```

