---
title: "Fitting interactions between continuous variables"
date: "2017-06-26"
categories: [R, graphs, generalised additive models, non-linearities]
---

Splitting up continuous variables is generally a 
[bad](../2015-10-16-nonlinear-relationships) idea.
In terms of statistical efficiency, the popular practice of dichotomising 
continuous variables at their median is comparable to throwing out a third of 
the dataset. Moreover, statistical models based on split-up continuous variables 
are prone to misinterpretation: threshold effects are easily read into the 
results when, in fact, none exist. Splitting up, or ‘binning’, continuous 
variables, then, is something to avoid. But what if you’re interested in how the 
effect of one continuous predictor varies according to the value of another 
continuous predictor? In other words, what if you’re interested in the 
interaction between two continuous predictors? Binning one of the predictors 
seems appealing since it makes the model easier to interpret. However, as I’ll 
show in this blog post, it’s fairly straightforward to fit and interpret 
interactions between continuous predictors.

## Data

To keep things easy, I will use a simulated dataset with two continuous predictors (`pred1` and `pred2`) and one continuous `outcome`. The dataset is available [here](http://homeweb.unifr.ch/VanhoveJ/Pub/Data/example_interaction.csv).


```{r}
df <- read.csv("http://homeweb.unifr.ch/VanhoveJ/Pub/Data/example_interaction.csv")
```

## Exploratory plots
First, let’s draw some quick graphs to get a sense of how the predictors and the outcome are related.

```{r}
#| fig.cap: "_Left:_ The predictors are negatively correlated with each other. _Middle:_ `pred1` is negatively correlated with the outcome. _Right:_ `pred2` seems to be nonlinearly correlated with the outcome."
# 3 plots next to each other
par(mfrow = c(1, 3))

# Intercorrelation pred1/pred2
plot(pred2 ~ pred1, df)
lines(lowess(df$pred1, df$pred2), col = "red")

# Pred1 vs. outcome
plot(outcome ~ pred1, df)
lines(lowess(df$pred1, df$outcome), col = "red")

# Pred2 vs. outcome
plot(outcome ~ pred2, df)
lines(lowess(df$pred2, df$outcome), col = "red")

# Normal plotting again
par(mfrow = c(1, 1))
```

When we suspect there’s an interaction between two predictors, it’s useful to draw **conditioning plots** as well. For these plots, the dataset is split up into a number of overlapping equal-sized regions defined by a conditioning variable, and the relationship between the predictor of interest and the outcome within each region is plotted.


```{r}
#| fig.cap: "The effect of `pred1` on the outcome in four regions defined by `pred2`. The negative relationship grows stronger as `pred2` increases (from left to right)."
# coplot's in the lattice package
library(lattice)
coplot(outcome ~ pred1 | pred2, data = df, 
       number = 4, rows = 1,
       panel = panel.smooth)
```

```{r}
#| fig.cap: "The effect of `pred2` on the outcome in four regions defined by `pred1`. The relationship may be positive for small values of `pred1`, but becomes negative for larger values (from left to right)."
coplot(outcome ~ pred2 | pred1, data = df, 
       number = 4, rows = 1,
       panel = panel.smooth)
```

## Fitting the model
I find it easiest to fit the interaction between two continuous variables as a 
wiggly regression surface. A wiggly regression surface is the generalisation of 
a wiggly curve, such as the one in Figure 3 in 
[this](../2015-10-16-nonlinear-relationships) earlier blog post, into two 
dimensions. The advantage of fitting a wiggly surface rather than a plane is 
that we don’t have to assume that the interaction is linear. Rather, the shape 
of the surface will be estimated from the data.
To fit such wiggly surfaces, we need the `gam()` function in the `mgcv` package. 
The wiggly regression surface is fitted by feeding the two predictors to a `te()` 
function within the `gam()` call.


```{r}
# install.packages("mgcv")
library(mgcv)
m1 <- gam(outcome ~ te(pred1, pred2), data = df)
```

A numerical summary of the model can be obtained using `summary()`, but as you can see, this doesn’t offer much in the way of interpretable results. 

```{r}
summary(m1)
```


To understand what the wiggly regression surface looks like, you need to 
**plot** it. This can be done using the `vis.gam()` function. The two plots 
below show the same regression surface: once as a three-dimensional plot and 
once as a contour plot. The latter can be read like a topographic map of hilly 
terrain: the contour lines connect points on the surface with the same height, 
and the closer the contour lines are to each other, the steeper the surface is.


```{r}
# 3-d graph
vis.gam(m1, plot.type = "persp", color = "terrain", 
        theta = 135, # changes the perspective
        main = "")
```

```{r}
# A contour plot
vis.gam(m1, plot.type = "contour", color = "terrain",
        main = "")
```

What can be gleaned from these plots? First of all, the outcome variable is 
largest for low `pred1` and `pred2` values. 
Second, `pred1` and `pred2` both have strong negative effects when the respective 
other variable is large, but their effect is quite weak when the other variable is low. 

The contour lines are useful to quantify this. Going along the y-axis for a 
low `pred1` value (say, 2.5), we only cross one contour line, and we seem to 
climb slightly. 
Going along the y-axis for a large `pred1` value (say, 12.5), we cross 8 contour 
lines and we descend considerably, from around 100 on the outcome variable to 
below 30. 
Similarly, going along the x-axis for a low `pred2` value (say, 0.2), we descend 
slightly (perhaps 20 units). 
Going alongside the x-axis for a large `pred2` value (say, 0.8), we cross 10 
contour lines. 
The effect of `pred1`, then, depends on the value of `pred2` and vice versa. 
Or in other words, `pred1` and `pred2` interact with each other.

## Improved visualisation
I find it useful to plot a contour plot with the wiggly regression surface with the interaction alongside a contour plot with a regression surface _without_ an interaction. I find this helps me to better understand what the interaction consists of. This is particularly helpful when the one or two of the predictors are nonlinearly related to the outcome. To draw the second contour plot, first fit a model without an interaction. By feeding each predictor to a separate `te()` function, potentially nonlinear regression curves are estimated for both predictors.

```{r}
m2 <- gam(outcome ~ te(pred1) + te(pred2), data = df)
```

Then plot both contour plots alongside each other. Here I also set the `too.far` parameter. This way, region of the graph for which no combinations of predictor variables are available are not plotted.

```{r}
# Two plots next to each other
par(mfrow = c(1, 2))
vis.gam(m2, plot.type = "contour", color = "terrain", 
        too.far = 0.1, main = "without interaction")
vis.gam(m1, plot.type = "contour", color = "terrain", 
        too.far = 0.1, main = "with interaction")
par(mfrow = c(1, 1))
```

## Decomposing a regression surface into main effects and an interaction
It may be useful to 'decompose' the regression surface into the main effects 
associated with each predictor and the interaction itself.
This can be done by fitting the main effects and the interaction
in separate `ti()` (**not** `te()`) functions. This fits non-linear main effects, 
which is what we want: this way, the interaction term doesn't absorb any 
nonlinearities not modelled by the linear main effects.

```{r}
m3 <- gam(outcome ~ ti(pred1) + ti(pred2) + # main effects
                    ti(pred1, pred2),       # interaction
          data = df)
summary(m3)
```

The wiggly regression surface of `m1` has here been decomposed
into the main effects of `pred1` and `pred2` and an interaction
between them. Both main effects and the interaction are significant.

The `edf` values in the table above, incidentally, express
how nonlinear the effects are estimated to be. Values close
to 1 indicate that the effect is linear, whereas a value of 2
suggests that the effect can be modelled using a linear
and a quadratic term etc. Here both main effects could
sensibly be modelled linearly, but the interaction couldn't.
(A linear interaction means that you multiply both predictors together and use 
this product as an additional predictor.)

There's much more to fitting nonlinearities and interactions using `gam()` 
than discussed here, but hopefully this was useful enough to dissuade you from 
carving up continuous variables when you're interested in their interaction.

## Software versions
```{r}
devtools::session_info()
```
