---
title: "Model comparisons for mixed-effects regression, or: When `anova()` doesn't do ANOVA"
date: "2026-08-12"
categories: [assumptions, mixed-effects models, multiple regression, R, significance]
format:
  html:
    code-fold: show
    message: false
    df-print: paged
    fig-width: 6
    fig-height: 5
---

I recently reviewed a paper in which the authors wrote that they compared
two mixed-effects models using ANOVA (analysis of variance).
That wasn't quite right, though:
while they did use R's `anova()` function to compare their two models,
this function doesn't actually run an ANOVA when the models being compared
are mixed-effects models.
Below, I explain what the `anova()` function actually does,
why you should take its output with a pinch of salt,
and how you can obtain more accurate results.

## Where the `anova()` function got its name
It's entirely reasonable to assume that comparing mixed-effects models
using `anova()` involves, well, ANOVA.
The function is a bit of a misnomer, however,
and it is instructive to reconstruct where it got its name from.
To do this, we need to consider how model comparisons for the more everyday
general linear model can be carried out.

I'll use some data that's built-in into R to fit two models.
These models are nested in the sense that every parameter that is estimated
in model `m0` is also estimated in model `m1`, but not vice versa:

```{r}
options(show.signif.stars = FALSE)

data(swiss) # see ?swiss for details
m0 <- lm(Fertility ~ Agriculture, data = swiss)
m1 <- lm(Fertility ~ Agriculture + Education, data = swiss)
```

When we apply `anova()` to the larger model (`m1`),
we obtain a standard analysis of variance table that shows a partitioning
of the variability in the outcome:

```{r}
anova(m1)
```

The sums of squares reported are sequential sums of squares 
(so-called Type-I sums of squares, in typical intransparent statistics lingo).
This means that the `Education` variable accounted for an additional 2329.8
units of sum of squares beyond what was already accounted for by
the `Agriculture` variable.[^1]

[^1]: If we switched the order of the predictors in the `lm()` call,
we'd obtain a different partitioning. 
The reason is that any variance in the outcome that could have been assigned to either
`Education` or `Agriculture` would now be fully assigned to `Education` rather
than to `Agriculture` as in model `m1`.

Informally, the _F_-test for the `Education` predictor tests if this predictor
can account for more variance than would be expected if `Education` did not 
have any linear association with `Fertility` once the linear effect
of `Agriculture` on `Fertility` has been accounted for.
The _F_-test is finite-sample exact if the model's errors are identically
and independently sampled from a normal distribution.
If the errors are independent but not necessarily 
stem from the same normal distribution,
then, by and large, the _F_-test is still serviceable 
as a large-sample approximation.[^2]

[^2]: When in doubt, you could try a permutation test. But 
in my experience, as long as the model is a relevant one,
these asymptotics kick in pretty quickly.
Hence: [_Before worrying about model assumptions, think about model relevance_](https://janhove.github.io/posts/2019-04-11-assumptions-relevance/).

Now, the variance partitioning above doesn't explicitly involve `m0`.
But it's there implicitly: We're effectively comparing a model with
`Agriculture` and `Education` as predictors to a model that only has 
`Agriculture` as a predictor.
We can make this comparison more explicit by feeding both models to
the `anova()` function:

```{r}
anova(m0, m1)
```


The result is the same.
This explains why the function for comparing
two nested linear models is called `anova()`.

The advantage of this second use of `anova()` is that it allows you to 
fit several predictors at once and check if these new predictors
jointly earn their keep:

```{r}
m2 <- lm(Fertility ~ Agriculture + Education + Examination + Infant.Mortality, 
         data = swiss)
```

A sum of squares partitioning lists each predictor separately:

```{r}
anova(m2)
```

The model comparison tests the joint contribution of the three additional predictors.
```{r}
anova(m0, m2)
```

## Comparing mixed-effects models
Let's now turn to comparisons of nested mixed-effects models,
specifically mixed-effects models with the same random effects
but where the fixed-effect parameters estimated in the null model
are a subset of those in the larger model.
I use a dataset on the academic performance of primary school children
that was used by Faraway (2006),
but I whittle down the dataset to just 12 schools
and one school year to make the point clearer.

```{r}
library(lme4)

set.seed(2026-08-12)
data(jsp, package = "faraway")
selected_schools <- sample(levels(jsp$school), 12)
jspr <- subset(jsp, year == 2 & school %in% selected_schools)
```

The goal is to model the pupils' `math` performance
in terms of their intelligence as measured by the `raven` test
and the `social` class of their father, represented here as a factor
with nine levels.
The question is whether `raven` interacts with `social` class in terms
of their effects on `math` performance.
To that end, two models are fitted,
both of which take into account that the pupils are nested in schools
by means of random effects:[^3]

```{r}
m0 <- lmer(math ~ raven + social + (1|school),
           data = jspr)
m1 <- lmer(math ~ raven * social + (1|school),
           data = jspr)
```

[^3]: I ignore the nesting of pupils in classes for the sake of simplicity.

We can apply `anova()` to model `m1`, 
but this doesn't actually run an _F_-test.
The reason is that it's not clear what the correct number of degrees of freedom
for the residual sum of squares should be.[^4]

```{r}
anova(m1)
```

[^4]: You can squeeze out a _p_-value from this table if you must
using the `lmerTest` package. But the tests this package implements
are approximations.

But like before, we can use `anova()` to obtain a model comparison:

```{r}
anova(m0, m1)
```

Despite being called `anova()`,
this function, when applied to two or more mixed-effects models,
runs a **likelihood ratio test**.
First, both models, which were fitted using restricted maximum likelihood
originally, are refitted using (standard) maximum likelihood
for reasons discussed by Faraway (2006, Chapter 8).
Then, the likelihood of the data is computed under both models.
Call these $L_0$ and $L_1$, respectively.
According to a theorem due to Wilks (1938),
the quantity
\[
  T := -2 \log\left(\frac{L_0}{L_1}\right)
  = -2(\log L_0 - \log L_1)
\]
is approximately $\chi^2$-distributed in large samples under the null hypothesis,
with the degrees of freedom equalling the number of additional parameters
in the larger model.

The output above gives us $\log L_0 = -789.53$ and $\log L_1 = -782.82$,
from which we obtain $-2 \log\left(\frac{L_0}{L_1}\right) = 13.419$.
This value is then compared against the theoretical asymptotic distribution,
i.e., a $\chi^2$ distribution with 8 degrees of freedom, resulting
in a _p_-value of 0.098:
```{r}
pchisq(13.419, df = 8, lower.tail = FALSE)
```

Since the quantity $L_0/L_1$ is a ratio of likelihoods,
the test is called a **likelihood ratio test** (LRT).

The operative words in the preceding explanation are _approximately_
and _asymptotic_. What these words mean is that, under the null hypothesis,
the quantity $T$ isn't actually $\chi^2$-distributed---but for ever larger
samples, the discrepancies between its distribution and the relevant
$\chi^2$ distribution become ever more negligible.
Unfortunately, there is no sensible answer to the question of which sample
size is `large enough', and the actual distribution of $T$ may differ
noticeably from a $\chi^2$ distribution in practice.

This is why it is recommended to **simulate** the distribution 
of $T$ under the null hypothesis rather than assume the asymptotics have
kicked in sufficiently.
To achieve this, we can use **parametric bootstrapping**, that is,
we simulate new outcomes from the null model (`m0`),
refit the null model and the larger model using these simulated data,
and extract the test statistic $T$ from the model comparison.
We do this a couple of thousand times,
and then compare the test statistic that we actually obtained
against the distribution of the test statistics obtained via simulation.

The following code snippet estimates the null distribution of the LRT statistic:
```{r}
# Refit models using (standard) maximum likelihood
m0 <- lmer(math ~ raven + social + (1|school),
           data = jspr, REML = FALSE)
m1 <- lmer(math ~ raven * social + (1|school),
           data = jspr, REML = FALSE)

# Obtain LRT statistic from model comparison
observed_stat <- anova(m0, m1)[2, 6]

# Simulate 2,000 datasets from null model
M <- 2000
simdat <- simulate(m0, M)

# Fit models to simulated data; extract LRT stats
ctrl <- lmerControl(check.conv.singular = "ignore") # fewer warnings in output
LRTs <- numeric(M)
for (i in seq_len(M)) {
  H0_m0 <- lmer(simdat[, i] ~ raven + social + (1|school),
                data = jspr, REML = FALSE, control = ctrl)
  H0_m1 <- lmer(simdat[, i] ~ raven * social + (1|school),
                data = jspr, REML = FALSE, control = ctrl)
  LRTs[i] <- anova(H0_m0, H0_m1)[2, 6]
}
```

@fig-simulated-asymptotic reveals pretty pronounced discrepancies between
the test statistic's estimated null distribution
and its asymptotic null distribution.

```{r}
#| label: fig-simulated-asymptotic
#| fig-cap: "The black curve shows the distribution function of the $\\chi^2$ distribution with 8 degrees of freedom, which is the asymptotic distribution of the likelihood ratio test statistic in our example. The red curve shows the estimated distribution function of the test statistic under the null distribution. The difference is quite substantial. The test statistic that was actually observed is highlighted by the dashed vertical line."
plot(ecdf(LRTs), col = "darkred",
     main = "Simulated vs asymptotic LRT null distribution",
     xlab = "T", ylab = "Cumulative probability")
curve(pchisq(x, 8), add = TRUE, col = "black")
abline(v = observed_stat, lty = 2, col = "#008cba")
```

A _p_-value can be obtained by computing the number of test statistics simulated
under the null hypothesis that are at least as large as the test statistic
that we obtained for our actual data:
```{r}
mean(LRTs >= observed_stat)
```

So rather than $p = 0.098$, we obtain $p = 0.12$.

Incidentally, while it's also possible to compare nested linear models using
a likelihood ratio test, there is little upside to doing so:
asymptotically, the _F_-test and the likelihood ratio test will converge 
to the same result, but the _F_-test will do so more quickly than
the likelihood ratio test.
So you can just stick to the _F_-test instead.
The parametric bootstrap approach, 
which is useful for comparing mixed-effects models,
has no added value when comparing linear models.

## tl;dr
When applied to a single linear model, `anova()` runs an ANOVA.
To the extent that the model's assumptions aren't met,
the results of this ANOVA are approximate but generally quite accurate.

When applied to two or more nested linear models, `anova()` runs a model
comparison by means of an _F_-test,
the results of which coincide with those of an ANOVA---hence the function's name.

When applied to two or more nested mixed-effects models, `anova()`
runs a model comparison by means of an asymptotic likelihood
ratio test.
_Even if_ the models' assumptions are met,
the results of this test are approximate.
Parametric bootstrapping can be used to obtain more accurate results.

## References
Faraway, Julian.
2006.
_Extending the linear model with R: Generalized linear, mixed effects and nonparametric regression models._
Boca Raton, FL: Chapman & Hall/CRC.

Wilks, Samuel S.
1938.
[The large-sample distribution of the likelihood ratio for testing composite hypotheses.](https://doi.org/10.1214/aoms/1177732360)
_The Annals of Mathematical Statistics_ 9(1).
60--62.

## Software versions
```{r, warning = FALSE}
devtools::session_info("attached")
```
